---
title: "Exercise 2 - A differential expression analysis, end to end"
output: html_document
---

**Curriculum link:** `stats.md` -> Topics 8, 13, 20, 29
**Core modules used:** 08, 13, 20, 29

## The brief

You are handed a raw count matrix: 12 samples (6 control, 6 treated).
Nobody tells you how it was made. Build the analysis yourself and, at every
step, **check the step rather than trusting it**.

| Q | Step | The thing that goes wrong |
|---|---|---|
| 1 | Normalisation | composition bias - a few huge genes distort every library |
| 2 | Dispersion | per-gene estimates are hopeless at n = 6 |
| 3 | Testing + FDR | the test is fine; calibration and filtering are not |
| 4 | Enrichment | inter-gene correlation inflates set-level p-values |
| 5 | Negative control | does the whole pipeline return nothing on a null? |

Question 5 matters most and is the one almost nobody runs.

```{r setup, message = FALSE}
suppressPackageStartupMessages(library(MASS))
MODULE_NAME <- "E2_rnaseq_end_to_end"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## The data

```{r data}
make_counts <- function(seed = 7, n_per_group = 6, n_genes = 2500, n_de = 200,
                        composition_bias = TRUE) {
  set.seed(seed)
  n <- 2*n_per_group
  group <- rep(0:1, each = n_per_group)
  ## True expression level per gene, log-normal as real data are.
  base <- exp(rnorm(n_genes, 4.0, 1.6))
  ## Mean-dispersion trend: low-count genes are much noisier (eq. 20.3).
  disp <- 0.10 + 18.0/(base + 12.0)
  lfc <- numeric(n_genes)
  de_idx <- sample(n_genes, n_de)
  lfc[de_idx] <- rnorm(n_de, 0, 1.1)
  ## Correlated modules: each responds to the processing BATCH plus a smaller
  ## unmeasured latent factor. Both move all of a module's genes together,
  ## which is what makes genes in a pathway non-independent.
  ##
  ## Batch is balanced within each group, so it is NOT confounded with
  ## treatment - but leaving it out of the design still inflates error rates.
  batch <- rep(rep(0:1, length.out = n_per_group), 2)
  n_mod <- 10; mod_size <- 50
  modules <- list()
  mod_factor <- matrix(0, n_genes, n)
  for (k in seq_len(n_mod)) {
    members <- ((k - 1)*mod_size + 1):(k*mod_size)
    modules[[sprintf("module_%02d", k)]] <- members
    mod_factor[members, ] <- rep(rnorm(1, 0, 0.55)*batch + rnorm(n, 0, 0.18),
                                 each = length(members))
  }
  ## Modules 1 and 2 are genuinely up-regulated; the rest are not.
  for (k in 1:2) {
    lfc[modules[[k]]] <- rnorm(mod_size, 0.9, 0.25)
    de_idx <- union(de_idx, modules[[k]])
  }
  lib <- runif(n, 0.5, 2.0)                    # library size varies 4-fold
  mu <- outer(base, rep(1, n)) * 2^outer(lfc, group) * exp(mod_factor) *
        rep(lib, each = n_genes)
  if (composition_bias) {
    ## 5 genes explode in the treated group: real biology, but it steals
    ## sequencing depth from every other treated gene.
    hogs <- order(base, decreasing = TRUE)[1:5]
    mu[hogs, group == 1] <- mu[hogs, group == 1]*60
  }
  ## Sequencing measures PROPORTIONS: renormalise each column to a fixed total.
  mu <- sweep(mu, 2, colSums(mu), "/") * rep(4e6*lib, each = n_genes)
  counts <- matrix(rnbinom(n_genes*n, size = 1/disp, mu = mu), n_genes)
  list(counts = counts, group = group, batch = batch, lfc_true = lfc,
       is_de = seq_len(n_genes) %in% de_idx, modules = modules,
       disp_true = disp, base = base)
}
header("The count matrix as received")
D <- make_counts()
counts <- D$counts; group <- D$group; batch <- D$batch
cat(sprintf("  %d genes x %d samples (%d control, %d treated)\n",
            nrow(counts), ncol(counts), sum(group == 0), sum(group == 1)))
cat(sprintf("  library sizes (millions): %s\n",
            paste(round(colSums(counts)/1e6, 2), collapse = " ")))
cat(sprintf("  ratio largest/smallest library = %.2f\n",
            max(colSums(counts))/min(colSums(counts))))
cat(sprintf("  median count %.0f, max count %s\n", median(counts),
            format(max(counts), big.mark = ",")))
cat(sprintf("  group : %s\n  batch : %s   <- recorded, balanced within group\n",
            paste(group, collapse = " "), paste(batch, collapse = " ")))
```

### Q1: Normalisation

Compare total-count (CPM), upper-quartile and median-of-ratios size factors
(eq. 20.1). Which recovers the truth, and how would you *detect* the
problem without knowing the truth?

```{r q1}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# sf_total <- function(c) { s <- colSums(c); s/exp(mean(log(s))) }
# sf_uq <- function(c) {
#   q <- apply(c, 2, function(x) quantile(x[x > 0], 0.75))
#   q/exp(mean(log(q)))
# }
# sf_median_ratio <- function(c) {
#   ## DESeq2 median-of-ratios (eq. 20.1). The geometric mean across samples is
#   ## a per-gene REFERENCE; a sample's size factor is the median of its ratios
#   ## to that reference. Because it is a MEDIAN, a handful of exploding genes
#   ## cannot move it.
#   keep <- apply(c > 0, 1, all)          # the geometric mean needs no zeros
#   lc <- log(c[keep,, drop = FALSE])
#   exp(apply(lc - rowMeans(lc), 2, median))
# }
# sfs <- list("total count (CPM)" = sf_total(counts),
#             "upper quartile" = sf_uq(counts),
#             "median of ratios" = sf_median_ratio(counts))
# cat(sprintf("  %-22s%11s%11s%11s\n", "method", "ctrl mean", "trt mean", "trt/ctrl"))
# for (nm in names(sfs)) {
#   s <- sfs[[nm]]
#   cat(sprintf("  %-22s%11.3f%11.3f%11.3f\n", nm, mean(s[group == 0]),
#               mean(s[group == 1]), mean(s[group == 1])/mean(s[group == 0])))
# }
# nde <- !D$is_de
# cat(sprintf("\n  %-22s%32s\n", "method", "median logFC of NON-DE genes"))
# for (nm in names(sfs)) {
#   norm <- sweep(counts, 2, sfs[[nm]], "/")
#   a <- log2(rowMeans(norm[nde, group == 1]) + 1)
#   b <- log2(rowMeans(norm[nde, group == 0]) + 1)
#   cat(sprintf("  %-22s%32.3f\n", nm, median(a - b)))
# }
# ## DIAGNOSIS WITHOUT THE TRUTH: check what share of each library the top
# ## genes consume.
# frac <- apply(counts, 2, function(x) sum(sort(x, decreasing = TRUE)[1:5])/sum(x))
# cat(sprintf("\n  share of library taken by the top 5 genes:\n"))
# cat(sprintf("    control : %.1f%%\n    treated : %.1f%%   <- composition bias\n",
#             100*mean(frac[group == 0]), 100*mean(frac[group == 1])))
#
# ## Total-count normalisation divides by a number the 5 exploding genes
# ## dominate, so every OTHER treated gene is scaled down and looks
# ## DOWN-regulated: the median non-DE logFC moves away from 0. Median-of-ratios
# ## and upper-quartile are robust to a few enormous genes and keep it near 0.
# ## This is why DESeq2/edgeR never use CPM for testing, and why "normalise to
# ## total reads" is wrong whenever composition changes - the same closure
# ## problem as microbiome data (Topic 26).
```

### Q2: Dispersion

Estimate the NB dispersion per gene by method of moments, then shrink it
towards a fitted mean-dispersion trend (eq. 20.4). Show numerically why the
raw per-gene estimate must not be used at n = 6.

```{r q2}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# sf <- sf_median_ratio(counts)
# norm <- sweep(counts, 2, sf, "/")
# keep <- rowMeans(norm) >= 1        # keep almost everything; Q3 studies filtering
# nrm <- norm[keep,, drop = FALSE]; dtrue <- D$disp_true[keep]
# mu_hat <- rowMeans(nrm)
# var_hat <- apply(nrm, 1, var)
# ## NB: Var = mu + disp*mu^2  ->  disp = (Var - mu)/mu^2   (eq. 20.2)
# disp_mom <- pmax((var_hat - mu_hat)/mu_hat^2, 1e-4)
# ## Trend: dispersion as a smooth function of the mean (eq. 20.3), in log space.
# o <- order(mu_hat)
# sm <- stats::filter(log(disp_mom[o]), rep(1/151, 151), sides = 2)
# sm[is.na(sm)] <- log(disp_mom[o])[is.na(sm)]
# disp_trend <- numeric(length(disp_mom)); disp_trend[o] <- exp(sm)
# ## Empirical-Bayes shrinkage towards the trend (eq. 20.4). The weight depends
# ## on the residual degrees of freedom, n - 2 = 10 here. trigamma(df/2) is the
# ## sampling variance of a log-variance estimate - NOT 2/df.
# dfres <- ncol(counts) - 2
# logres <- log(disp_mom) - log(disp_trend)
# prior_var <- max(var(logres) - trigamma(dfres/2), 0.01)
# wt <- prior_var/(prior_var + trigamma(dfres/2))
# disp_shrunk <- exp(log(disp_trend) + wt*logres)
# cat(sprintf("  %d genes kept (mean normalised count >= 1)\n", sum(keep)))
# cat(sprintf("  residual df = %d, shrinkage weight towards the trend = %.2f\n",
#             dfres, 1 - wt))
# cat(sprintf("\n  %-26s%18s%18s\n", "estimator", "corr with truth", "median |error|"))
# for (nm in list(c("per-gene MoM", "disp_mom"), c("trend only", "disp_trend"),
#                 c("shrunk (empirical Bayes)", "disp_shrunk"))) {
#   est <- get(nm[2])
#   cat(sprintf("  %-26s%18.3f%18.3f\n", nm[1], cor(log(est), log(dtrue)),
#               median(abs(log(est) - log(dtrue)))))
# }
#
# ## With 10 residual degrees of freedom a variance estimate has a relative
# ## standard error of about sqrt(2/10) = 45%, so the per-gene dispersion is
# ## extremely noisy - and worst exactly where it hurts, in the low-count genes.
# ## Shrinking each gene towards a trend fitted across thousands of genes
# ## borrows strength (Topic 33) and is the single reason DESeq2/edgeR work at
# ## n = 3-6.
```

### Q3: Testing, calibration and FDR

Fit a negative-binomial GLM per gene with a size-factor offset. Check
whether the TEST is calibrated before trusting the FDR, then show what
independent filtering buys.

```{r q3}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# kept_idx <- which(keep)
# truth_kept <- D$is_de[kept_idx]
# in_mod <- kept_idx %in% unlist(D$modules)
# off <- log(sf)
# fit_all <- function(use_batch, test = c("lrt", "wald")) {
#   test <- match.arg(test)
#   p <- rep(1, length(kept_idx)); lfc <- numeric(length(kept_idx))
#   for (i in seq_along(kept_idx)) {
#     yv <- counts[kept_idx[i], ]
#     fam <- MASS::negative.binomial(theta = 1/disp_shrunk[i])
#     d1 <- if (use_batch) data.frame(yv, group, batch) else data.frame(yv, group)
#     f1 <- try(glm(yv ~ ., fam, data = d1, offset = off), silent = TRUE)
#     if (inherits(f1, "try-error")) next
#     lfc[i] <- coef(f1)["group"]/log(2)
#     if (test == "wald") {
#       ## dispersion = 1 is ESSENTIAL: negative.binomial() is not on
#       ## summary.glm's "known dispersion" list, so by default summary()
#       ## estimates an EXTRA scale parameter and switches to a t reference.
#       p[i] <- coef(summary(f1, dispersion = 1))["group", 4]
#     } else {
#       d0 <- if (use_batch) data.frame(yv, batch) else data.frame(yv)
#       f0 <- try(glm(yv ~ ., fam, data = d0, offset = off), silent = TRUE)
#       if (inherits(f0, "try-error")) next
#       p[i] <- pchisq(f0$deviance - f1$deviance, 1, lower.tail = FALSE)
#     }
#   }
#   list(p = p, lfc = lfc)
# }
# ## --- (a) Is the TEST calibrated? Measure the null p-value rate directly. ---
# cat("  Rate of p < 0.05 among genes with NO true effect (should be 0.05):\n\n")
# cat(sprintf("  %-20s%-8s%12s%12s%12s\n", "design", "test", "all nulls",
#             "in module", "elsewhere"))
# cal <- list()
# for (dn in c("~ group", "~ group + batch")) {
#   for (tn in c("wald", "lrt")) {
#     r <- fit_all(use_batch = (dn == "~ group + batch"), test = tn)
#     cal[[paste(dn, tn)]] <- r
#     nul <- !truth_kept
#     cat(sprintf("  %-20s%-8s%12.4f%12.4f%12.4f\n", dn, toupper(tn),
#                 mean(r$p[nul] < 0.05), mean(r$p[nul & in_mod] < 0.05),
#                 mean(r$p[nul & !in_mod] < 0.05)))
#   }
# }
# ## WALD is anticonservative here and the LRT is closer to nominal. A Wald test
# ## uses the curvature of the likelihood AT THE ESTIMATE; with 12 samples that
# ## quadratic approximation is poor and it errs towards small p-values (the
# ## extreme case is the Hauck-Donner effect, Topic 13). Prefer the LRT at small
# ## n - it is what edgeR's glmLRT does.
# ##
# ## Also note that module genes are only modestly worse than the rest (about
# ## 0.087 vs 0.069), despite carrying a whole extra source of variation. The
# ## dispersion was estimated FROM THESE DATA, so the batch-driven spread inside
# ## each module was absorbed into a larger per-gene dispersion. Unmodelled
# ## structure gets LAUNDERED into the variance estimate. That protects the type
# ## I error - and costs POWER, since real effects in those genes are now tested
# ## against an inflated variance.
#
# ## --- (b) Independent filtering, using the best-calibrated test. ---
# best <- cal[["~ group + batch lrt"]]
# pvals <- best$p
# report <- function(name, mask) {
#   q <- rep(1, length(pvals)); q[mask] <- p.adjust(pvals[mask], "BH")
#   disc <- q < 0.05
#   cat(sprintf("  %-36s%8d%9d%8.3f%8.3f\n", name, sum(mask), sum(disc),
#               sum(disc & !truth_kept)/max(sum(disc), 1),
#               sum(disc & truth_kept)/sum(truth_kept)))
# }
# cat(sprintf("\n  %-36s%8s%9s%8s%8s\n", "analysis (LRT, ~ group + batch)",
#             "tested", "discov.", "FDP", "power"))
# report("no filtering", rep(TRUE, length(pvals)))
# for (th in c(5, 30, 100)) report(sprintf("independent filter: mean >= %d", th),
#                                  mu_hat >= th)
#
# ## Filtering barely changes anything here, and the reason is worth knowing:
# ## this simulation has very few genes near zero - about 1% below a mean count
# ## of 5 - so there is almost nothing to remove. In real RNA-seq 30-50% of
# ## annotated genes are effectively unexpressed in any given tissue, and
# ## dropping them removes a third or more of the multiplicity burden for free.
# ## That is where independent filtering earns its keep.
# ##
# ## What the table DOES show is that filtering does not break anything: FDP and
# ## power stay flat as the threshold rises, until it climbs high enough to
# ## start discarding real signal and power falls.
# ##
# ## It is legitimate only because mean count is independent of the p-value
# ## under the null. Filter on the observed fold change instead and FDR control
# ## collapses - the adjective in "INDEPENDENT filtering" is load-bearing.
# ##
# ## Finally, note the realised FDP sits above the nominal 0.05. BH controls the
# ## EXPECTED FDP over repeated experiments, not the FDP of your one gene list,
# ## and a mildly liberal test plus correlated genes pushes a single realisation
# ## up. Calibration is something to CHECK, not assume.
```

### Q4: Enrichment on your own result

Test each module for enrichment, naively and with a CAMERA-style variance
inflation factor (eq. 29.5). Two modules are genuinely up-regulated.

```{r q4}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# ## Work on a z-scale: turn each gene's p-value and sign into a z.
# z <- sign(best$lfc)*qnorm(pmax(pvals, 1e-300)/2, lower.tail = FALSE)
# logn <- log2(norm[kept_idx,, drop = FALSE] + 1)
# gene_pos <- setNames(seq_along(kept_idx), kept_idx)
# set_test <- function(members) {
#   idx <- gene_pos[as.character(intersect(members, kept_idx))]
#   idx <- idx[!is.na(idx)]
#   if (length(idx) < 10) return(NULL)
#   m <- length(idx); zs <- z[idx]
#   ## Naive: assumes the m genes are independent.
#   t_naive <- mean(zs)/(sd(z)/sqrt(m))
#   ## CAMERA: estimate the mean inter-gene correlation from residuals and
#   ## inflate the variance by VIF = 1 + (m-1)*rhobar (eq. 29.5).
#   sub <- logn[idx,, drop = FALSE]
#   resid <- sub - rowMeans(sub)
#   resid <- resid/(apply(resid, 1, sd) + 1e-9)
#   Cm <- cor(t(resid))
#   rho_bar <- (sum(Cm) - m)/(m*(m - 1))
#   vif <- 1 + (m - 1)*rho_bar
#   list(m = m, rho = rho_bar, vif = vif,
#        p_naive = 2*pnorm(abs(t_naive), lower.tail = FALSE),
#        p_cam = 2*pnorm(abs(t_naive/sqrt(max(vif, 1e-6))), lower.tail = FALSE))
# }
# cat(sprintf("  %-12s%6s%10s%8s%12s%12s%8s\n", "module", "size", "rho-bar",
#             "VIF", "naive p", "CAMERA p", "truth"))
# rows <- list()
# for (nm in names(D$modules)) {
#   r <- set_test(D$modules[[nm]])
#   if (is.null(r)) next
#   tr <- if (nm %in% c("module_01", "module_02")) "UP" else "-"
#   rows[[nm]] <- c(r$p_naive, r$p_cam, tr == "UP")
#   cat(sprintf("  %-12s%6d%10.3f%8.1f%12.2e%12.2e%8s\n", nm, r$m, r$rho,
#               r$vif, r$p_naive, r$p_cam, tr))
# }
# R <- do.call(rbind, rows); thresh <- 0.05/nrow(R)
# cat(sprintf("\n  Bonferroni over %d sets:\n", nrow(R)))
# cat(sprintf("    naive : %d significant, %d of them FALSE\n",
#             sum(R[, 1] < thresh), sum(R[, 1] < thresh & R[, 3] == 0)))
# cat(sprintf("    CAMERA: %d significant, %d of them FALSE\n",
#             sum(R[, 2] < thresh), sum(R[, 2] < thresh & R[, 3] == 0)))
#
# ## Every module is internally correlated by construction, so the naive test -
# ## which divides by sqrt(m) as if the genes were independent - understates the
# ## variance of the set mean and can fire on modules with no true signal.
# ## Dividing by sqrt(VIF) restores calibration, at the cost of some power.
# ## This is the SAME correction as the design effect in E1: correlated genes
# ## and correlated cells are one problem wearing two hats.
```

### Q5: The negative control

Run the pipeline on **permuted** group labels. A correct pipeline finds
essentially nothing - but only if the permutation respects the structure.

```{r q5}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# ## Permutation is expensive: every run refits two GLMs per gene. Use a random
# ## subset of genes - the discovery COUNT scales with it, but the comparison
# ## between permutation schemes, which is what we care about, does not.
# set.seed(3)
# perm_genes <- sort(sample(length(kept_idx), 700))
# pipeline <- function(labels, use_batch) {
#   p <- rep(1, length(perm_genes))
#   for (j in seq_along(perm_genes)) {
#     i <- perm_genes[j]; yv <- counts[kept_idx[i], ]
#     fam <- MASS::negative.binomial(theta = 1/disp_shrunk[i])
#     d1 <- if (use_batch) data.frame(yv, g = labels, batch) else data.frame(yv, g = labels)
#     d0 <- if (use_batch) data.frame(yv, batch) else data.frame(yv)
#     f1 <- try(glm(yv ~ ., fam, data = d1, offset = off), silent = TRUE)
#     f0 <- try(glm(yv ~ ., fam, data = d0, offset = off), silent = TRUE)
#     if (inherits(f1, "try-error") || inherits(f0, "try-error")) next
#     p[j] <- pchisq(f0$deviance - f1$deviance, 1, lower.tail = FALSE)
#   }
#   sum(p.adjust(p, "BH") < 0.05)
# }
# permute_free <- function() sample(group)
# permute_within_batch <- function() {
#   ## Shuffle labels WITHIN each batch, so the permuted design stays balanced
#   ## with respect to batch - the only valid null here.
#   lab <- group
#   for (b in unique(batch)) lab[batch == b] <- sample(group[batch == b])
#   lab
# }
# n_real <- pipeline(group, TRUE)
# cat(sprintf("  real labels, design ~ group + batch : %d discoveries among %d genes\n\n",
#             n_real, length(perm_genes)))
# cat(sprintf("  %-24s%-18s%7s%6s  runs\n", "permutation scheme", "design",
#             "mean", "max"))
# for (pn in c("free (ignores batch)", "within batch")) {
#   pf <- if (pn == "within batch") permute_within_batch else permute_free
#   for (dn in c("~ group", "~ group + batch")) {
#     hits <- replicate(6, pipeline(pf(), dn == "~ group + batch"))
#     cat(sprintf("  %-24s%-18s%7.1f%6d  %s\n", pn, dn, mean(hits), max(hits),
#                 paste(hits, collapse = " ")))
#   }
# }
#
# ## Read these against the discoveries the REAL labels produce. Every scheme
# ## collapses to a handful of genes, so the pipeline passes its negative
# ## control: it is not manufacturing signal out of nothing. That is the first
# ## thing to establish, and most published pipelines never establish it.
# ##
# ## Read the rows against each other. One is clearly the worst: permuting
# ## freely AND leaving batch out of the design. Fixing EITHER of those -
# ## permuting within batch, or putting batch in the design - brings it back
# ## down, and doing both is no better than doing one. The nuisance structure
# ## has to be accounted for somewhere; it does not much matter where. The
# ## differences are modest here only because batch was balanced within group
# ## by design; under a confounded design (E1) they would not be.
# ##
# ## Look at the MAXIMUM, not just the mean. Most permutations return zero, but
# ## the occasional one returns several - those are the runs where the shuffled
# ## labels lined up with batch. A negative control must be run MANY times: one
# ## permutation returning 0 proves nothing.
# ##
# ## Two rules worth keeping:
# ##   * permute within the strata that structure the data (batch, subject,
# ##     litter, plate) - the exchangeability rule from Module 07;
# ##   * put known nuisance variables in the DESIGN.
```

## Debrief

```{r debrief}
header("The generative truth")
cat(sprintf("  %d of %d genes are truly differential.\n", sum(D$is_de), nrow(counts)))
cat("  Dispersion follows 0.10 + 18/(mean + 12): low-count genes are far noisier.
  Modules 01 and 02 (50 genes each) are genuinely up-regulated; modules 03-10
  are correlated but NOT differential - they exist to catch a naive set test.
  All 10 modules respond to the processing BATCH, which is recorded and
  balanced within each arm - so it biases nothing, but it correlates genes and
  must appear in the design and in any permutation scheme.
  5 very high-expression genes are multiplied 60x in the treated group, which
  is real biology but creates composition bias in the total-count normaliser.

  The pipeline in order, and what each step protects you from:

    normalisation  -> composition bias (a median-based factor, never CPM)
    dispersion     -> the impossibility of estimating variance at n = 6
    testing        -> the mean-variance relationship of counts (NB, not t)
    filtering      -> multiplicity burden from genes with no power
    enrichment     -> inter-gene correlation (eq. 29.5 = eq. 1.8)
    permutation    -> everything you forgot\n")

png(file.path(OUT, "rnaseq_pipeline.png"), width = 1250, height = 420, res = 110)
par(mfrow = c(1, 3), mar = c(4.4, 4.4, 2.6, 1))
barplot(colSums(counts)/1e6, col = c("steelblue", "darkorange")[group + 1],
        xlab = "sample", ylab = "library size (millions)",
        main = "Library sizes vary 4-fold")
mu_p <- rowMeans(counts) + 1; var_p <- apply(counts, 1, var) + 1
plot(mu_p, var_p, pch = 16, cex = 0.25, col = adjustcolor("grey40", 0.4),
     log = "xy", xlab = "mean count", ylab = "variance",
     main = "Counts are overdispersed (eq. 20.2)")
gx <- 10^seq(0, 5, length.out = 50)
lines(gx, gx, lty = 2); lines(gx, gx + 0.3*gx^2, col = "firebrick", lwd = 2)
legend("topleft", c("Poisson (Var = mean)", "NB, disp = 0.3"),
       lty = c(2, 1), col = c("black", "firebrick"), lwd = c(1, 2), bty = "n",
       cex = 0.7)
frac <- apply(counts, 2, function(x) sum(sort(x, decreasing = TRUE)[1:5])/sum(x))
barplot(100*frac, col = c("steelblue", "darkorange")[group + 1],
        xlab = "sample", ylab = "% of library, top 5 genes",
        main = "Composition bias")
invisible(dev.off())
cat("Figure written to", file.path(OUT, "rnaseq_pipeline.png"), "\n")
```

## What to take away

1. Normalise with a **robust** size factor. Total-count normalisation breaks
   exactly when the biology is interesting.
2. You cannot estimate a per-gene variance at $n=6$. Shrink towards a trend.
3. Check that the test is calibrated before trusting the FDR; prefer the
   LRT to Wald at small $n$.
4. Independent filtering raises power for free; filtering on anything
   correlated with the test statistic destroys FDR control.
5. Set-level tests must account for inter-gene correlation.
6. **Permute the labels and re-run** - within the right strata.

**Next:** `E3_prediction_audit.R`
