---
title: "Exercise 4 - Answer a causal question from observational data"
output: html_document
---

**Curriculum link:** `stats.md` -> Topics 11, 15, 32, 34
**Core modules used:** 11, 32, 34, 28b

## The brief

A registry contains 4,000 patients. Some received drug A; the rest did not.
Does drug A improve the biomarker response?

You are given `severity`, `biomarker_0`, `age`, `treated`, a
post-treatment `toxicity` flag, and the outcome `biomarker_1`.

The temptation is to regress the outcome on everything available. Resist
it. Which variables you adjust for is a **causal** decision, not a
statistical one, and the data cannot make it for you. Adjusting for more
variables is not safer - one of the variables above will make your answer
*worse*.

## The causal structure

```
       severity --------------+
          |                   |
          v                   v
       treated ---------> biomarker_1
          |                   ^
          v                   |
       toxicity <----- frailty (UNMEASURED)

       age -> severity        (age also affects the outcome directly)
       biomarker_0 -> biomarker_1,  biomarker_0 -> treated
```

* `severity` and `biomarker_0` are **confounders** -> adjust.
* `toxicity` is a **collider**: treatment and the unmeasured `frailty`
  both cause it, and frailty also causes the outcome. Conditioning on it
  opens the path `treated -> toxicity <- frailty -> biomarker_1`.

````{r setup, message = FALSE}
MODULE_NAME <- "E4_causal_question"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
TRUE_ATE <- -1.50     # the estimand: average treatment effect on biomarker_1
````

## The data

````{r data}
make_registry <- function(seed = 3, n = 4000, ate = TRUE_ATE, unmeasured = 0) {
  set.seed(seed)
  age <- rnorm(n, 64, 11)
  ## Frailty is NEVER returned. It always affects the outcome and the toxicity
  ## risk; `unmeasured` additionally makes it affect TREATMENT, which is what
  ## turns it into a confounder (Q5).
  frailty <- rnorm(n)
  severity <- 0.045*(age - 64) + rnorm(n) + unmeasured*frailty
  biomarker_0 <- 10 + 1.4*severity + rnorm(n, 0, 1.2)
  ## CONFOUNDING BY INDICATION: sicker patients are more likely to be treated.
  lp_t <- -0.4 + 0.95*severity + 0.22*(biomarker_0 - 10) + unmeasured*frailty
  treated <- rbinom(n, 1, plogis(lp_t))
  biomarker_1 <- 8 + ate*treated + 1.7*severity + 0.55*(biomarker_0 - 10) +
    0.02*(age - 64) + 1.6*frailty + rnorm(n, 0, 1.2)
  ## COLLIDER: a common effect of treatment and frailty.
  toxicity <- rbinom(n, 1, plogis(-1.0 + 2.0*treated + 2.2*frailty))
  data.frame(age, severity, biomarker_0, treated, toxicity, biomarker_1)
}
header("The registry")
reg <- make_registry()
cat(sprintf("  n = %d, treated = %d (%.1f%%)\n", nrow(reg), sum(reg$treated),
            100*mean(reg$treated)))
cat(sprintf("\n  %-16s%12s%12s%12s\n", "variable", "untreated", "treated",
            "std. diff"))
for (v in c("age", "severity", "biomarker_0", "toxicity", "biomarker_1")) {
  a <- reg[[v]][reg$treated == 0]; b <- reg[[v]][reg$treated == 1]
  sd_p <- sqrt((var(a) + var(b))/2)
  cat(sprintf("  %-16s%12.3f%12.3f%12.3f\n", v, mean(a), mean(b),
              (mean(b) - mean(a))/sd_p))
}
cat("\n  A standardised difference above ~0.1 marks an imbalanced covariate.\n")
cat("  The treated patients are SICKER at baseline - that is the whole\n")
cat("  problem, and it is visible before any modelling.\n")
````

### Q1: The naive answer, and why it has the wrong sign

````{r q1}
````
````{r YOUR CODE HERE}

````
````{r SOLUTION (uncomment to check)}
# naive <- mean(reg$biomarker_1[reg$treated == 1]) -
#          mean(reg$biomarker_1[reg$treated == 0])
# cat(sprintf("  naive difference in means : %+.3f\n", naive))
# cat(sprintf("  true ATE                  : %+.3f\n", TRUE_ATE))
# cat(sprintf("  bias                      : %+.3f\n", naive - TRUE_ATE))
# cat("\n  The naive estimate has the WRONG SIGN. A drug that genuinely lowers\n")
# cat("  the biomarker appears to raise it, because the patients who received\n")
# cat("  it were sicker to begin with.\n")
#
# ## This is confounding by indication, the default state of every
# ## observational treatment comparison in medicine. Sample size does not help:
# ## with n = 4,000 the naive estimate is precisely and confidently wrong.
# ## Precision is not accuracy.
````

### Q2: Adjust for the confounders, two ways

Estimate the ATE by outcome regression and by IPTW (eq. 32.4), then check
covariate balance after weighting.

````{r q2}
````
````{r YOUR CODE HERE}

````
````{r SOLUTION (uncomment to check)}
# CONF <- c("severity", "biomarker_0", "age")
# ## (a) Outcome regression: the coefficient on `treated` is the ATE provided
# ##     the model is right and all confounders are in it.
# m_out <- lm(biomarker_1 ~ treated + severity + biomarker_0 + age, reg)
# ate_reg <- coef(m_out)["treated"]
# ## (b) IPTW: model treatment, then weight by 1/P(own treatment). This builds
# ##     a pseudo-population in which treatment is independent of the measured
# ##     confounders.
# ps_model <- glm(treated ~ severity + biomarker_0 + age, binomial, reg)
# ps <- fitted(ps_model)
# ## Stabilised weights: same estimand, much lower variance.
# p_t <- mean(reg$treated)
# w <- ifelse(reg$treated == 1, p_t/ps, (1 - p_t)/(1 - ps))
# wm <- function(x, ww) sum(x*ww)/sum(ww)
# i1 <- reg$treated == 1; i0 <- !i1
# ate_iptw <- wm(reg$biomarker_1[i1], w[i1]) - wm(reg$biomarker_1[i0], w[i0])
# naive <- mean(reg$biomarker_1[i1]) - mean(reg$biomarker_1[i0])
# cat(sprintf("  %-34s%10s%10s\n", "estimator", "ATE", "bias"))
# cat(sprintf("  %-34s%10.3f%10.3f\n", "naive difference", naive, naive - TRUE_ATE))
# cat(sprintf("  %-34s%10.3f%10.3f\n", "outcome regression", ate_reg, ate_reg - TRUE_ATE))
# cat(sprintf("  %-34s%10.3f%10.3f\n", "IPTW (stabilised)", ate_iptw, ate_iptw - TRUE_ATE))
# cat(sprintf("  %-34s%10.3f%10.3f\n", "TRUTH", TRUE_ATE, 0))
# cat(sprintf("\n  %-16s%18s%22s\n", "covariate", "std diff before",
#             "std diff after IPTW"))
# for (v in CONF) {
#   a <- reg[[v]][i0]; b <- reg[[v]][i1]
#   sd_p <- sqrt((var(a) + var(b))/2)
#   cat(sprintf("  %-16s%18.3f%22.3f\n", v, (mean(b) - mean(a))/sd_p,
#               (wm(b, w[i1]) - wm(a, w[i0]))/sd_p))
# }
# cat(sprintf("\n  weight diagnostics: max = %.2f, effective n = %.0f of %d\n",
#             max(w), sum(w)^2/sum(w^2), nrow(reg)))
#
# ## Both estimators recover the truth, because between them they use the SAME
# ## assumption: all confounders are measured and correctly modelled. They are
# ## not independent checks of each other. What IPTW adds is the BALANCE TABLE,
# ## which IS checkable: if the standardised differences do not collapse, your
# ## propensity model is wrong and you know it before looking at the outcome.
````

### Q3: The collider

Add `toxicity` to the outcome regression. It is measured, it is strongly
associated with the outcome, and adding it "controls for more". What
happens?

````{r q3}
````
````{r YOUR CODE HERE}

````
````{r SOLUTION (uncomment to check)}
# m_coll <- lm(biomarker_1 ~ treated + severity + biomarker_0 + age + toxicity, reg)
# ## Stratifying on the collider is the same mistake in another costume.
# m_strat <- lm(biomarker_1 ~ treated + severity + biomarker_0 + age,
#               subset(reg, toxicity == 0))
# cat(sprintf("  %-46s%10s%10s\n", "model", "ATE", "bias"))
# for (nm in list(c("correct: confounders only", coef(m_out)["treated"]),
#                 c("+ toxicity (a COLLIDER)", coef(m_coll)["treated"]),
#                 c("restricted to toxicity == 0", coef(m_strat)["treated"]),
#                 c("TRUTH", TRUE_ATE)))
#   cat(sprintf("  %-46s%10.3f%10.3f\n", nm[[1]], as.numeric(nm[[2]]),
#               as.numeric(nm[[2]]) - TRUE_ATE))
# cat(sprintf("\n  toxicity is strongly 'significant': p = %.2e,\n",
#             coef(summary(m_coll))["toxicity", "Pr(>|t|)"]))
# cat(sprintf("  and adding it RAISES R^2 from %.3f to %.3f.\n",
#             summary(m_out)$r.squared, summary(m_coll)$r.squared))
#
# ## Every statistical signal says to include toxicity: it is overwhelmingly
# ## significant, it raises R^2, it lowers the residual variance. And it makes
# ## the answer WRONG - here it overstates the benefit by about a third.
# ##
# ## Toxicity is a common effect of treatment and of frailty, and frailty also
# ## causes the outcome. Conditioning on a common effect induces an association
# ## between its causes (eq. 32.2), so the closed path
# ##     treated -> toxicity <- frailty -> biomarker_1
# ## is opened and its bias lands on the treatment coefficient. Restricting to
# ## non-toxic patients does the same damage - slightly worse, in fact - which
# ## is why "we excluded patients who experienced toxicity" is a red flag in a
# ## methods section.
# ##
# ## NOTHING IN THE DATA distinguishes a collider from a confounder. Only the
# ## knowledge that toxicity happened AFTER treatment, and shares a cause with
# ## the outcome, settles it. That is why you draw the DAG before you fit.
````

### Q4: Doubly robust estimation

AIPW (eq. 32.6) is consistent if **either** the outcome model or the
propensity model is correct. Demonstrate that by misspecifying each.

````{r q4}
````
````{r YOUR CODE HERE}

````
````{r SOLUTION (uncomment to check)}
# estimators <- function(df, out_formula, ps_formula) {
#   mo <- lm(out_formula, df)
#   mu1 <- predict(mo, transform(df, treated = 1))
#   mu0 <- predict(mo, transform(df, treated = 0))
#   e <- pmin(pmax(fitted(glm(ps_formula, binomial, df)), 0.02), 0.98)
#   a <- df$treated; y <- df$biomarker_1
#   pt <- mean(a)
#   ws <- ifelse(a == 1, pt/e, (1 - pt)/(1 - e))
#   ate_g <- mean(mu1 - mu0)
#   ate_ipw <- sum(y[a == 1]*ws[a == 1])/sum(ws[a == 1]) -
#              sum(y[a == 0]*ws[a == 0])/sum(ws[a == 0])
#   ## AIPW = g-formula estimate + a weighted residual correction.
#   aipw <- mean(a*(y - mu1)/e - (1 - a)*(y - mu0)/(1 - e) + mu1 - mu0)
#   c(g = ate_g, ipw = ate_ipw, aipw = aipw)
# }
# GOOD_OUT <- biomarker_1 ~ treated + severity + biomarker_0 + age
# BAD_OUT  <- biomarker_1 ~ treated                       # omits confounders
# GOOD_PS  <- treated ~ severity + biomarker_0 + age
# BAD_PS   <- treated ~ age                               # omits real drivers
# cat(sprintf("  %-16s%-14s%10s%10s%10s\n", "outcome model", "PS model",
#             "g-form", "IPTW", "AIPW"))
# for (o in list(c("correct", "GOOD_OUT"), c("WRONG", "BAD_OUT")))
#   for (p_ in list(c("correct", "GOOD_PS"), c("WRONG", "BAD_PS"))) {
#     r <- estimators(reg, get(o[2]), get(p_[2]))
#     cat(sprintf("  %-16s%-14s%10.3f%10.3f%10.3f\n", o[1], p_[1],
#                 r["g"], r["ipw"], r["aipw"]))
#   }
# cat(sprintf("  %-30s%10.3f%10.3f%10.3f\n", "TRUTH", TRUE_ATE, TRUE_ATE, TRUE_ATE))
#
# ## Read the AIPW column: it lands near the truth in the first three rows and
# ## fails only when BOTH models are wrong. That is double robustness - two
# ## chances to be right instead of one.
# ##
# ## It is not magic. AIPW does NOT protect against an unmeasured confounder,
# ## because neither model can contain a variable you do not have. It protects
# ## against getting the functional form wrong, not against the wrong variable
# ## list. Q5 is about the other failure.
````

### Q5: What would it take to explain the result away?

````{r q5}
````
````{r YOUR CODE HERE}

````
````{r SOLUTION (uncomment to check)}
# cat(sprintf("  %-14s%10s%11s%10s%10s\n", "unmeasured", "naive", "adjusted",
#             "AIPW", "bias"))
# for (u in c(0, 0.3, 0.6, 1.0, 1.6)) {
#   r <- make_registry(seed = 3, unmeasured = u)
#   nv <- mean(r$biomarker_1[r$treated == 1]) - mean(r$biomarker_1[r$treated == 0])
#   e <- estimators(r, GOOD_OUT, GOOD_PS)
#   cat(sprintf("  %-14.1f%10.3f%11.3f%10.3f%10.3f\n", u, nv, e["g"], e["aipw"],
#               e["aipw"] - TRUE_ATE))
# }
# cat(sprintf("  %-14s%10s%11.3f%10.3f%10.3f\n", "TRUTH", "", TRUE_ATE, TRUE_ATE, 0))
#
# ## Every method above adjusts only for what it can see. As the unmeasured
# ## frailty strengthens, ALL of them drift together - the doubly robust
# ## estimator no better than the rest. No amount of methodological
# ## sophistication substitutes for measuring the confounder.
# ##
# ## Since you can never rule this out, report how strong the unmeasured
# ## confounding would have to be. For a risk ratio RR the E-value is
# ##     E = RR + sqrt(RR*(RR - 1))                          (eq. 32.8)
# ## the minimum association (with BOTH treatment and outcome) an unmeasured
# ## confounder would need to explain the estimate away.
# for (rr in c(1.2, 1.5, 2.0, 3.0))
#   cat(sprintf("    RR = %.1f  ->  E-value = %.2f\n", rr, rr + sqrt(rr*(rr - 1))))
# cat("\n  An E-value of 1.7 means a confounder would need ~1.7-fold\n")
# cat("  associations with both treatment and outcome - plausible, so the\n")
# cat("  finding is fragile. An E-value of 5 would be hard to dismiss.\n")
````

## Debrief

````{r debrief}
header("The generative truth")
cat(sprintf("  TRUE average treatment effect = %+.2f on biomarker_1.\n\n", TRUE_ATE))
cat("  Variable roles - the only thing that matters, and the only thing the
  data cannot tell you:

    severity      CONFOUNDER  causes treatment and outcome    -> ADJUST
    biomarker_0   CONFOUNDER  causes treatment and outcome    -> ADJUST
    age           CONFOUNDER  via severity, plus direct       -> ADJUST
    toxicity      COLLIDER    caused by treatment AND frailty -> DO NOT ADJUST
    frailty       UNMEASURED  causes the outcome and toxicity -> cannot adjust
                              (and, in Q5, treatment too -> then a confounder)

  The four lessons, in order of how often they are violated:

    1. The naive comparison had the WRONG SIGN. Confounding by indication is
       the default, not the exception.
    2. Adjusting for MORE variables is not safer. Adding the collider made a
       correct analysis wrong, while improving every fit statistic.
    3. Doubly robust estimation gives two chances at the functional form. It
       gives no protection against a variable you did not measure.
    4. Because you can never prove there is no unmeasured confounding,
       quantify how much would be needed. That is what an E-value is for.

  Which variables to adjust for is decided by the DAG - by what you know
  about how the data were generated - and never by stepwise selection, by
  p-values, or by which model fits best.\n")

png(file.path(OUT, "causal_question.png"), width = 1200, height = 420, res = 110)
par(mfrow = c(1, 3), mar = c(4.4, 4.2, 2.6, 1))
plot(reg$severity, reg$biomarker_1, pch = 16, cex = 0.25,
     col = adjustcolor(c("steelblue", "darkorange")[reg$treated + 1], 0.35),
     xlab = "severity (confounder)", ylab = "biomarker_1",
     main = "Treated patients start sicker")
legend("topleft", c("untreated", "treated"), col = c("steelblue", "darkorange"),
       pch = 16, bty = "n", cex = 0.75)
ps <- fitted(glm(treated ~ severity + biomarker_0 + age, binomial, reg))
h0 <- hist(ps[reg$treated == 0], breaks = 30, plot = FALSE)
h1 <- hist(ps[reg$treated == 1], breaks = 30, plot = FALSE)
plot(h0, col = adjustcolor("steelblue", 0.6), border = "white", freq = FALSE,
     xlab = "propensity score", main = "Overlap (positivity) is adequate",
     ylim = c(0, max(h0$density, h1$density)))
plot(h1, col = adjustcolor("darkorange", 0.6), border = "white", freq = FALSE,
     add = TRUE)
rr <- seq(1.01, 4, length.out = 100)
plot(rr, rr + sqrt(rr*(rr - 1)), type = "l", lwd = 2, col = "firebrick",
     xlab = "observed risk ratio", ylab = "E-value",
     main = "Eq. (32.8): how much would it take?")
invisible(dev.off())
cat("Figure written to", file.path(OUT, "causal_question.png"), "\n")
````

## What to take away

1. Draw the DAG **before** fitting. Adjustment sets come from the DAG.
2. Confounding by indication routinely flips the sign.
3. A collider is statistically indistinguishable from a confounder.
4. Never adjust for anything measured **after** treatment.
5. Doubly robust is not robust to unmeasured confounding.
6. Report an E-value with every observational estimate.

**This is the final exercise.** Return to `stats.md` for the theory, or to
`statsR/bioinformatics/` to apply it.
