---
title: "Foundations 2 - Probability, independence, and conditional probability"
output: html_document
---

**Curriculum link:** `stats.md` -> Part 0, §0.4, equations (0.1)-(0.9)
**Assumes:** `F1`.

## Why you need this

Your sample could have come out otherwise. Probability is the mathematics
of "could have come out otherwise", which is why it sits underneath every
p-value, confidence interval and error rate in this course.

You need surprisingly little of it. This module covers all of it.

```{r setup, message = FALSE}
MODULE_NAME <- "F2_probability_basics"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
set.seed(102)
```

## 1. A probability is a long-run frequency

```{r longrun}
header("1. Probability is what happens in the long run")
flips <- rbinom(20000, 1, 0.5)          # 0 = tails, 1 = heads
cat(sprintf("  %-26s%20s\n", "after this many flips", "proportion heads"))
for (n in c(10, 100, 1000, 10000, 20000))
  cat(sprintf("  %-26s%20.4f\n", format(n, big.mark = ","), mean(flips[1:n])))
cat("
  The proportion wanders a lot early on and settles down later. That is all a
  probability is: the value the proportion settles to.

  Note what it does NOT tell you - whether the next flip is heads. A
  probability is a statement about a PROCESS repeated, never about a single
  outcome. That is exactly the distinction that makes confidence intervals and
  p-values so easy to misread (F5).\n")
```

## 2. The two combining rules, eq. (0.2)-(0.3)

```{r rules}
header("2. 'or' adds; 'and' multiplies (but only if independent)")
die <- sample(1:6, 200000, replace = TRUE)
cat("  OR rule, eq. (0.2) - for outcomes that cannot both happen:\n")
cat(sprintf("    P(1) = %.4f,  P(2) = %.4f\n", mean(die == 1), mean(die == 2)))
cat(sprintf("    P(1) + P(2)   = %.4f   <- the rule\n", mean(die == 1) + mean(die == 2)))
cat(sprintf("    P(1 or 2)     = %.4f   <- simulated\n", mean(die %in% c(1, 2))))
a <- rbinom(200000, 1, 0.5); b <- rbinom(200000, 1, 0.5)
cat("\n  AND rule, eq. (0.3) - for INDEPENDENT events:\n")
cat(sprintf("    P(A heads) = %.4f,  P(B heads) = %.4f\n", mean(a), mean(b)))
cat(sprintf("    product       = %.4f   <- the rule\n", mean(a)*mean(b)))
cat(sprintf("    P(both heads) = %.4f   <- simulated\n", mean(a == 1 & b == 1)))
```

## 3. Independence, and what happens when it fails

```{r independence}
header("3. When independence fails, multiplying gives the wrong answer")
## Two coins welded together: they always land the same way. Each is still a
## perfectly fair coin - P(heads) = 0.5 for each, separately.
welded <- rbinom(200000, 1, 0.5)
cc <- welded; dd <- welded
cat("  Two coins that are NOT independent (they always match):\n")
cat(sprintf("    P(C heads) = %.4f   <- still a fair coin\n", mean(cc)))
cat(sprintf("    P(D heads) = %.4f   <- still a fair coin\n", mean(dd)))
cat(sprintf("    product (what eq. 0.3 would give) = %.4f\n", mean(cc)*mean(dd)))
cat(sprintf("    P(both heads), actually           = %.4f\n", mean(cc == 1 & dd == 1)))
cat(sprintf("    the rule is wrong by a factor of  %.1fx\n",
            mean(cc == 1 & dd == 1)/(mean(cc)*mean(dd))))
cat("
  Each coin is individually fair. The rule still fails, because the rule was
  never about the individual coins - it was about their INDEPENDENCE.

  Now scale that up. This is the whole of pseudoreplication.\n")

## Two designs, both producing 120 numbers:
##   (a) 120 genuinely independent measurements
##   (b) 6 mice x 20 cells, where the 20 cells from one mouse share that
##       mouse's own level - so they are NOT independent of each other.
N_UNITS <- 6; N_SUB <- 20; SD_UNIT <- 1; SD_CELL <- 1; REPS <- 4000
indep_means <- numeric(REPS); clust_means <- numeric(REPS)
clust_naive <- numeric(REPS)
for (i in seq_len(REPS)) {
  indep_means[i] <- mean(rnorm(N_UNITS*N_SUB, 0, SD_CELL))
  mouse_level <- rnorm(N_UNITS, 0, SD_UNIT)
  cells <- rep(mouse_level, each = N_SUB) + rnorm(N_UNITS*N_SUB, 0, SD_CELL)
  clust_means[i] <- mean(cells)
  clust_naive[i] <- sd(cells)/sqrt(N_UNITS*N_SUB)
}
cat(sprintf("\n  %-40s%6s%22s\n", "design", "rows", "TRUE SD of the mean"))
cat(sprintf("  %-40s%6d%22.4f\n", "(a) 120 independent measurements",
            N_UNITS*N_SUB, sd(indep_means)))
cat(sprintf("  %-40s%6d%22.4f\n", "(b) 6 mice x 20 cells",
            N_UNITS*N_SUB, sd(clust_means)))
cat(sprintf("\n  For design (b), what s/sqrt(120) CLAIMS : %8.4f\n", mean(clust_naive)))
cat(sprintf("  What the uncertainty actually IS       : %8.4f\n", sd(clust_means)))
cat(sprintf("  The naive formula is too small by      : %8.1fx\n",
            sd(clust_means)/mean(clust_naive)))
cat(sprintf("  Treating the MOUSE as the unit gives   : %8.4f\n",
            sqrt(SD_UNIT^2 + SD_CELL^2/N_SUB)/sqrt(N_UNITS)))
cat("
  Both designs produce 120 numbers. Design (b) is far less informative,
  because its 120 rows contain only 6 independent facts about the population.

  The naive formula divides by sqrt(120) as if there were 120. The last line
  treats the mouse as the unit and divides by sqrt(6) instead - and that
  number matches the truth.

  This is not a subtle effect. In exercise E1 it turns p = 0.06 into
  p = 2e-12. Independence is not a technicality - it is the load-bearing
  assumption of applied statistics, and eq. (0.3) is where it enters.\n")
```

## 4. Conditional probability, eq. (0.4)

```{r conditional}
header("4. P(A given B) is not P(B given A)")
N <- 1000000
PREVALENCE <- 0.01; SENSITIVITY <- 0.99; SPECIFICITY <- 0.95
diseased <- runif(N) < PREVALENCE
positive <- ifelse(diseased, runif(N) < SENSITIVITY, runif(N) < (1 - SPECIFICITY))
cat(sprintf("  A disease %.0f%% of people have.\n", 100*PREVALENCE))
cat(sprintf("  A test that is %.0f%% sensitive and %.0f%% specific.\n\n",
            100*SENSITIVITY, 100*SPECIFICITY))
cat(sprintf("  %-46s%10s\n", "quantity", "value"))
cat(sprintf("  %-46s%10.3f\n", "P(test positive | you have it)", mean(positive[diseased])))
cat(sprintf("  %-46s%10.3f   <- !\n", "P(you have it | test positive)",
            mean(diseased[positive])))
cat(sprintf("\n  of %s positive tests, only %s are real cases\n",
            format(sum(positive), big.mark = ","),
            format(sum(positive & diseased), big.mark = ",")))
cat("
  The test is 99% accurate at detecting the disease, and yet a positive result
  means you probably do NOT have it. Nothing is wrong with the test. Healthy
  people vastly outnumber sick ones, so even a small false-positive RATE
  produces a large NUMBER of false positives.

  This asymmetry - P(A|B) is not P(B|A) - is the most misread idea in
  statistics, and it is exactly the mistake people make with p-values:

    a p-value is   P(data this extreme | no real effect)
    people read it P(no real effect | data this extreme)

  Those are as different as the two rows above. F5 returns to this.\n")
```

## 5. Expectation and variance, eq. (0.5)-(0.7)

```{r moments}
header("5. Expectation (where it sits) and variance (how spread out)")
MU <- 10; SIGMA <- 3
x <- rnorm(500000, MU, SIGMA)
cat(sprintf("  %-40s%-24s%10s\n", "quantity", "formula", "value"))
cat(sprintf("  %-40s%-24s%10.4f\n", "expected value E[X] = mu", "mean of X", mean(x)))
cat(sprintf("  %-40s%-24s%10.4f\n", "variance = E[(X - mu)^2]",
            "mean squared distance", mean((x - MU)^2)))
cat(sprintf("  %-40s%-24s%10.4f\n", "standard deviation", "sqrt(variance)",
            sqrt(mean((x - MU)^2))))
cat(sprintf("
  Why square the distances (eq. 0.6)? Because without squaring, the positives
  and negatives cancel exactly:

    mean of (X - mu)     = %9.5f   <- always ~0, useless
    mean of |X - mu|     = %9.5f   <- works, but awkward algebra
    mean of (X - mu)^2   = %9.5f   <- the variance

  Squaring makes the units wrong (squared grams), so we take the square root
  at the end and get the STANDARD DEVIATION - a spread in the original units.

  Rule of thumb for a bell-shaped variable: about 68%% of values fall within
  1 SD of the mean and 95%% within 2 SD. Check it:
    within 1 SD: %.1f%%
    within 2 SD: %.1f%%\n",
  mean(x - MU), mean(abs(x - MU)), mean((x - MU)^2),
  100*mean(abs(x - MU) < SIGMA), 100*mean(abs(x - MU) < 2*SIGMA)))
```

## 6. The variance rule that explains everything, eq. (0.9)

```{r varsum}
header("6. Variances add - for independent things (0.9)")
n <- 200000
X <- rnorm(n); Y <- rnorm(n); Z <- X
cat(sprintf("  %-34s%16s%18s\n", "", "Var(X)+Var(Y)", "Var(X+Y) actual"))
cat(sprintf("  %-34s%16.4f%18.4f\n", "X, Y independent", var(X) + var(Y), var(X + Y)))
cat(sprintf("  %-34s%16.4f%18.4f   <- rule fails\n", "X, Z identical (dependent)",
            var(X) + var(Z), var(X + Z)))
cat("\n  Now the consequence. Average n independent values and the variance\n")
cat("  of the average is divided by n, so its SD is divided by sqrt(n):\n\n")
cat(sprintf("  %-8s%18s%20s%16s\n", "n", "SD of one value", "SD of the average",
            "sigma/sqrt(n)"))
for (k in c(1, 4, 16, 64, 256)) {
  avgs <- replicate(20000, mean(rnorm(k)))
  cat(sprintf("  %-8d%18.4f%20.4f%16.4f\n", k, 1, sd(avgs), 1/sqrt(k)))
}
cat("
  The last two columns match. That is eq. (0.11), the STANDARD ERROR, and it
  falls straight out of eq. (0.9).

  Two consequences worth carrying with you:

    * To halve your uncertainty you must QUADRUPLE your sample. Precision is
      expensive, and gets more expensive the more you have.

    * The rule required INDEPENDENCE. If your n values are 20 cells from each
      of 6 mice, the variance does not divide by 120, and dividing by 120
      anyway is exactly the error in section 3.

  Everything in F4 is this table.\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "probability_basics.png"), width = 1250, height = 420, res = 110)
par(mfrow = c(1, 3), mar = c(4.4, 4.4, 2.6, 1))
running <- cumsum(flips)/seq_along(flips)
plot(running, type = "l", log = "x", col = "steelblue", ylim = c(0, 1),
     xlab = "number of flips", ylab = "proportion heads",
     main = "A probability is a long-run frequency")
abline(h = 0.5, col = "firebrick", lty = 2, lwd = 1.5)
barplot(c(sum(positive & diseased), sum(positive & !diseased)),
        names.arg = c("true\npositives", "false\npositives"),
        col = c("seagreen", "firebrick"), ylab = "people testing positive",
        main = sprintf("P(disease | positive) = %.2f", mean(diseased[positive])))
ks <- c(1, 2, 4, 8, 16, 32, 64, 128, 256)
sds <- sapply(ks, function(k) sd(replicate(8000, mean(rnorm(k)))))
plot(ks, sds, type = "b", pch = 16, log = "xy", col = "steelblue",
     xlab = "n averaged", ylab = "SD of the average",
     main = "Eq. (0.9) => eq. (0.11)")
lines(ks, 1/sqrt(ks), col = "firebrick", lty = 2, lwd = 2)
legend("bottomleft", c("simulated", "sigma/sqrt(n)"),
       col = c("steelblue", "firebrick"), lty = c(1, 2), lwd = 2, bty = "n", cex = 0.75)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "probability_basics.png"), "\n")
```

# PROBLEMS

### Problem 1: The base rate

```{r problem1}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# cat(sprintf("  %-14s%18s%18s\n", "prevalence", "P(pos | disease)", "P(disease | pos)"))
# for (prev in c(0.001, 0.01, 0.10, 0.30, 0.60)) {
#   dis <- runif(400000) < prev
#   pos <- ifelse(dis, runif(400000) < 0.99, runif(400000) < 0.05)
#   cat(sprintf("  %-13.1f%%%18.3f%18.3f\n", 100*prev, mean(pos[dis]), mean(dis[pos])))
# }
#
# ## The middle column never moves: the TEST has not changed. The right column
# ## swings from almost 0 to almost 1, purely because of how common the disease
# ## is in the population being tested.
# ##
# ## The same test is nearly useless for screening the general population and
# ## very informative in a high-risk clinic. This is why screening programmes
# ## target risk groups, and why "the test is 99% accurate" means nothing on
# ## its own.
# ##
# ## The statistical version: your interpretation of a positive RESULT depends
# ## on how plausible the thing was BEFORE you tested. That is Bayes' theorem,
# ## and it is why p = 0.04 means something quite different for a
# ## pre-registered hypothesis than for the 500th gene you looked at.
```

### Problem 2: How much does dependence cost you?

```{r problem2}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# n_groups <- 6; per_group <- 20
# cat(sprintf("  %-12s%8s%18s%20s%14s\n", "group SD", "ICC",
#             "true SD of mean", "naive s/sqrt(120)", "too small by"))
# for (sd_group in c(0, 0.3, 0.7, 1.5)) {
#   tr <- numeric(2000); nv <- numeric(2000)
#   for (i in 1:2000) {
#     g <- rnorm(n_groups, 0, sd_group)
#     y <- rep(g, each = per_group) + rnorm(n_groups*per_group, 0, 1)
#     tr[i] <- mean(y); nv[i] <- sd(y)/sqrt(length(y))
#   }
#   icc <- sd_group^2/(sd_group^2 + 1)
#   cat(sprintf("  %-12.1f%8.2f%18.4f%20.4f%13.1fx\n", sd_group, icc,
#               sd(tr), mean(nv), sd(tr)/mean(nv)))
# }
#
# ## With no group effect the two columns agree - the 120 values really are
# ## independent. As the group effect grows they diverge fast, and the naive
# ## formula understates the uncertainty several-fold.
# ##
# ## The ICC column is the fraction of total variance that lives BETWEEN
# ## groups. Even an ICC of 0.1 - which sounds negligible - does real damage,
# ## because the penalty depends on the GROUP SIZE too: it is 1 + (m-1) x ICC,
# ## the "design effect" of eq. (1.8). With m = 20 cells per mouse, an ICC of
# ## 0.1 means each mouse contributes about a third of what you thought.
# ##
# ## The fix is never a better formula. It is to analyse at the level of the
# ## unit you randomised (F1, Problem 3) or to model the grouping explicitly
# ## with a mixed model (Topic 14).
```

### Problem 3: Does the rule of thumb survive a skewed variable?

```{r problem3}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# normal_x <- rnorm(200000, 100, 20)
# skewed_x <- rlnorm(200000, log(100), 0.9)
# cat(sprintf("  %-22s%9s%9s%14s%14s%16s\n", "variable", "mean", "median",
#             "within 1 SD", "within 2 SD", "below the mean"))
# for (nm in c("normal", "skewed (lognormal)")) {
#   v <- if (nm == "normal") normal_x else skewed_x
#   m <- mean(v); s <- sd(v)
#   cat(sprintf("  %-22s%9.1f%9.1f%13.1f%%%13.1f%%%15.1f%%\n", nm, m, median(v),
#               100*mean(abs(v - m) < s), 100*mean(abs(v - m) < 2*s), 100*mean(v < m)))
# }
# lv <- log(skewed_x); m <- mean(lv); s <- sd(lv)
# cat("\n  after taking logs of the skewed variable:\n")
# cat(sprintf("  %-22s%9.1f%9.1f%13.1f%%%13.1f%%%15.1f%%\n", "log(skewed)", m,
#             median(lv), 100*mean(abs(lv - m) < s), 100*mean(abs(lv - m) < 2*s),
#             100*mean(lv < m)))
#
# ## For the skewed variable the rule of thumb breaks, and note the last
# ## column: far MORE than half the values sit below the mean. A few large
# ## values drag the mean above the typical value, so "the average expression"
# ## is not a description of a typical gene.
# ##
# ## Taking logs restores the bell shape, and with it the rule of thumb and the
# ## entire standard toolkit. This is why expression, concentrations and
# ## survival times are analysed on a log scale - not tradition, but because
# ## the log turns multiplicative, skewed variation into additive, symmetric
# ## variation. Topic 2 covers when a transformation is appropriate and what it
# ## does to the interpretation of your effect size.
```

## What to take away

1. A probability is a **long-run frequency**, never a statement about one
   outcome.
2. "Or" adds; "and" multiplies **only under independence**.
3. **Independence is the assumption biology breaks**, and eq. (0.3) is
   where the damage begins.
4. $P(A \mid B) \neq P(B \mid A)$ - the p-value misreading.
5. Variance is the average **squared** distance from the mean.
6. Variances **add for independent things**: which is why averaging works
   and why $\mathrm{SE} = \sigma/\sqrt{n}$.

**Next:** `F3_distributions.R`
