---
title: "Foundations 3 - Distributions"
output: html_document
---

**Curriculum link:** `stats.md` -> Part 0, §0.5, equation (0.10)
**Assumes:** `F1`, `F2`.

## What a distribution is

A **distribution** answers: which values does this variable take, and how
often? The mean and SD are two summaries of it; the distribution is the
whole picture.

You do not need to memorise formulas. You need to recognise **which
situation produces which shape**, because that decides which method is
valid. Each distribution below corresponds to a *mechanism*, and naming the
mechanism usually names the method.

```{r setup, message = FALSE}
MODULE_NAME <- "F3_distributions"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
set.seed(103)

## A small text histogram, so you can see shapes without opening a plot.
sparkline <- function(v, bins = 28, height = 7) {
  h <- hist(v, breaks = bins, plot = FALSE)
  cnt <- h$counts/max(h$counts)
  for (lev in height:1)
    cat("    |", paste(ifelse(cnt*height >= lev - 0.5, "#", " "), collapse = ""), "\n", sep = "")
  cat("    +", strrep("-", length(cnt)), "\n", sep = "")
  cat(sprintf("     %-*.1f%*.1f\n", length(cnt) %/% 2, min(h$breaks),
              length(cnt) - length(cnt) %/% 2, max(h$breaks)))
}
```

## 1. The normal distribution: what averages look like

```{r normal}
header("1. Normal: the bell curve")
MU <- 100; SIGMA <- 15
x <- rnorm(200000, MU, SIGMA)
cat(sprintf("  X ~ Normal(mu = %d, sigma = %d)\n", MU, SIGMA))
sparkline(x)
cat(sprintf("\n  %-12s%10s%12s\n", "within", "theory", "observed"))
for (k in 1:3) {
  theory <- c(0.6827, 0.9545, 0.9973)[k]
  cat(sprintf("  %-12s%10.4f%12.4f\n", paste(k, "sigma"), theory,
              mean(abs(x - MU) < k*SIGMA)))
}
cat("
  The 68 / 95 / 99.7 rule. Worth memorising - it turns any normal-ish number
  into an instant sanity check.

  The normal distribution is everywhere NOT because biological measurements
  are naturally bell-shaped (mostly they are not) but because AVERAGES are,
  whatever the raw data looks like. That is the central limit theorem (F4).\n")
```

## 2. Binomial: counting successes out of $n$ tries

```{r binomial}
header("2. Binomial: how many out of n?")
n_trials <- 50; p_success <- 0.30
b <- rbinom(200000, n_trials, p_success)
cat(sprintf("  X ~ Binomial(n = %d, p = %.2f)\n", n_trials, p_success))
sparkline(b)
cat(sprintf("\n  %-28s%-16s%10s%12s\n", "quantity", "formula", "theory", "observed"))
cat(sprintf("  %-28s%-16s%10.3f%12.3f\n", "mean", "n*p", n_trials*p_success, mean(b)))
cat(sprintf("  %-28s%-16s%10.3f%12.3f\n", "variance", "n*p*(1-p)",
            n_trials*p_success*(1-p_success), var(b)))
cat("
  Note the variance formula: it is LARGEST at p = 0.5 and shrinks towards 0 as
  p approaches 0 or 1. A proportion near 0 or 1 is intrinsically less variable
  than one near a half - which is why the uncertainty on a proportion is not
  the same at every level, and why proportions get their own model (Topic 13)
  rather than being fed to a t-test.\n")
for (p_ in c(0.02, 0.5, 0.98))
  cat(sprintf("    p = %-5.2f variance = %6.2f\n", p_, n_trials*p_*(1-p_)))
```

## 3. Poisson: counting events in a window, eq. (0.10)

```{r poisson}
header("3. Poisson: mean equals variance (0.10)")
for (lam in c(2, 10, 100)) {
  po <- rpois(200000, lam)
  cat(sprintf("  lambda = %-5d mean = %8.3f   variance = %8.3f   ratio = %.3f\n",
              lam, mean(po), var(po), var(po)/mean(po)))
}
sparkline(rpois(200000, 10))
cat("
  mean = variance is the Poisson's signature, and its weakness. It leaves NO
  freedom: tell me the mean and I have told you the spread.

  Notice also that as lambda grows the shape becomes symmetric and bell-like.
  A Poisson with a large mean is well approximated by a normal - which is why
  high-count genes can be treated more casually than low-count ones.\n")
```

## 4. Negative binomial: counts in the real world

```{r negbin}
header("4. Negative binomial: Poisson plus biological variability")
MEAN <- 100
cat(sprintf("  All of these have the same mean (%d). Only the spread differs.\n\n", MEAN))
cat(sprintf("  %-40s%9s%11s%10s\n", "model", "mean", "variance", "var/mean"))
pois <- rpois(200000, MEAN)
cat(sprintf("  %-40s%9.1f%11.1f%10.2f\n", "Poisson (technical noise only)",
            mean(pois), var(pois), var(pois)/mean(pois)))
for (disp in c(0.05, 0.2, 0.5)) {
  ## A negative binomial IS a Poisson whose rate varies between samples.
  ## Simulate it exactly that way, so the mechanism is visible.
  rate <- rgamma(200000, shape = 1/disp, scale = MEAN*disp)
  nb <- rpois(200000, rate)
  cat(sprintf("  %-40s%9.1f%11.1f%10.2f\n",
              sprintf("Neg. binomial, dispersion = %.2f", disp),
              mean(nb), var(nb), var(nb)/mean(nb)))
}
cat(sprintf("
  Var = mu + dispersion * mu^2. The first term is counting noise; the second
  is biological variability between samples, and it dominates at high counts.

  This is THE reason RNA-seq is analysed with DESeq2 or edgeR rather than a
  t-test or a Poisson model. Use Poisson on data that is really negative
  binomial and you understate the variance badly - at mean %d with dispersion
  0.2, by a factor of %.0f. Every p-value inherits that error.

  Topic 20 is about estimating that dispersion when you have 3 samples, which
  is the central technical difficulty of differential expression.\n",
  MEAN, (MEAN + 0.2*MEAN^2)/MEAN))
```

## 5. Uniform: and the most useful diagnostic in this course

```{r uniform}
header("5. Uniform: p-values under the null are flat")
null_p <- replicate(20000, t.test(rnorm(10), rnorm(10))$p.value)
cat("  20,000 t-tests on data with NO real difference:\n")
sparkline(null_p)
cat(sprintf("\n  %-18s%8.4f   (should be 0.05)\n", "p < 0.05", mean(null_p < 0.05)))
cat(sprintf("  %-18s%8.4f   (should be 0.10)\n", "p < 0.10", mean(null_p < 0.10)))
cat(sprintf("  %-18s%8.4f   (should be 0.50)\n", "p < 0.50", mean(null_p < 0.50)))
real_p <- replicate(20000, t.test(rnorm(10, 0.9), rnorm(10))$p.value)
cat("\n  The same test when there IS a real difference:\n")
sparkline(real_p)
cat("
  FLAT means the test is behaving. SPIKED AT ZERO means there is signal.

  That gives you a diagnostic you can use on any analysis: plot a histogram of
  all your p-values. If you expect mostly nulls and it is NOT flat, something
  is wrong - unmodelled batch, non-independence, a broken assumption. If it
  slopes upward towards 1, your test is conservative. A p-value histogram is
  the cheapest and most informative plot in genomics (Topic 8).\n")
```

## 6. Distributions of *statistics*, not measurements

```{r tdist}
header("6. t, chi-squared and F describe test statistics")
cat("  The t distribution is the normal's heavier-tailed cousin. It is what\n")
cat("  you get when you must ESTIMATE sigma instead of knowing it.\n\n")
cat(sprintf("  %-8s%14s%24s\n", "df", "P(|t| > 2)", "P(|t| > 2) if normal"))
for (df in c(2, 5, 10, 30, 1000))
  cat(sprintf("  %-8d%14.4f%24.4f\n", df, 2*pt(2, df, lower.tail = FALSE),
              2*pnorm(2, lower.tail = FALSE)))
cat("
  With few degrees of freedom the tails are FATTER - extreme values are more
  likely than a normal would suggest. That is exactly right: when you have
  little data, your estimate of sigma is itself unreliable, and the t
  distribution builds in that extra doubt.

  As df grows, t converges on the normal. By df = 30 the difference barely
  matters, which is the origin of the folklore that 'n = 30 is enough'.

  The practical consequence: with a small sample, the critical value for a 95%
  interval is not 1.96 but something larger, so your interval is WIDER. Small
  studies are automatically made more humble - if you use the right
  distribution.\n")
```

## 7. Skew, and why biology uses logs

```{r skew}
header("7. Skewed data, and what the log transform does")
expr <- rlnorm(200000, log(50), 1.1)
cat("  A typical gene expression variable (right-skewed):\n")
sparkline(expr)
cat(sprintf("\n  mean   = %8.1f\n  median = %8.1f   <- much less than the mean\n",
            mean(expr), median(expr)))
cat(sprintf("  %.1f%% of values are BELOW the mean\n", 100*mean(expr < mean(expr))))
cat("\n  The same data after taking logs:\n")
sparkline(log2(expr))
cat(sprintf("\n  mean   = %8.2f\n  median = %8.2f   <- now they agree\n",
            mean(log2(expr)), median(log2(expr))))
cat(sprintf("  %.1f%% of values are below the mean\n",
            100*mean(log2(expr) < mean(log2(expr)))))
cat("
  For the raw variable, 'the mean' is not a typical value - a minority of
  large values drags it above most of the data. Reporting mean +/- SD here
  would be actively misleading.

  The log fixes it, and not by accident. Biological quantities tend to vary
  MULTIPLICATIVELY - a treatment doubles expression, it does not add 40 units
  - and the log turns multiplication into addition. That converts skewed,
  proportional variation into symmetric, additive variation, which is exactly
  what the standard toolkit is built for.

  This is why expression is analysed as log2 fold change, why concentrations
  are logged, and why '2-fold change' is a more natural unit than '+50'.\n")
```

## 8. Figure

```{r figure}
png(file.path(OUT, "distributions.png"), width = 1250, height = 780, res = 110)
par(mfrow = c(2, 3), mar = c(4.2, 4.2, 2.6, 1))
hist(x, breaks = 70, col = "steelblue", border = NA, main = "Normal: 68 / 95 rule",
     xlab = "value", freq = FALSE)
for (k in 1:2) abline(v = MU + c(-1, 1)*k*SIGMA, col = "firebrick", lty = 2)
hist(b, breaks = 30, col = "seagreen", border = NA, freq = FALSE,
     main = sprintf("Binomial(n=%d, p=%.1f)", n_trials, p_success), xlab = "successes")
plot(0:60, dpois(0:60, 10), type = "h", col = "darkorange", lwd = 2,
     main = "Poisson: mean = variance", xlab = "count", ylab = "probability")
lines(0:60, dpois(0:60, 2), type = "h", col = "steelblue", lwd = 2)
legend("topright", c("lambda=2", "lambda=10"), col = c("steelblue", "darkorange"),
       lwd = 2, bty = "n", cex = 0.7)
grid_ <- 0:320
plot(grid_, dpois(grid_, MEAN), type = "l", lwd = 2, col = "steelblue",
     main = "Same mean, different spread", xlab = "count", ylab = "probability")
for (i in seq_along(c(0.05, 0.3))) {
  disp <- c(0.05, 0.3)[i]
  lines(grid_, dnbinom(grid_, size = 1/disp, mu = MEAN), lwd = 2,
        col = c("darkorange", "firebrick")[i])
}
legend("topright", c("Poisson", "NB 0.05", "NB 0.3"),
       col = c("steelblue", "darkorange", "firebrick"), lwd = 2, bty = "n", cex = 0.7)
hist(null_p, breaks = 40, col = "grey70", border = NA, freq = FALSE,
     main = "p-values: flat under the null", xlab = "p-value")
hist(real_p, breaks = 40, col = adjustcolor("firebrick", 0.6), border = NA,
     freq = FALSE, add = TRUE)
abline(h = 1, lty = 2)
hist(log2(expr), breaks = 70, col = "seagreen", border = NA, freq = FALSE,
     main = "Skewed data, after log2", xlab = "log2(expression)")
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "distributions.png"), "\n")
```

# PROBLEMS

### Problem 1: How wrong is Poisson on real count data?

```{r problem1}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# poisson_test <- function(a, b) {
#   ## Compare two groups of counts ASSUMING Poisson (variance = mean).
#   ma <- mean(a); mb <- mean(b)
#   se <- sqrt(ma/length(a) + mb/length(b))
#   2*pnorm(abs(ma - mb)/max(se, 1e-9), lower.tail = FALSE)
# }
# cat(sprintf("  %-20s%10s%22s\n", "true dispersion", "var/mean", "false positive rate"))
# for (disp in c(0, 0.05, 0.2, 0.5)) {
#   hits <- 0; ratios <- numeric(3000)
#   for (i in 1:3000) {
#     if (disp == 0) { a <- rpois(6, 100); b <- rpois(6, 100) }
#     else {
#       a <- rpois(6, rgamma(6, 1/disp, scale = 100*disp))
#       b <- rpois(6, rgamma(6, 1/disp, scale = 100*disp))
#     }
#     ratios[i] <- var(c(a, b))/mean(c(a, b))
#     hits <- hits + (poisson_test(a, b) < 0.05)
#   }
#   cat(sprintf("  %-20.2f%10.1f%21.1f%%\n", disp, mean(ratios), 100*hits/3000))
# }
#
# ## With genuinely Poisson data the test is correct - about 5%. Add even
# ## modest overdispersion and the false positive rate climbs steeply, because
# ## the test divides by a standard error that is far too small.
# ##
# ## THE GROUPS ARE IDENTICAL IN EVERY ROW. Every "discovery" is false. This is
# ## what happens when you analyse RNA-seq with a model that assumes counting
# ## noise is the only noise, and it is why DESeq2 and edgeR spend most of
# ## their effort estimating the dispersion rather than testing.
```

### Problem 2: When does the normal approximation start working?

```{r problem2}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# skewness <- function(v) mean((v - mean(v))^3)/sd(v)^3
# cat(sprintf("  %-14s%12s%18s%14s\n", "n averaged", "skewness",
#             "P(within 2 SD)", "normal says"))
# for (n in c(1, 2, 5, 15, 50, 200)) {
#   m <- replicate(40000, mean(rlnorm(n, 0, 1.8)))
#   z <- (m - mean(m))/sd(m)
#   cat(sprintf("  %-14d%12.2f%18.3f%14.3f\n", n, skewness(m),
#               mean(abs(z) < 2), 0.9545))
# }
#
# ## Skewness measures asymmetry; 0 is symmetric. It falls steadily towards 0
# ## as n grows, and the coverage climbs towards the normal's 95.45%.
# ##
# ## Two lessons. First, the CLT really works - even this source becomes
# ## bell-shaped. Second, "n = 30 is enough" is FOLKLORE, not a theorem: for a
# ## source this skewed, n = 50 is still visibly off. The nastier the source,
# ## the larger n must be. When in doubt use a method that does not rely on the
# ## approximation at all - a permutation test or a bootstrap (Topics 7, 5).
```

### Problem 3: Reading a p-value histogram

```{r problem3}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# cases <- list()
# cases[["all null, correct test"]] <-
#   replicate(4000, t.test(rnorm(8), rnorm(8))$p.value)
# cases[["10% real effects"]] <-
#   sapply(1:4000, function(i) t.test(rnorm(8, if (i %% 10 == 0) 1.5 else 0),
#                                     rnorm(8))$p.value)
# cases[["null, but clustered data"]] <- replicate(4000, {
#   ga <- rnorm(2); gb <- rnorm(2)
#   t.test(rep(ga, each = 4) + rnorm(8, 0, 0.3),
#          rep(gb, each = 4) + rnorm(8, 0, 0.3))$p.value })
# cases[["null, conservative test"]] <-
#   pmin(1, 2.2*replicate(4000, t.test(rnorm(8), rnorm(8))$p.value))
# cat(sprintf("  %-28s%9s%34s\n", "analysis", "p<0.05", "shape"))
# for (nm in names(cases)) {
#   ps <- cases[[nm]]
#   lo <- mean(ps < 0.2); hi <- mean(ps > 0.8)
#   shape <- if (lo > 2.5*hi) "spike at 0 (signal, or a bug)"
#            else if (hi > 1.5*lo) "sloping up (conservative)" else "flat (healthy)"
#   cat(sprintf("  %-28s%9.3f%34s\n", nm, mean(ps < 0.05), shape))
# }
#
# ## (a) FLAT, 5% below 0.05. The test is calibrated - the baseline you compare
# ##     everything else against.
# ## (b) FLAT with a SPIKE at zero. Healthy, with real signal. The height of
# ##     the flat part estimates what fraction of features are null, which is
# ##     exactly how Storey's q-value works (Topic 8).
# ## (c) A spike at zero WITH NO REAL EFFECTS ANYWHERE. Indistinguishable from
# ##     (b) by eye, which is what makes it dangerous. If you know most
# ##     features should be null and you see this, suspect non-independence or
# ##     an unmodelled batch before you celebrate.
# ## (d) SLOPING UP towards 1. The test is conservative - you lose power but
# ##     make no false claims. Common with rank tests on tiny samples.
# ##
# ## Plot this histogram for every genome-scale analysis you run. It costs
# ## nothing and catches errors that staring at the top hits never will.
```

## What to take away

1. A distribution is a **mechanism**, not just a shape.
2. **Poisson** means counting noise only; **negative binomial** adds the
   biological variability real data always has.
3. **p-values are uniform under the null**: the cheapest diagnostic you own.
4. $t$, $\chi^2$ and $F$ describe **computed statistics**, not measurements.
5. Biology varies **multiplicatively**, so logs make it additive.

**Next:** `F4_estimates_and_standard_errors.R`
