---
title: "Applied 36 - Microbiome and compositional data"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 26, equations (26.1)-(26.6)

## Why this module is different

We simulate ABSOLUTE abundances and then close them, so we can answer the
question real data cannot: **which findings are biology and which are
arithmetic?**

```{r setup, message = FALSE}
MODULE_NAME <- "36_microbiome_compositional"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. Simulate ABSOLUTE abundances, then close them, eq. (26.1)

```{r simulate}
header("1. Absolute truth -> relative observation (26.1)")
simulate_microbiome <- function(n = 120, D = 180, n_changed = 45, log_effect = 1.9,
                                depth_lo = 5000, depth_hi = 60000) {
  group <- factor(rep(c("ctrl","case"), c(n %/% 2, n - n %/% 2)), levels = c("ctrl","case"))
  base <- exp(rnorm(D, 5.5, 2.2))
  true_log_fc <- numeric(D)
  changed <- sample(D, n_changed)
  ## Deliberately ASYMMETRIC: most changed taxa INCREASE, so the total shifts.
  true_log_fc[changed] <- abs(rnorm(n_changed, 0, log_effect))
  A <- t(sapply(seq_len(n), function(i) {
    eff <- if (group[i] == "case") true_log_fc else 0
    base * exp(eff + rnorm(D, 0, 0.55))
  }))
  A[matrix(runif(n*D), n) < 0.25] <- 0            # structural zeros
  depth <- sample(depth_lo:depth_hi, n, TRUE)
  Y <- t(sapply(seq_len(n), function(i) rmultinom(1, depth[i], A[i, ]/sum(A[i, ]))))
  dimnames(Y) <- list(sprintf("S%03d", seq_len(n)), sprintf("Taxon%03d", seq_len(D)))
  dimnames(A) <- dimnames(Y)
  list(counts = Y, absolute = A,
       meta = data.frame(group = group, depth = depth, row.names = rownames(Y)),
       truth = data.frame(log_fc = true_log_fc, changed = true_log_fc != 0,
                          row.names = colnames(Y)))
}
set.seed(3601)
sim <- simulate_microbiome()
counts <- sim$counts; absolute <- sim$absolute; meta <- sim$meta; truth <- sim$truth
D <- ncol(counts)
cat(sprintf("  %d samples x %d taxa\n", nrow(counts), D))
cat(sprintf("  sequencing depth: %s - %s (%.1fx range)\n",
            format(min(meta$depth), big.mark=","), format(max(meta$depth), big.mark=","),
            max(meta$depth)/min(meta$depth)))
cat(sprintf("  zeros: %.1f%% of the table\n", 100*mean(counts == 0)))
cat(sprintf("  taxa with a TRUE absolute increase: %d\n", sum(truth$changed)))
tc <- mean(rowSums(absolute)[meta$group == "ctrl"])
ts <- mean(rowSums(absolute)[meta$group == "case"])
cat(sprintf("\n  TRUE total absolute load: ctrl %s, case %s  (%.2fx)\n",
            format(round(tc), big.mark=","), format(round(ts), big.mark=","), ts/tc))
cat("  The total went UP. Sequencing cannot see that - it only ever returns a\n")
cat("  FIXED number of reads. Everything below follows from that fact.\n")
```

## 2. What closure does to the truth, eq. (2.7)

```{r closure}
header("2. Relative abundance of UNCHANGED taxa must fall (2.7)")
rel <- counts/rowSums(counts)
cat(sprintf("  %-26s%16s%16s%12s\n", "taxa", "mean rel. ctrl", "mean rel. case",
            "log2 ratio"))
for (row_ in list(list("TRULY increased", truth$changed),
                  list("TRULY unchanged", !truth$changed))) {
  rc <- mean(rel[meta$group == "ctrl", row_[[2]]])
  rs <- mean(rel[meta$group == "case", row_[[2]]])
  cat(sprintf("  %-26s%16.5f%16.5f%12.3f\n", row_[[1]], rc, rs, log2(rs/rc)))
}
cat("\n  The UNCHANGED taxa APPEAR to decrease in relative terms, because a\n")
cat("  fixed number of reads is shared with taxa that genuinely grew. This is\n")
cat("  arithmetic, not ecology (eq. 2.7).\n")
```

## 3. Log-ratio transformations, eq. (26.2)-(26.3)

```{r logratio}
header("3. clr / alr transforms and the zero problem (26.2)-(26.3)")
clr <- function(X, pseudocount = 0.5) {                          # eq. (26.2)
  Xp <- X + pseudocount; Xp <- Xp/rowSums(Xp)
  lg <- log(Xp); lg - rowMeans(lg)
}
Z <- clr(counts)
cat(sprintf("  clr rows sum to zero: max |row sum| = %.2e\n", max(abs(rowSums(Z)))))
cat(sprintf("  -> the clr covariance matrix is SINGULAR (rank %d of %d), which is\n",
            D - 1, D))
cat("     why you cannot invert it directly for a graphical model (10.6).\n")
tx <- rownames(truth)[truth$changed][1]
cat(sprintf("\n  Pseudocount sensitivity - the same taxon's clr effect:\n"))
cat(sprintf("  %14s%26s%12s\n", "pseudocount", "clr effect (case-ctrl)", "p-value"))
for (pc in c(0.1, 0.5, 1, 5)) {
  Zp <- clr(counts, pc)
  a <- Zp[meta$group == "ctrl", tx]; b <- Zp[meta$group == "case", tx]
  cat(sprintf("  %14.1f%26.4f%12.2e\n", pc, mean(b) - mean(a), t.test(a, b)$p.value))
}
cat("\n  Report the pseudocount and show conclusions survive changing it.\n")
cat("  Distinguish SAMPLING zeros (present but unseen at this depth - a\n")
cat("  pseudocount is reasonable) from STRUCTURAL zeros (genuinely absent -\n")
cat("  these should be MODELLED, not filled).\n")
```

## 4. Differential abundance: four strategies vs the known truth

```{r da}
header("4. Differential abundance against a KNOWN absolute truth")
da_relative_ttest <- function(counts, meta) {
  rel <- counts/rowSums(counts)
  apply(rel, 2, function(v) tryCatch(t.test(v ~ meta$group)$p.value, error = function(e) 1))
}
da_clr_ttest <- function(counts, meta, pc = 0.5) {
  Z <- clr(counts, pc)
  apply(Z, 2, function(v) tryCatch(t.test(v ~ meta$group)$p.value, error = function(e) 1))
}
da_ancombc_like <- function(counts, meta, pc = 0.5) {
  ## ANCOM-BC-style bias correction, eq. (26.4):
  ##     E[log Y_dj] = log A_dj + log d_j
  ## The sample-specific sampling fraction log d_j is a location shift COMMON
  ## to every taxon within a sample. Estimate it robustly (the median log
  ## abundance), subtract, then test taxon-wise linear models.
  logY <- log(counts + pc)
  corrected <- logY - apply(logY, 1, median)
  g <- as.numeric(meta$group == "case")
  X <- cbind(1, g)
  beta <- qr.coef(qr(X), corrected)
  resid <- corrected - X %*% beta
  dof <- nrow(X) - 2
  se <- sqrt(colSums(resid^2)/dof * chol2inv(qr.R(qr(X)))[2, 2])
  2*pt(abs(beta[2, ]/pmax(se, 1e-12)), dof, lower.tail = FALSE)
}
score <- function(p, label) {
  q <- p.adjust(p, "BH"); rej <- q < 0.05
  tp <- sum(rej & truth$changed); fp <- sum(rej & !truth$changed)
  cat(sprintf("  %-40s%7d%6d%6d%9.2f%8.3f\n", label, sum(rej), tp, fp,
              tp/sum(truth$changed), fp/max(sum(rej), 1)))
}
cat(sprintf("  Truth = an ABSOLUTE increase in %d taxa.\n", sum(truth$changed)))
cat(sprintf("  %-40s%7s%6s%6s%9s%8s\n", "method", "rej", "TP", "FP", "sens", "FDP"))
score(da_relative_ttest(counts, meta), "t-test on relative abundance")
score(da_clr_ttest(counts, meta), "clr + t-test (26.2)")
score(da_ancombc_like(counts, meta), "ANCOM-BC-style bias correction (26.4)")
cat("\n  The naive relative test produces a large number of false positives -\n")
cat("  the UNCHANGED taxa whose SHARE fell because the changed ones grew.\n")
cat("  Estimating and removing the sample-specific sampling fraction (26.4) is\n")
cat("  what ANCOM-BC2 does, and it is why it controls FDR where naive\n")
cat("  relative-abundance tests do not.\n")
```

## 5. Alpha diversity is depth-dependent, eq. (26.5)

```{r diversity}
header("5. Alpha diversity and rarefaction (26.5)")
shannon <- function(x) { p <- x[x > 0]/sum(x); -sum(p*log(p)) }
chao1 <- function(x) {
  s <- sum(x > 0); f1 <- sum(x == 1); f2 <- sum(x == 2)
  if (f2 > 0) s + f1^2/(2*f2) else s + f1*(f1-1)/2
}
div <- data.frame(observed = rowSums(counts > 0),
                  shannon = apply(counts, 1, shannon),
                  chao1 = apply(counts, 1, chao1),
                  depth = meta$depth, group = meta$group)
cat("  Correlation of each index with sequencing DEPTH (a pure artefact):\n")
for (c_ in c("observed","shannon","chao1"))
  cat(sprintf("    %-10s Spearman = %+.3f\n", c_,
              cor(div$depth, div[[c_]], method = "spearman")))
set.seed(5)
min_depth <- min(meta$depth)
rare <- t(sapply(seq_len(nrow(counts)), function(i)
  rmultinom(1, min_depth, counts[i, ]/sum(counts[i, ]))))
cat(sprintf("\n  After rarefying to %s reads:\n", format(min_depth, big.mark=",")))
cat(sprintf("    observed   Spearman with depth = %+.3f\n",
            cor(meta$depth, rowSums(rare > 0), method = "spearman")))
cat(sprintf("    shannon    Spearman with depth = %+.3f\n",
            cor(meta$depth, apply(rare, 1, shannon), method = "spearman")))
cat("\n  Diversity estimation is one of the FEW places rarefying is defensible -\n")
cat("  richness is intrinsically depth-dependent. For DIFFERENTIAL ABUNDANCE,\n")
cat("  rarefying throws away data and power; use an offset or a bias-corrected\n")
cat("  model instead.\n")
for (c_ in c("observed","shannon")) {
  v <- if (c_ == "observed") rowSums(rare > 0) else apply(rare, 1, shannon)
  cat(sprintf("    %-10s ctrl %8.3f  case %8.3f  p = %.4f\n", c_,
              mean(v[meta$group == "ctrl"]), mean(v[meta$group == "case"]),
              t.test(v ~ meta$group)$p.value))
}
```

## 6. Beta diversity: PERMANOVA and the dispersion trap, eq. (26.6)

```{r permanova}
header("6. PERMANOVA must be paired with a DISPERSION test (26.6)")
aitchison_distance <- function(counts, pc = 0.5) dist(clr(counts, pc))   # eq. (18.5)
bray_curtis_matrix <- function(counts) {
  R <- counts/rowSums(counts); n <- nrow(R)
  D <- matrix(0, n, n)
  for (i in seq_len(n)) D[i, ] <- rowSums(abs(sweep(R, 2, R[i, ])))/rowSums(sweep(R, 2, R[i, ], "+"))
  as.dist(D)
}
permanova <- function(D, group, n_perm = 299) {
  D2 <- as.matrix(D)^2; n <- nrow(D2); g <- as.character(group)
  pf <- function(lab) {
    total <- sum(D2)/(2*n); within <- 0
    for (lv in unique(lab)) { m <- lab == lv; within <- within + sum(D2[m, m])/(2*sum(m)) }
    a <- length(unique(lab))
    ((total - within)/(a - 1))/(within/(n - a))
  }
  f <- pf(g)
  null <- replicate(n_perm, pf(sample(g)))
  c(F = f, p = (1 + sum(null >= f))/(n_perm + 1))
}
permdisp <- function(D, group, n_perm = 299) {
  Dm <- as.matrix(D); g <- as.character(group)
  disp <- numeric(length(g))
  for (lv in unique(g)) {
    m <- which(g == lv)
    medoid <- m[which.min(rowSums(Dm[m, m, drop = FALSE]))]
    disp[m] <- Dm[m, medoid]
  }
  lv <- unique(g)
  obs <- abs(mean(disp[g == lv[1]]) - mean(disp[g == lv[2]]))
  null <- replicate(n_perm, {
    pg <- sample(g); abs(mean(disp[pg == lv[1]]) - mean(disp[pg == lv[2]]))
  })
  c(diff = obs, p = (1 + sum(null >= obs))/(n_perm + 1))
}
set.seed(3602)
for (row_ in list(list("Aitchison (18.5)", aitchison_distance(counts)),
                  list("Bray-Curtis (18.4)", bray_curtis_matrix(counts)))) {
  pv <- permanova(row_[[2]], meta$group); dv <- permdisp(row_[[2]], meta$group)
  cat(sprintf("  %-22s PERMANOVA F = %6.3f, p = %.4f   |   dispersion diff = %.4f, p = %.4f\n",
              row_[[1]], pv["F"], pv["p"], dv["diff"], dv["p"]))
}
cat("\n  TRAP DEMO - groups with IDENTICAL centroids but different spread.\n")
cat("  How often does PERMANOVA wrongly declare a location difference?\n\n")
cat(sprintf("  %-34s%10s%27s\n", "design", "sd ratio", "PERMANOVA rejection rate"))
for (cfg in list(c(60, 60, 1), c(100, 20, 2), c(20, 100, 3))) {
  na_ <- cfg[1]; nb_ <- cfg[2]
  lab <- c("balanced (60 vs 60)", "unbalanced, SMALL grp dispersed",
           "unbalanced, LARGE grp dispersed")[cfg[3]]
  for (sdr in c(2, 3)) {
    rate <- mean(replicate(25, {
      Zt <- rbind(matrix(rnorm(na_*20, 0, 1), na_), matrix(rnorm(nb_*20, 0, sdr), nb_))
      permanova(dist(Zt), rep(c("a","b"), c(na_, nb_)), n_perm = 149)["p"] < 0.05
    }))
    cat(sprintf("  %-34s%10.1f%26.0f%%\n", lab, sdr, 100*rate))
  }
}
cat("\n  The result is sharper than the usual warning. PERMANOVA is fairly\n")
cat("  ROBUST to dispersion differences when the design is BALANCED. It breaks\n")
cat("  down badly when the groups are UNBALANCED and the SMALLER group is the\n")
cat("  more dispersed one - then it rejects almost always, with identical\n")
cat("  centroids. (Anderson & Walsh 2013 document exactly this asymmetry.)\n")
cat("\n  Two practical consequences:\n")
cat("   * Balance your group sizes - yet another payoff from design.\n")
cat("   * Always report PERMDISP alongside PERMANOVA, and say which of\n")
cat("     'different centroids' and 'different spread' your data support.\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "microbiome.png"), width = 1100, height = 800, res = 110)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.5, 1))
ma_c <- colMeans(absolute[meta$group == "ctrl", ]); ma_s <- colMeans(absolute[meta$group == "case", ])
mr_c <- colMeans(rel[meta$group == "ctrl", ]); mr_s <- colMeans(rel[meta$group == "case", ])
ok <- ma_c > 0 & mr_c > 0
plot(log2(ma_s[ok]/ma_c[ok]), log2(mr_s[ok]/mr_c[ok]), pch = 16, cex = 0.5,
     col = ifelse(truth$changed[ok], "firebrick", adjustcolor("grey50", 0.5)),
     xlab = "TRUE log2 FC (absolute)", ylab = "observed log2 FC (relative)",
     main = "Eq. (2.7): closure shifts everything down")
abline(h = 0); abline(v = 0); abline(0, 1, lty = 2)
plot(div$depth, div$observed, pch = 16, cex = 0.6, col = "steelblue",
     xlab = "sequencing depth", ylab = "observed taxa",
     main = "Eq. (26.5): richness tracks depth")
points(div$depth, rowSums(rare > 0), pch = 16, cex = 0.6, col = "darkorange")
legend("bottomright", c("raw","rarefied"), col = c("steelblue","darkorange"),
       pch = 16, bty = "n", cex = 0.7)
pco <- cmdscale(aitchison_distance(counts), k = 2)
plot(pco, pch = 16, cex = 0.7, col = c("steelblue","darkorange")[meta$group],
     xlab = "PCoA 1", ylab = "PCoA 2", main = "Aitchison PCoA (eq. 18.5)")
legend("topright", levels(meta$group), col = c("steelblue","darkorange"),
       pch = 16, bty = "n", cex = 0.7)
set.seed(6)
Zc <- rbind(matrix(rnorm(100*20, 0, 1), 100), matrix(rnorm(20*20, 0, 3), 20))
plot(Zc[, 1], Zc[, 2], pch = 16, cex = 0.6,
     col = c(rep("steelblue", 100), rep("darkorange", 20)),
     main = "Trap: same centroid, unbalanced dispersion", xlab = "", ylab = "")
legend("topright", c("large n, tight","small n, dispersed"),
       col = c("steelblue","darkorange"), pch = 16, bty = "n", cex = 0.65)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "microbiome.png"), "\n")
```

# PROBLEMS

### Problem 1: Spurious correlation from closure

Correlate taxa whose ABSOLUTE abundances are independent, on the relative and
clr scales. Which is honest?

```{r problem1}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# sub <- colnames(counts)[1:40]
# Cabs <- cor(absolute[, sub]); Crel <- cor(rel[, sub]); Cclr <- cor(clr(counts)[, sub])
# off <- upper.tri(Cabs)
# for (row_ in list(list("ABSOLUTE (the truth)", Cabs), list("relative (closed)", Crel),
#                   list("clr", Cclr)))
#   cat(sprintf("  %-24s mean off-diag r = %+.4f, max |r| = %.3f\n", row_[[1]],
#               mean(row_[[2]][off]), max(abs(row_[[2]][off]))))
#
# ## Closure forces the AVERAGE correlation negative (eq. 2.7). The clr partially
# ## removes it but is NOT a complete fix: clr rows also sum to zero, so clr
# ## values carry a residual negative constraint. For network inference use
# ## proportionality (phi, rho_p) or SPIEC-EASI, which estimate a sparse
# ## PRECISION matrix (eq. 10.6) on log-ratios rather than raw correlations.
```

### Problem 2: Absolute quantification rescues the analysis

Suppose you also measured total bacterial load by qPCR. Multiply relative
abundances by the (known) total and redo the test. How much does it improve?

```{r problem2}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# total_load <- rowSums(absolute)                    # the "qPCR" measurement
# abs_est <- rel * total_load
# logA <- log(abs_est + 1)
# g <- as.numeric(meta$group == "case"); X <- cbind(1, g)
# beta <- qr.coef(qr(X), logA); resid <- logA - X %*% beta
# se <- sqrt(colSums(resid^2)/(nrow(X)-2) * chol2inv(qr.R(qr(X)))[2,2])
# p_abs <- 2*pt(abs(beta[2, ]/pmax(se, 1e-12)), nrow(X)-2, lower.tail = FALSE)
# cat(sprintf("  %-40s%7s%6s%6s%9s%8s\n", "method","rej","TP","FP","sens","FDP"))
# score(da_ancombc_like(counts, meta), "bias-corrected relative (26.4)")
# score(p_abs, "ABSOLUTE abundance (rel x qPCR total)")
#
# ## With an absolute anchor the question becomes answerable DIRECTLY and both
# ## sensitivity and FDP improve. This is why spike-ins and qPCR total-load
# ## measurements are worth the effort: they convert an unidentifiable question
# ## into an identifiable one. Without them, state that your claim is about
# ## RELATIVE abundance.
```

### Problem 3: Prevalence filtering

Filter taxa present in at least 10%, 25% and 50% of samples. Why must the
filter be outcome-independent?

```{r problem3}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# cat(sprintf("  %19s%12s%7s%6s%6s%9s%8s\n", "prevalence filter","taxa kept",
#             "rej","TP","FP","sens","FDP"))
# for (thr in c(0, 0.10, 0.25, 0.50)) {
#   keep <- colMeans(counts > 0) >= thr
#   p <- da_ancombc_like(counts[, keep, drop = FALSE], meta)
#   q <- p.adjust(p, "BH"); rej <- q < 0.05
#   t0 <- truth[names(p), ]
#   cat(sprintf("  %18.0f%%%12d%7d%6d%6d%9.2f%8.3f\n", 100*thr, sum(keep), sum(rej),
#               sum(rej & t0$changed), sum(rej & !t0$changed),
#               sum(rej & t0$changed)/sum(truth$changed),
#               sum(rej & !t0$changed)/max(sum(rej), 1)))
# }
#
# ## Prevalence is computed from the COUNTS alone, with no reference to the
# ## group labels, so it is OUTCOME-INDEPENDENT and cannot break FDR control
# ## (Module 08, section 6). Filtering on "taxa that differ between groups"
# ## would. Note the trade-off: aggressive filtering removes rare taxa that may
# ## be exactly the interesting ones. Pre-specify the rule.
```

## What to take away

1. State whether your claim is about RELATIVE or ABSOLUTE abundance.
2. Transform with log-ratios before distance, ordination, correlation or PCA.
3. Handle zeros explicitly; report the pseudocount and a sensitivity analysis.
4. Use bias-corrected differential abundance (26.4).
5. **Pair every PERMANOVA with a dispersion test.**
6. Rarefy only for diversity, never for differential abundance.

**Next:** `37_spatial_omics.R`
