---
title: "Applied 37 - Spatial transcriptomics and spatial omics"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 27, equations (27.1)-(27.5)

## The central trap

A tissue section contains thousands of spots but is **one observation of one
patient**. Spot-level p-values describe that section, not the condition.

```{r setup, message = FALSE}
suppressPackageStartupMessages(library(nlme))
MODULE_NAME <- "37_spatial_omics"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. Simulate replicated tissue sections

```{r simulate}
header("1. A multi-patient spatial experiment")
simulate_spatial <- function(n_patients = 8, side = 26, G = 250, n_svg = 30) {
  coords0 <- expand.grid(x = 0:(side-1), y = 0:(side-1))
  coords0$x <- coords0$x + (coords0$y %% 2)*0.5          # hexagonal-ish offset
  n_spot <- nrow(coords0)
  condition <- rep(c("ctrl","case"), each = n_patients/2)
  is_svg <- c(rep(TRUE, n_svg), rep(FALSE, G - n_svg))
  base <- exp(rnorm(G, -0.7, 1.1))
  blocks <- list(); metas <- list()
  for (pid in seq_len(n_patients)) {
    patient_eff <- rnorm(G, 0, 0.45)                     # patient RANDOM effect
    cx <- side/2 + rnorm(1, 0, 1.5); cy <- side/2 + rnorm(1, 0, 1.5)
    radius <- side*0.22 * if (condition[pid] == "case") 1.45 else 1
    dist_ <- sqrt((coords0$x - cx)^2 + (coords0$y - cy)^2)
    in_domain <- as.numeric(dist_ < radius)
    field <- exp(-(dist_/(side*0.25))^2)
    depth <- rlnorm(n_spot, 0, 0.32)
    mu <- outer(base*exp(patient_eff), depth)
    mu[is_svg, ] <- mu[is_svg, ] * rep(exp(1.4*field), each = sum(is_svg))
    Y <- matrix(rpois(length(mu), mu), nrow = G)
    ids <- sprintf("P%02d_S%04d", pid-1, seq_len(n_spot))
    dimnames(Y) <- list(sprintf("G%03d", seq_len(G)), ids)
    blocks[[pid]] <- Y
    metas[[pid]] <- data.frame(patient = sprintf("P%02d", pid-1),
                               condition = condition[pid],
                               x = coords0$x, y = coords0$y, in_domain = in_domain,
                               total_umi = colSums(Y), row.names = ids)
  }
  list(counts = do.call(cbind, blocks), spot_meta = do.call(rbind, metas),
       truth = data.frame(is_svg = is_svg, row.names = sprintf("G%03d", seq_len(G))))
}
set.seed(3701)
sim <- simulate_spatial()
counts <- sim$counts; spot_meta <- sim$spot_meta; truth <- sim$truth
spot_meta$condition <- factor(spot_meta$condition, levels = c("ctrl","case"))
stopifnot(identical(colnames(counts), rownames(spot_meta)))
cat(sprintf("  %d genes x %s spots from %d patients\n", nrow(counts),
            format(ncol(counts), big.mark=","), length(unique(spot_meta$patient))))
cat(sprintf("  spots per section: %d\n", sum(spot_meta$patient == "P00")))
cat(sprintf("  truly spatially variable genes: %d\n", sum(truth$is_svg)))
cat(sprintf("\n  n for CONDITION-level claims = %d patients, NOT %s spots\n",
            length(unique(spot_meta$patient)), format(ncol(counts), big.mark=",")))
```

## 2. Spatial autocorrelation: Moran's I, eq. (27.1)

Note `E_0[I] = -1/(n-1)`, **not zero** - a routinely misreported detail.

```{r morans}
header("2. Moran's I and Geary's C (27.1)-(27.2)")
knn_weights <- function(coords, k = 6) {
  d <- as.matrix(dist(coords)); diag(d) <- Inf
  W <- matrix(0, nrow(d), nrow(d))
  for (i in seq_len(nrow(d))) W[i, order(d[i, ])[1:k]] <- 1/k
  W
}
morans_i <- function(x, W) {                                     # eq. (27.1)
  n <- length(x); z <- x - mean(x)
  den <- sum(z^2)
  if (den == 0) return(0)
  (n/sum(W)) * as.numeric(t(z) %*% W %*% z)/den
}
gearys_c <- function(x, W) {                                     # eq. (27.2)
  n <- length(x); z <- x - mean(x); den <- sum(z^2)
  if (den == 0) return(1)
  num <- sum(W * outer(x, x, "-")^2)
  ((n-1)*num)/(2*sum(W)*den)
}
p0 <- spot_meta$patient == "P00"
coords0 <- as.matrix(spot_meta[p0, c("x","y")])
W <- knn_weights(coords0, 6)
n_spots0 <- nrow(coords0)
cat(sprintf("  section P00: %d spots, k=6 neighbour graph\n", n_spots0))
cat(sprintf("  E_0[Moran's I] = -1/(n-1) = %.5f  (NOT zero)\n", -1/(n_spots0-1)))
cat("  E_0[Geary's C] = 1.0\n")
sub <- counts[, p0]
logn <- log1p(t(t(sub)/colSums(sub))*1e4)
cat(sprintf("\n  %-22s%15s%15s\n", "gene class", "mean Moran I", "mean Geary C"))
for (row_ in list(list("truly spatial (SVG)", which(truth$is_svg)[1:30]),
                  list("non-spatial", which(!truth$is_svg)[1:30])))
  cat(sprintf("  %-22s%15.4f%15.4f\n", row_[[1]],
              mean(sapply(row_[[2]], function(g) morans_i(logn[g, ], W))),
              mean(sapply(row_[[2]], function(g) gearys_c(logn[g, ], W)))))
```

## 3. Testing spatial variability by permutation

```{r svg-test}
header("3. Permutation test for spatially variable genes")
moran_perm_test <- function(x, W, n_perm = 99) {
  obs <- morans_i(x, W)
  null <- replicate(n_perm, morans_i(sample(x), W))
  c(I = obs, p = (1 + sum(null >= obs))/(n_perm + 1))
}
set.seed(3702)
genes_test <- c(which(truth$is_svg)[1:20], which(!truth$is_svg)[1:60])
svg_res <- t(sapply(genes_test, function(g) moran_perm_test(logn[g, ], W)))
rownames(svg_res) <- rownames(logn)[genes_test]
q <- p.adjust(svg_res[, "p"], "BH")
is_svg_t <- truth$is_svg[genes_test]
cat(sprintf("  tested %d genes in ONE section\n", length(genes_test)))
cat(sprintf("  detected: %d  (TP %d, FP %d)\n", sum(q < 0.05),
            sum(q < 0.05 & is_svg_t), sum(q < 0.05 & !is_svg_t)))
cat(sprintf("  sensitivity %.2f\n", sum(q < 0.05 & is_svg_t)/sum(is_svg_t)))
cat("\n  This is a valid WITHIN-SECTION claim: 'this gene is spatially organised\n")
cat("  in this tissue'. It says NOTHING about the condition.\n")
cat("  NOTE we permute VALUES over LOCATIONS, which preserves both the\n")
cat("  expression distribution and the tissue geometry. The normal\n")
cat("  approximation for Moran's I is poor on irregular lattices.\n")
```

## 4. THE TRAP: spot-level condition testing

```{r trap}
header("4. Spots are not patients (eq. 1.5 again)")
dom_frac <- aggregate(in_domain ~ patient + condition, spot_meta, mean)
d_spot <- spot_meta
d_spot$case <- as.numeric(d_spot$condition == "case")
m_spot <- glm(in_domain ~ case, data = d_spot, family = binomial())
a <- dom_frac$in_domain[dom_frac$condition == "ctrl"]
b <- dom_frac$in_domain[dom_frac$condition == "case"]
p_patient <- t.test(a, b)$p.value
m_mixed <- lme(in_domain ~ case, random = ~ 1 | patient, data = d_spot)
cat(sprintf("  domain fraction: ctrl %.3f, case %.3f\n", mean(a), mean(b)))
cat(sprintf("\n  %-44s%14s%14s\n", "analysis", "p-value", "effective n"))
cat(sprintf("  %-44s%14.2e%14s\n", "spot-level logistic (WRONG)",
            coef(summary(m_spot))["case", 4], format(nrow(d_spot), big.mark=",")))
cat(sprintf("  %-44s%14.2e%14d\n", "per-patient summary, t-test", p_patient, nrow(dom_frac)))
cat(sprintf("  %-44s%14.2e%14d\n", "mixed model, patient random effect",
            summary(m_mixed)$tTable["case", "p-value"], nrow(dom_frac)))
cat("\n  The spot-level p-value is astronomically small because it counts\n")
cat("  thousands of spots as independent. The design effect (eq. 1.8) applies\n")
cat("  exactly as it did for cells in Module 31.\n")
cat("\n  NEGATIVE CONTROL - assign patients to arms at random:\n")
set.seed(11)
fake <- t(replicate(40, {
  lv <- sample(rep(c("ctrl","case"), each = nrow(dom_frac)/2))
  names(lv) <- dom_frac$patient
  d2 <- spot_meta; d2$fake <- as.numeric(lv[as.character(d2$patient)] == "case")
  ms <- glm(in_domain ~ fake, data = d2, family = binomial())
  dd <- dom_frac; dd$fake <- lv[as.character(dd$patient)]
  c(coef(summary(ms))["fake", 4] < 0.05,
    t.test(dd$in_domain[dd$fake == "ctrl"], dd$in_domain[dd$fake == "case"])$p.value < 0.05)
}))
cat(sprintf("    spot-level false-positive rate    : %.0f%%   <- should be 5%%\n",
            100*mean(fake[, 1])))
cat(sprintf("    patient-level false-positive rate : %.0f%%\n", 100*mean(fake[, 2])))
```

## 5. Counts stay counts: the total-UMI offset, eq. (27.4)

```{r offset}
header("5. Spot depth varies enormously - use an offset (27.4)")
cat(sprintf("  total UMI per spot: %s - %s (%.1fx)\n",
            format(min(spot_meta$total_umi), big.mark=","),
            format(max(spot_meta$total_umi), big.mark=","),
            max(spot_meta$total_umi)/max(min(spot_meta$total_umi), 1)))
g_test <- 1
d <- spot_meta[p0, ]; d$y <- as.numeric(sub[g_test, ])
m_no <- glm(y ~ in_domain, data = d, family = poisson())
m_off <- glm(y ~ in_domain, data = d, family = poisson(), offset = log(total_umi))
cat(sprintf("\n  gene %s, domain effect:\n", rownames(counts)[g_test]))
cat(sprintf("    without offset: %+.4f (p = %.2e)\n", coef(m_no)["in_domain"],
            coef(summary(m_no))["in_domain", 4]))
cat(sprintf("    with offset    : %+.4f (p = %.2e)\n", coef(m_off)["in_domain"],
            coef(summary(m_off))["in_domain", 4]))
cat("  Without the offset the model partly measures how deeply each spot was\n")
cat("  sequenced, not how much RNA the cells contained (eq. 13.8).\n")
```

## 6. Edge effects and neighbourhood-graph sensitivity

```{r edges}
header("6. Edge effects and the neighbour graph")
d_edge <- pmin(coords0[,1] - min(coords0[,1]), max(coords0[,1]) - coords0[,1],
               coords0[,2] - min(coords0[,2]), max(coords0[,2]) - coords0[,2])
is_edge <- d_edge < 2
Dm <- as.matrix(dist(coords0))
n_within2 <- rowSums(Dm > 0 & Dm <= 2)
cat(sprintf("  spots within 2 units: interior mean %.1f, edge mean %.1f\n",
            mean(n_within2[!is_edge]), mean(n_within2[is_edge])))
svg0 <- which(truth$is_svg)[1]
cat(sprintf("\n  Moran's I for a spatial gene:\n"))
cat(sprintf("    all spots             : %.4f\n", morans_i(logn[svg0, ], W)))
cat(sprintf("    interior only (buffer): %.4f\n",
            morans_i(logn[svg0, !is_edge], knn_weights(coords0[!is_edge, ], 6))))
cat("  Boundary spots have fewer TRUE neighbours, which biases neighbourhood\n")
cat("  statistics. Use a buffer zone, an edge-corrected estimator, or a\n")
cat("  toroidal correction - and SAY WHICH.\n")
cat(sprintf("\n  %16s%24s%22s\n", "k (neighbours)", "mean I, spatial genes",
            "mean I, non-spatial"))
for (k in c(4, 6, 12, 30)) {
  Wk <- knn_weights(coords0, k)
  cat(sprintf("  %16d%24.4f%22.4f\n", k,
              mean(sapply(which(truth$is_svg)[1:15], function(g) morans_i(logn[g, ], Wk))),
              mean(sapply(which(!truth$is_svg)[1:15], function(g) morans_i(logn[g, ], Wk)))))
}
cat("\n  Larger neighbourhoods detect broader gradients; smaller ones detect\n")
cat("  fine structure. There is NO single correct k - report the choice and\n")
cat("  show sensitivity to it.\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "spatial.png"), width = 1100, height = 800, res = 110)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.5, 1))
pal <- hcl.colors(30, "viridis")
plot(coords0, pch = 15, cex = 0.5, col = pal[cut(logn[svg0, ], 30)],
     xlab = "x", ylab = "y",
     main = sprintf("SVG (Moran I = %.2f)", morans_i(logn[svg0, ], W)))
ns <- which(!truth$is_svg)[1]
plot(coords0, pch = 15, cex = 0.5, col = pal[cut(logn[ns, ], 30)],
     xlab = "x", ylab = "y",
     main = sprintf("Non-spatial gene (Moran I = %.2f)", morans_i(logn[ns, ], W)))
hist(svg_res[!is_svg_t, "I"], breaks = 25, col = adjustcolor("steelblue", 0.6),
     border = "white", xlim = range(svg_res[, "I"]),
     main = "Eq. (27.1): the null is NOT zero", xlab = "Moran's I")
hist(svg_res[is_svg_t, "I"], breaks = 25, col = adjustcolor("firebrick", 0.6),
     border = "white", add = TRUE)
abline(v = -1/(n_spots0-1), col = "red", lty = 2, lwd = 2)
stripchart(in_domain ~ condition, data = dom_frac, vertical = TRUE, pch = 16,
           cex = 1.4, col = c("steelblue","darkorange"),
           ylab = "domain fraction per SECTION",
           main = sprintf("Patient-level unit: n = %d, p = %.3f", nrow(dom_frac), p_patient))
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "spatial.png"), "\n")
```

# PROBLEMS

### Problem 1: Colocalisation with a geometry-preserving null

Label spots as cell type A or B and ask whether A and B are closer than
chance. Permute LABELS, keeping COORDINATES fixed. Why not permute
coordinates?

```{r problem1}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# set.seed(21)
# lab <- ifelse(spot_meta$in_domain[p0] > 0, "A", "B")
# flip <- runif(length(lab)) < 0.12
# lab <- ifelse(flip, ifelse(lab == "A", "B", "A"), lab)
# cross_nn <- function(coords, lab) {
#   A <- coords[lab == "A",, drop = FALSE]; B <- coords[lab == "B",, drop = FALSE]
#   if (!nrow(A) || !nrow(B)) return(NA)
#   mean(apply(as.matrix(dist(rbind(A, B)))[1:nrow(A), -(1:nrow(A)), drop = FALSE], 1, min))
# }
# obs <- cross_nn(coords0, lab)
# null <- replicate(299, cross_nn(coords0, sample(lab)))
# cat(sprintf("  observed mean A->nearest-B distance: %.3f\n", obs))
# cat(sprintf("  permutation null mean              : %.3f\n", mean(null)))
# cat(sprintf("  p (A closer to B than chance)      : %.4f\n",
#             (1 + sum(null <= obs))/300))
#
# ## Permuting LABELS keeps the tissue's shape, density and any holes exactly as
# ## observed, so the null asks precisely 'given this tissue geometry, is the
# ## arrangement of labels special?'. Permuting COORDINATES would destroy the
# ## geometry and test a different, uninteresting null (uniform tissue).
# ## Colocalisation is also SCALE-DEPENDENT: repeat at several radii (a cross-K
# ## function) rather than reporting a single number.
```

### Problem 2: Deconvolution uncertainty

Simulate spot compositions as mixtures of two cell types, estimate by
constrained least squares, and propagate the uncertainty into a downstream
comparison. How much does ignoring it inflate significance?

```{r problem2}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# set.seed(31)
# n_sp <- 300; G2 <- 60
# prof <- matrix(rlnorm(G2*2), G2)
# w_true <- rbeta(n_sp, 2, 2)
# Yd <- prof %*% rbind(w_true, 1 - w_true) * matrix(rlnorm(G2*n_sp, 0, 0.35), G2)
# w_hat <- apply(Yd, 2, function(y) {
#   w <- c(0.5, 0.5)
#   for (it in 1:400) {
#     g <- crossprod(prof, prof %*% w - y)
#     w <- pmax(w - 0.3*g/max(sum(prof^2), 1e-9), 0); if (sum(w) > 0) w <- w/sum(w)
#   }
#   w[1]
# })
# rmse <- sqrt(mean((w_hat - w_true)^2))
# cat(sprintf("  correlation(estimated, true) = %.3f\n", cor(w_hat, w_true)))
# cat(sprintf("  RMSE of the proportion estimate = %.3f\n", rmse))
# grp <- rep(0:1, each = n_sp/2)                   # NO real difference
# p_naive <- t.test(w_hat[grp == 0], w_hat[grp == 1])$p.value
# boot_p <- replicate(200, {
#   wb <- pmin(pmax(w_hat + rnorm(n_sp, 0, rmse), 0), 1)
#   t.test(wb[grp == 0], wb[grp == 1])$p.value
# })
# cat(sprintf("  p treating w_hat as OBSERVED    : %.4f\n", p_naive))
# cat(sprintf("  median p with uncertainty added : %.4f\n", median(boot_p)))
#
# ## Deconvolved proportions are ESTIMATES with real error, but almost every
# ## downstream analysis treats them as measured quantities. Propagate the
# ## uncertainty (bootstrap or posterior draws) or state plainly that you did
# ## not - the same understatement-of-uncertainty issue as a batch-corrected
# ## matrix in Module 19.
```

## What to take away

1. **Never treat spots or cells as patient replicates.**
2. Build the neighbourhood graph deliberately and report sensitivity to it.
3. Use permutation nulls that preserve tissue geometry.
4. Keep counts as counts with a total-UMI offset (27.4).
5. `E_0[Moran's I] = -1/(n-1)`, not 0.
6. Propagate deconvolution uncertainty, or say that you did not.

**Next:** `38_survival_biomarkers.R`
