---
title: "Applied 39 - Gene set enrichment done honestly"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 29, equations (29.1)-(29.5)

## The question

A differential expression list is 400 genes long. Which **pathways** are
involved? Almost every trap in this module comes from one fact: **genes in a
pathway are correlated**, and the standard tests assume they are not.

```{r setup, message = FALSE}
MODULE_NAME <- "39_gene_set_enrichment"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. Over-representation analysis, eq. (29.1)

ORA asks: is the overlap between my DE list and a pathway bigger than chance?
Under the null the overlap is **hypergeometric** - sampling `n` genes without
replacement from `N`, of which `K` are in the set:

$$P(X = k) = \frac{\binom{K}{k}\binom{N-K}{n-k}}{\binom{N}{n}} \tag{29.1}$$

```{r ora}
header("1. ORA and the hypergeometric null (29.1)")
N <- 20000; K <- 200; n <- 400; k <- 20
## phyper(k-1, ..., lower.tail = FALSE) = P(X >= k). Using P(X > k) is an
## off-by-one that silently makes every p-value too small.
p_over <- phyper(k - 1, K, N - K, n, lower.tail = FALSE)
cat(sprintf("  N = %d genes tested, K = %d in the pathway\n", N, K))
cat(sprintf("  n = %d DE genes, k = %d of them in the pathway\n", n, k))
cat(sprintf("  expected overlap under the null = n*K/N = %.1f\n", n*K/N))
cat(sprintf("  fold enrichment = %.2f\n", k/(n*K/N)))
cat(sprintf("  P(X >= %d) = %.3e   (one-sided hypergeometric, eq. 29.1)\n", k, p_over))
cat(sprintf("  Fisher exact 2x2 agrees: p = %.3e\n",
            fisher.test(matrix(c(k, n-k, K-k, N-K-n+k), 2), alternative = "greater")$p.value))
cat("\n  The BACKGROUND (N) is the most abused choice in all of ORA.\n")
for (Nb in c(20000, 12000, 6000)) {
  pb <- phyper(k - 1, K, Nb - K, n, lower.tail = FALSE)
  cat(sprintf("    background = %5d genes -> p = %.2e, fold = %.2f\n",
              Nb, pb, k/(n*K/Nb)))
}
cat("  Same data, three different answers. The background must be the genes\n")
cat("  that COULD have been detected - genes that passed your expression\n")
cat("  filter - not all of Ensembl. Using the whole genome as background\n")
cat("  manufactures enrichment for anything expressed in your tissue.\n")
```

## 2. ORA throws away the ranking

```{r ora-threshold}
header("2. ORA depends on a threshold you chose arbitrarily")
set.seed(3901)
G <- 5000
truth <- rep(0, G); pathway <- 1:100
## A coordinated but MODEST shift across a whole pathway - the biologically
## realistic case, and the one ORA is worst at.
truth[pathway] <- 0.45
stat <- rnorm(G, truth)
pv <- 2*pnorm(-abs(stat))
cat(sprintf("  Pathway of %d genes, each shifted by %.2f SD (a coordinated,\n",
            length(pathway), 0.45))
cat("  modest effect). Vary ONLY the DE-list cutoff:\n\n")
cat(sprintf("  %-14s%10s%10s%14s\n", "p cutoff", "DE genes", "overlap", "ORA p"))
for (cut in c(0.001, 0.01, 0.05, 0.1, 0.2)) {
  de <- which(pv < cut); ov <- length(intersect(de, pathway))
  pp <- phyper(ov - 1, length(pathway), G - length(pathway), length(de),
               lower.tail = FALSE)
  cat(sprintf("  %-14s%10d%10d%14.2e\n", sprintf("p < %g", cut), length(de), ov, pp))
}
cat("\n  The conclusion swings by orders of magnitude on a cutoff nobody can\n")
cat("  justify, because ORA sees only in/out and discards the ranking. A\n")
cat("  gene at p = 0.051 carries almost the same evidence as one at 0.049.\n")
```

## 3. GSEA: use the whole ranked list, eq. (29.4)

Walk down the ranked list, adding when a gene is in the set and subtracting
when it is not. The enrichment score is the maximum deviation from zero:

$$ES = \max_{i}\left|\sum_{j \le i} w_j\right| \tag{29.4}$$

```{r gsea}
header("3. GSEA running enrichment score (29.4)")
running_es <- function(stat, in_set, p = 1) {
  o <- order(stat, decreasing = TRUE)
  s <- stat[o]; m <- in_set[o]
  ## Hits are weighted by |stat|^p (p = 1 is the GSEA default); misses are
  ## weighted uniformly so that the walk returns to zero at the end.
  w <- abs(s)^p
  hit <- ifelse(m, w/sum(w[m]), 0)
  miss <- ifelse(m, 0, 1/sum(!m))
  cumsum(hit - miss)
}
in_set <- seq_len(G) %in% pathway
res <- running_es(stat, in_set)
es <- res[which.max(abs(res))]
cat(sprintf("  observed ES = %+.4f (at rank %d of %d)\n",
            es, which.max(abs(res)), G))
## Permuting GENE LABELS keeps the ranked statistics fixed and reshuffles set
## membership. It is a COMPETITIVE null: 'is this set more extreme than a
## random set of the same size?'
set.seed(3902)
NPERM <- 2000
null_es <- replicate(NPERM, {
  r <- running_es(stat, sample(in_set))
  r[which.max(abs(r))]
})
p_gsea <- (1 + sum(abs(null_es) >= abs(es)))/(NPERM + 1)
nes <- es/mean(abs(null_es[sign(null_es) == sign(es)]))
cat(sprintf("  gene-label permutation null: mean %+.4f, SD %.4f\n",
            mean(null_es), sd(null_es)))
cat(sprintf("  NES = %+.3f, permutation p = %.2e   (%d permutations)\n",
            nes, p_gsea, NPERM))
cat(sprintf("  ORA at p < 0.05 for comparison: p = %.2e\n",
            {de <- which(pv < 0.05)
             phyper(length(intersect(de, pathway)) - 1, length(pathway),
                    G - length(pathway), length(de), lower.tail = FALSE)}))
cat("\n  GSEA uses every gene's rank and needs no cutoff. At the conventional\n")
cat("  p < 0.05 cutoff ORA misses this pathway completely (p = 0.38); at\n")
cat("  p < 0.01 it scrapes in at 0.045. Its verdict is decided by the cutoff,\n")
cat("  GSEA's is not - which is the whole argument for rank-based tests when\n")
cat("  the effect is coordinated and modest rather than large in a few genes.\n")
```

## 4. Competitive vs self-contained, and why it matters

* **Self-contained**: 'is ANY gene in this set differentially expressed?'
  Null = no gene in the set changes. Permutes **sample labels**.
* **Competitive**: 'is this set MORE differential than genes outside it?'
  Null = set genes behave like the rest. Permutes **gene labels**.

```{r competitive}
header("4. Competitive vs self-contained nulls (29.2)-(29.3)")
## Vectorised Welch t-statistic - used from here on, so the simulations below
## can afford enough replicates for the rates to be meaningful.
tstat <- function(mat, g) {
  a <- mat[, g == 1, drop = FALSE]; b <- mat[, g == 0, drop = FALSE]
  (rowMeans(a) - rowMeans(b)) /
    sqrt(apply(a, 1, var)/ncol(a) + apply(b, 1, var)/ncol(b))
}
set.seed(3903)
## A GLOBAL shift: EVERY gene moves a little (e.g. a change in library
## composition or a global stress response). No set is special.
R4 <- 200; m4 <- 100; G4 <- 4000
hit_self <- hit_comp <- hit_ora <- 0
for (i in 1:R4) {
  sg <- rnorm(G4, 0.30)                     # every gene shifted, none special
  ism <- seq_len(G4) %in% 1:m4
  hit_self <- hit_self + (t.test(sg[ism])$p.value < 0.05)
  rr <- running_es(sg, ism); ee <- rr[which.max(abs(rr))]
  nl <- replicate(100, {q <- running_es(sg, sample(ism)); q[which.max(abs(q))]})
  hit_comp <- hit_comp + ((1 + sum(abs(nl) >= abs(ee)))/101 < 0.05)
  dd <- which(2*pnorm(-abs(sg)) < 0.05)
  hit_ora <- hit_ora + (phyper(length(intersect(dd, which(ism))) - 1, m4,
                               G4 - m4, max(length(dd), 1),
                               lower.tail = FALSE) < 0.05)
}
cat(sprintf("  %d datasets in which EVERY gene is shifted by 0.30 SD.\n", R4))
cat("  The set is in no way special, so a COMPETITIVE test should fire 5% of\n")
cat("  the time and a SELF-CONTAINED test should fire almost always.\n\n")
cat(sprintf("  %-46s%14s\n", "test", "fires"))
cat(sprintf("  %-46s%13.0f%%   <- as designed\n", "self-contained (set mean vs 0)",
            100*hit_self/R4))
cat(sprintf("  %-46s%13.0f%%\n", "competitive GSEA (gene permutation)", 100*hit_comp/R4))
cat(sprintf("  %-46s%13.0f%%\n", "competitive ORA (hypergeometric)", 100*hit_ora/R4))
cat("\n  The self-contained test is RIGHT by its own definition - genes in the\n")
cat("  set really did change - but it answers a question you did not ask. On\n")
cat("  a global shift it declares essentially EVERY pathway significant, which\n")
cat("  is useless for interpretation. Competitive tests correctly report that\n")
cat("  nothing is special about this set. Know which one your tool runs:\n")
cat("  limma's `roast` is self-contained, `camera` is competitive, `fgsea` is\n")
cat("  competitive, and classic GSEA with sample permutation is self-contained.\n")
```

## 5. Inter-gene correlation, eq. (29.5) - the big one

Gene-label permutation assumes genes are **exchangeable and independent**.
They are not: pathway genes are co-regulated. The variance of the set's mean
statistic is inflated by exactly the design effect from Topic 1:

$$\mathrm{VIF} = 1 + (m-1)\bar\rho \tag{29.5}$$

```{r correlation}
header("5. Inter-gene correlation inflates everything (29.5)")
set.seed(3904)
simulate_correlated <- function(G = 2000, m = 100, rho = 0.10, n = 20) {
  ## Genes in the set share a latent factor -> exchangeable correlation rho.
  z <- matrix(rnorm(G*n), G)
  lat <- rnorm(n)
  z[1:m, ] <- sqrt(rho)*matrix(rep(lat, each = m), m) + sqrt(1-rho)*z[1:m, ]
  ## NO differential expression at all: both groups have the same distribution.
  z
}
R5 <- 200; grp5 <- rep(0:1, each = 10)
cat(sprintf("  %d null datasets per row; Monte Carlo SE on a 5%% rate is ~%.1f%%\n",
            R5, 100*sqrt(0.05*0.95/R5)))
cat(sprintf("  %-8s%12s%14s%18s%12s\n", "rho", "VIF (29.5)", "GSEA p<.05",
            "self-cont. p<.05", "ORA p<.05"))
for (rho in c(0.0, 0.02, 0.05, 0.10)) {
  hg <- hs <- ho <- 0
  for (i in 1:R5) {
    z <- simulate_correlated(rho = rho)
    s <- tstat(z, grp5); iset <- seq_len(nrow(z)) %in% 1:100
    r <- running_es(s, iset); e <- r[which.max(abs(r))]
    nl <- replicate(100, {q <- running_es(s, sample(iset)); q[which.max(abs(q))]})
    hg <- hg + ((1 + sum(abs(nl) >= abs(e)))/101 < 0.05)
    hs <- hs + (t.test(s[iset])$p.value < 0.05)
    de <- which(2*pt(-abs(s), 18) < 0.05)
    ho <- ho + (phyper(length(intersect(de, which(iset))) - 1, 100, nrow(z) - 100,
                       max(length(de), 1), lower.tail = FALSE) < 0.05)
  }
  cat(sprintf("  %-8.2f%12.1f%13.0f%%%17.0f%%%11.0f%%\n", rho, 1 + 99*rho,
              100*hg/R5, 100*hs/R5, 100*ho/R5))
}
cat("\n  There is NO differential expression anywhere in these simulations, so\n")
cat("  every column should read 5%. The rho = 0 row confirms the tests are\n")
cat("  calibrated when genes really are independent. A correlation of just\n")
cat("  rho = 0.05 - far below what real pathways show - already wrecks the\n")
cat("  false positive rate, because VIF = 1 + (m-1)*rho grows with SET SIZE.\n")
cat("  This is the design effect of eq. (1.8) under another name.\n")
cat("\n  Note ORA stays roughly calibrated here: thresholding at p < 0.05\n")
cat("  discards most of the shared latent signal, so ORA is less sensitive to\n")
cat("  correlation than the rank-based tests - one of its very few advantages.\n")
cat("\n  The fix is a test that estimates rho-bar and divides by sqrt(VIF) -\n")
cat("  that is precisely what limma's `camera` does - or a null that permutes\n")
cat("  SAMPLES, which preserves the correlation structure between genes.\n")
```

## 6. Sample permutation preserves gene-gene correlation

```{r sample-perm}
header("6. Sample permutation: the null that respects correlation")
set.seed(3905)
G2 <- 2000; m <- 100; n <- 24; rho <- 0.10
z <- matrix(rnorm(G2*n), G2)
lat <- rnorm(n)
z[1:m, ] <- sqrt(rho)*matrix(rep(lat, each = m), m) + sqrt(1-rho)*z[1:m, ]
grp <- rep(0:1, each = n/2)
s_obs <- tstat(z, grp); iset <- seq_len(G2) %in% 1:m
## Two statistics, two nulls each.
##   gene permutation  - reshuffles SET MEMBERSHIP, keeping the statistics fixed.
##                       Destroys the correlation among set genes.
##   sample permutation- reshuffles GROUP LABELS and recomputes the statistics.
##                       Genes keep their correlations, so the null is correct.
mean_gene <- replicate(1000, mean(s_obs[sample(iset)]))
mean_samp <- replicate(1000, {sp <- tstat(z, sample(grp)); mean(sp[iset])})
es_of <- function(s, i) {r <- running_es(s, i); r[which.max(abs(r))]}
es_gene <- replicate(500, es_of(s_obs, sample(iset)))
es_samp <- replicate(500, {sp <- tstat(z, sample(grp)); es_of(sp, iset)})
m_obs <- mean(s_obs[iset]); e_obs <- es_of(s_obs, iset)
cat(sprintf("  one null dataset: %d genes, set of %d, rho = %.2f, n = %d\n",
            G2, m, rho, n))
cat(sprintf("  VIF = 1 + (m-1)*rho = %.1f, so the correct null SD for the set\n",
            1 + (m-1)*rho))
cat(sprintf("  MEAN statistic is sqrt(VIF) = %.2f times the independence null.\n\n",
            sqrt(1 + (m-1)*rho)))
cat(sprintf("  %-22s%12s%10s%12s\n", "statistic", "null", "SD", "p-value"))
cat(sprintf("  %-22s%12s%10.4f%12.3f\n", "set mean t", "gene perm", sd(mean_gene),
            (1 + sum(abs(mean_gene) >= abs(m_obs)))/1001))
cat(sprintf("  %-22s%12s%10.4f%12.3f\n", "set mean t", "sample perm", sd(mean_samp),
            (1 + sum(abs(mean_samp) >= abs(m_obs)))/1001))
cat(sprintf("  %-22s%12s%10.4f%12.3f\n", "GSEA ES", "gene perm", sd(es_gene),
            (1 + sum(abs(es_gene) >= abs(e_obs)))/501))
cat(sprintf("  %-22s%12s%10.4f%12.3f\n", "GSEA ES", "sample perm", sd(es_samp),
            (1 + sum(abs(es_samp) >= abs(e_obs)))/501))
cat(sprintf("\n  set mean t : observed SD ratio %.2f  vs sqrt(VIF) = %.2f  <- matches\n",
            sd(mean_samp)/sd(mean_gene), sqrt(1 + (m-1)*rho)))
cat(sprintf("  GSEA ES    : observed SD ratio %.2f\n", sd(es_samp)/sd(es_gene)))
cat("\n  For the set MEAN the theory is exact: the correlation-respecting null\n")
cat("  is wider by almost exactly sqrt(VIF), which is the whole content of\n")
cat("  eq. (29.5). The ES is a maximum of a bounded random walk rather than a\n")
cat("  mean, so its null widens too but by a smaller, non-analytic factor -\n")
cat("  a useful reminder that VIF quantifies the variance of an AVERAGE, and\n")
cat("  does not transfer unchanged to every other statistic.\n")
cat("\n  Sample permutation needs enough samples to permute: with n per group\n")
cat("  below ~7 there are too few distinct label assignments for a two-sided\n")
cat("  p to even reach 0.05, which is why small studies are pushed towards\n")
cat("  `camera`-style analytic corrections instead.\n")
```

## 7. Selection bias: length, expression and set size

```{r bias}
header("7. Detection bias: long, highly expressed genes win by default")
set.seed(3906)
G3 <- 6000
len <- exp(rnorm(G3, log(2000), 0.6))          # gene length, bp
expr <- exp(rnorm(G3, log(200), 1.1))          # mean count per sample
## EVERY gene has exactly the same true log2 fold change. The only thing that
## differs is PRECISION: counts scale with length x expression, and a Poisson
## count of size mu has relative SE 1/sqrt(mu). So the standard error of the
## log fold change shrinks for long, highly expressed genes.
LFC_TRUE <- 0.35
counts <- expr*(len/2000)
se_lfc <- sqrt(2/counts) + 0.05                # +0.05: irreducible biological CV
lfc_hat <- rnorm(G3, LFC_TRUE, se_lfc)
p3 <- 2*pnorm(-abs(lfc_hat/se_lfc))
de3 <- p3 < 0.05
cat(sprintf("  %d genes, ALL with the same true log2 FC of %.2f.\n", G3, LFC_TRUE))
cat(sprintf("  %d (%.0f%%) reach p < 0.05.\n\n", sum(de3), 100*mean(de3)))
cat(sprintf("  %-26s%12s%12s\n", "", "DE genes", "non-DE"))
cat(sprintf("  %-26s%12.0f%12.0f\n", "median length (bp)",
            median(len[de3]), median(len[!de3])))
cat(sprintf("  %-26s%12.0f%12.0f\n", "median expression",
            median(expr[de3]), median(expr[!de3])))
cat(sprintf("  %-26s%12.3f%12.3f\n", "median SE of logFC",
            median(se_lfc[de3]), median(se_lfc[!de3])))
cat("\n  No gene is more differential than any other, yet the DE list is\n")
cat("  systematically enriched for long, highly expressed genes - purely\n")
cat("  because they are measured more precisely. A pathway made of such genes\n")
cat("  (ribosome, spliceosome, oxidative phosphorylation) will look enriched\n")
cat("  for reasons that have nothing to do with biology.\n")
cat("\n  `goseq` fixes this by fitting a spline for P(DE | length) and using a\n")
cat("  length-weighted null instead of the plain hypergeometric. GSEA on a\n")
cat("  ranked list is less affected but not immune, since power still shapes\n")
cat("  the ranking that GSEA walks down.\n")
cat("\n  Set SIZE is the other systematic bias:\n")
cat(sprintf("  %-14s%16s%18s\n", "set size m", "VIF at rho=0.05", "independent genes"))
for (mm in c(10, 50, 200, 1000))
  cat(sprintf("  %-14d%16.1f%18.1f\n", mm, 1 + (mm-1)*0.05, mm/(1 + (mm-1)*0.05)))
cat("  A 1000-gene GO term carries only ~20 genes' worth of INDEPENDENT\n")
cat("  information. Restrict to sets of ~15-500 genes: smaller sets are\n")
cat("  unstable, larger ones are both correlation-inflated and too vague to\n")
cat("  interpret even when real.\n")
```

## 8. Multiple testing across overlapping sets

```{r multiple}
header("8. Thousands of OVERLAPPING sets")
set.seed(3907)
## Build a GO-like hierarchy: one broad parent term, split into children, split
## again into grandchildren. A parent contains every gene of its descendants,
## so these tests are not remotely independent.
G8 <- 5000
parent <- 1:600
children <- split(parent, rep(1:6, each = 100))
grandkids <- unlist(lapply(children, function(ch) split(ch, rep(1:4, each = 25))),
                    recursive = FALSE)
## Plus 1000 unrelated background terms, as a real MSigDB run would have.
others <- lapply(1:1000, function(i) sample(601:G8, sample(20:200, 1)))
sets <- c(list(parent = parent), children, grandkids, others)
names(sets) <- c("parent", paste0("child", 1:6),
                 paste0("grand", 1:24), paste0("bg", 1:1000))
## ONE biological truth: a single child term (100 genes) is really affected.
## Its 4 grandchildren are subsets of it, and `parent` contains it - so those
## 5 other terms are enriched purely by inclusion, not by separate biology.
truth8 <- rep(0, G8); truth8[children[[1]]] <- 2.0
s8 <- rnorm(G8, truth8)
p8 <- sapply(sets, function(g) {
  ov <- intersect(which(2*pnorm(-abs(s8)) < 0.05), g)
  de <- which(2*pnorm(-abs(s8)) < 0.05)
  phyper(length(ov) - 1, length(g), G8 - length(g), length(de), lower.tail = FALSE)
})
padj8 <- p.adjust(p8, "BH")
sig <- names(padj8)[padj8 < 0.05]
cat(sprintf("  %d sets tested. TRUTH: exactly ONE 100-gene term (child1) is\n",
            length(sets)))
cat("  affected. Every other term is either disjoint from it or related to it\n")
cat("  only by containment.\n")
cat(sprintf("  %-40s%14s\n", "", "discoveries"))
cat(sprintf("  %-40s%14d\n", "raw p < 0.05", sum(p8 < 0.05)))
cat(sprintf("  %-40s%14d\n", "BH q < 0.05", sum(padj8 < 0.05)))
cat(sprintf("  %-40s%14d\n", "BY q < 0.05", sum(p.adjust(p8, "BY") < 0.05)))
cat(sprintf("  %-40s%14d\n", "Bonferroni", sum(p.adjust(p8, "bonferroni") < 0.05)))
lineage <- c("parent", "child1", paste0("grand", 1:4))
cat(sprintf("\n  BH-significant terms (%d): %s\n", length(sig),
            paste(sig, collapse = ", ")))
cat(sprintf("  in the child1 lineage : %d\n", sum(sig %in% lineage)))
cat(sprintf("  outside it (genuine false positives): %d\n", sum(!sig %in% lineage)))
cat("\n  Almost every discovery is the true term, one of its ANCESTORS, or one\n")
cat("  of its CHILDREN - all significant only because they share genes with\n")
cat("  child1. The FDR control is not wrong: these sets really are enriched.\n")
cat("  But reporting them as separate discoveries multiplies ONE biological\n")
cat("  finding into several, which is how enrichment tables end up with 40\n")
cat("  rows that all say 'immune response'.\n")
cat("\n  Two separate issues, often confused:\n")
cat("    VALIDITY - BH assumes positive regression dependence (Topic 6). GO's\n")
cat("      nesting induces exactly that, so BH is fine. BY is valid under any\n")
cat("      dependence but is far too conservative to be worth it here.\n")
cat("    INTERPRETATION - the discoveries are REDUNDANT. Collapse them before\n")
cat("      counting: topGO's `elim`/`weight` algorithms test a term\n")
cat("      conditional on its children, and REVIGO-style semantic clustering\n")
cat("      groups terms after the fact. Report the most specific significant\n")
cat("      term in each lineage, not the whole lineage.\n")
```

## 9. Figure

```{r figure}
png(file.path(OUT, "gene_set_enrichment.png"), width = 1100, height = 800, res = 110)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.5, 1))
plot(res, type = "l", col = "steelblue", lwd = 2, xlab = "rank in ordered list",
     ylab = "running ES", main = sprintf("GSEA walk (ES = %+.3f)", es))
abline(h = 0, lty = 3)
rug(which(in_set[order(stat, decreasing = TRUE)]), col = "firebrick", lwd = 0.5)
hist(null_es, breaks = 40, col = adjustcolor("grey60", 0.7), border = "white",
     main = "Gene-permutation null", xlab = "ES")
abline(v = es, col = "firebrick", lwd = 2)
plot(density(mean_samp), col = "firebrick", lwd = 2, xlab = "set mean t",
     main = "Correlation widens the null", ylim = c(0, max(density(mean_gene)$y)))
lines(density(mean_gene), col = "steelblue", lwd = 2)
legend("topright", c("sample perm.","gene perm."), col = c("firebrick","steelblue"),
       lwd = 2, bty = "n", cex = 0.7)
curve(1 + (x - 1)*0.05, 1, 500, col = "firebrick", lwd = 2, log = "x",
      xlab = "set size m", ylab = "VIF", main = "Eq. (29.5): VIF = 1+(m-1)rho")
curve(1 + (x - 1)*0.02, 1, 500, col = "steelblue", lwd = 2, add = TRUE)
legend("topleft", c("rho = 0.05","rho = 0.02"), col = c("firebrick","steelblue"),
       lwd = 2, bty = "n", cex = 0.7)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "gene_set_enrichment.png"), "\n")
```

# PROBLEMS

### Problem 1: The background set

You profile brain tissue; 12,000 of 20,000 genes are expressed. A neuronal
pathway (200 genes, 190 expressed in brain) overlaps your 400-gene DE list
in 20 genes. Compute ORA with the whole genome vs the expressed genes.

```{r problem1}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# k <- 20; n <- 400
# cat(sprintf("  %-34s%8s%8s%12s%10s\n","background","N","K","expected","p"))
# for (bg in list(c("whole genome", 20000, 200), c("expressed in brain", 12000, 190))) {
#   Nb <- as.numeric(bg[2]); Kb <- as.numeric(bg[3])
#   cat(sprintf("  %-34s%8.0f%8.0f%12.2f%10.2e\n", bg[1], Nb, Kb, n*Kb/Nb,
#               phyper(k - 1, Kb, Nb - Kb, n, lower.tail = FALSE)))
# }
#
# ## The genome background asks 'are these genes over-represented relative to
# ## ALL human genes?' - and neuronal genes are, in brain, trivially. The
# ## correct background is the set of genes that could have entered the DE list
# ## at all: those that passed your expression filter and were actually tested.
# ## The genome background is the commonest reason a pathway analysis
# ## returns 'neuron differentiation' from a brain experiment.
```

### Problem 2: Set size and the VIF

Measure the false positive rate of a competitive test under a global null
with rho = 0.05, for set sizes 20, 100 and 400. Compare to eq. (29.5).

```{r problem2}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# set.seed(77)
# cat(sprintf("  %-12s%12s%16s\n", "set size m", "VIF (29.5)", "self-cont. FPR"))
# for (m_ in c(20, 100, 400)) {
#   hits <- 0; R <- 120
#   for (i in 1:R) {
#     nn <- 20; Gg <- 2000
#     zz <- matrix(rnorm(Gg*nn), Gg); la <- rnorm(nn)
#     zz[1:m_, ] <- sqrt(0.05)*matrix(rep(la, each = m_), m_) + sqrt(0.95)*zz[1:m_, ]
#     gg <- rep(0:1, each = nn/2)
#     ss <- tstat(zz, gg)
#     hits <- hits + (t.test(ss[1:m_])$p.value < 0.05)
#   }
#   cat(sprintf("  %-12d%12.1f%15.0f%%\n", m_, 1 + (m_-1)*0.05, 100*hits/R))
# }
#
# ## The FPR climbs with set size exactly as VIF = 1 + (m-1)*rho predicts,
# ## because the effective number of independent genes is m/VIF, not m. A
# ## 400-gene set at rho = 0.05 carries only ~20 genes' worth of independent
# ## information. This is why big, vague GO terms dominate enrichment results.
```

### Problem 3: Leading edge, and what it is not

Extract the leading-edge genes (those before the ES peak) from section 3.
What fraction of the true pathway do they recover, and how stable are they?

```{r problem3}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# o <- order(stat, decreasing = TRUE)
# peak <- which.max(abs(res))
# lead <- o[1:peak][in_set[o[1:peak]]]
# cat(sprintf("  ES peak at rank %d; leading edge = %d genes\n", peak, length(lead)))
# cat(sprintf("  recovers %.0f%% of the %d true pathway genes\n",
#             100*length(lead)/length(pathway), length(pathway)))
# set.seed(5)
# jac <- replicate(30, {
#   s2 <- rnorm(G, truth)                     # a replicate experiment
#   r2 <- running_es(s2, in_set); o2 <- order(s2, decreasing = TRUE)
#   l2 <- o2[1:which.max(abs(r2))][in_set[o2[1:which.max(abs(r2))]]]
#   length(intersect(lead, l2))/length(union(lead, l2))
# })
# cat(sprintf("  Jaccard overlap with 30 replicate experiments: %.2f (SD %.2f)\n",
#             mean(jac), sd(jac)))
#
# ## The leading edge is UNSTABLE: repeat the experiment and you get a
# ## substantially different gene list, because which genes fall before the peak
# ## depends on noise in the ranking. Treat it as a hypothesis-generating
# ## summary of WHY a set scored, never as 'the genes driving the pathway'.
```

## What to take away

1. The **background** is the genes you could have detected. Nothing else.
2. ORA discards the ranking and hinges on an arbitrary cutoff; prefer a
   rank-based test when effects are coordinated and modest.
3. Know whether your test is **competitive** or **self-contained** - they
   answer different questions and disagree on a global shift.
4. Inter-gene correlation inflates competitive tests by
   $\mathrm{VIF} = 1+(m-1)\bar\rho$ - the design effect again. Use `camera`
   or permute **samples**, not genes.
5. Length and expression bias the DE list; `goseq` reweights for it.
6. BH across overlapping GO terms is valid but the terms are redundant -
   collapse before counting.

**Next:** `40_networks_and_multiomics.R`
