---
title: "Applied 40 - Networks and multi-omics integration"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 30, equations (30.1)-(30.6)

## The question

We have RNA, methylation and protein on the same patients. Which molecular
**modules** exist, which are shared across assays, and would any of it be
found again in a second cohort?

The recurring theme: networks and factor models have enormous numbers of
parameters and very few patients. Almost everything they produce looks
structured whether or not there is structure to find.

```{r setup, message = FALSE}
MODULE_NAME <- "40_networks_and_multiomics"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. A multi-omics cohort built from known latent factors, eq. (30.6)

The generative model behind every integration method (MOFA, iCluster, JIVE):
a few latent factors $\mathbf{z}$ drive all assays through loadings
$\mathbf{W}^{(a)}$, plus assay-specific noise:

$$\mathbf{X}^{(a)} = \mathbf{Z}\,\mathbf{W}^{(a)\top} + \mathbf{E}^{(a)} \tag{30.6}$$

```{r simulate}
header("1. A MultiAssayExperiment-like cohort (30.6)")
set.seed(4001)
simulate_multiomics <- function(n = 60, p_rna = 400, p_met = 300, p_pro = 150) {
  ## Four latent factors, only some of which touch every assay.
  z_shared  <- rnorm(n)      # biology visible in ALL assays
  z_rna_met <- rnorm(n)      # regulatory axis: methylation -> expression only
  z_batch   <- rnorm(n)      # technical, RNA only
  z_pro     <- rnorm(n)      # protein-specific (post-translational)
  Z <- cbind(shared = z_shared, rna_met = z_rna_met, batch = z_batch, pro = z_pro)
  load <- function(p, idx, strength) {
    W <- matrix(0, p, 4)
    for (j in seq_along(idx)) if (length(idx[[j]]))
      W[idx[[j]], j] <- strength[j]*sign(rnorm(length(idx[[j]])))
    W
  }
  W_rna <- load(p_rna, list(1:80, 81:160, 161:220, integer(0)), c(1.4, 1.2, 1.0, 0))
  W_met <- load(p_met, list(1:60, 61:140, integer(0), integer(0)), c(1.1, 1.3, 0, 0))
  W_pro <- load(p_pro, list(1:40, integer(0), integer(0), 41:90), c(1.2, 0, 0, 1.3))
  list(
    rna = Z %*% t(W_rna) + matrix(rnorm(n*p_rna), n),
    met = Z %*% t(W_met) + matrix(rnorm(n*p_met), n),
    pro = Z %*% t(W_pro) + matrix(rnorm(n*p_pro), n),
    Z = Z,
    ## An outcome driven by the shared factor only - the thing worth finding.
    y = 0.9*z_shared + rnorm(n, 0, 1)
  )
}
d <- simulate_multiomics()
cat(sprintf("  n = %d patients\n", nrow(d$rna)))
for (a in c("rna","met","pro"))
  cat(sprintf("  %-4s : %4d features\n", a, ncol(d[[a]])))
cat(sprintf("\n  total features = %d, patients = %d  -> p/n = %.0f\n",
            ncol(d$rna) + ncol(d$met) + ncol(d$pro), nrow(d$rna),
            (ncol(d$rna) + ncol(d$met) + ncol(d$pro))/nrow(d$rna)))
cat("  4 latent factors: shared (all assays), rna_met (two assays),\n")
cat("  batch (RNA only, technical), pro (protein only).\n")
cat("\n  This p >> n regime is the whole difficulty. A correlation matrix on\n")
cat(sprintf("  400 genes has %s free parameters estimated from %d patients.\n",
            format(400*399/2, big.mark = ","), nrow(d$rna)))
```

## 2. Co-expression networks, eq. (30.1)-(30.3)

WGCNA's recipe: soft-threshold the correlation into an adjacency,
$$a_{ij} = |r_{ij}|^\beta \tag{30.1}$$
convert to a topological overlap matrix that counts shared neighbours,
then cluster the resulting dissimilarity.

```{r network}
header("2. Soft thresholding, TOM and module eigengenes (30.1)-(30.3)")
X <- scale(d$rna)
r <- cor(X)
cat("  Soft thresholding (eq. 30.1) raises |r| to a power beta, so weak edges\n")
cat("  fade smoothly instead of being cut at an arbitrary hard threshold.\n\n")
cat(sprintf("  %-8s%16s%16s%14s\n", "beta", "mean adjacency", "|r|>0.3 kept", "scale-free R2"))
for (beta in c(1, 3, 6, 12)) {
  A <- abs(r)^beta; diag(A) <- 0
  k <- rowSums(A)
  ## WGCNA picks beta by how well the degree distribution follows a power law.
  h <- hist(k[k > 0], breaks = 12, plot = FALSE)
  ok <- h$counts > 0
  fit <- lm(log10(h$counts[ok]) ~ log10(h$mids[ok]))
  cat(sprintf("  %-8d%16.4f%15.1f%%%14.3f\n", beta, mean(A),
              100*mean(abs(r)[abs(r) > 0.3]^beta > 0.1),
              summary(fit)$r.squared))
}
BETA <- 6
A <- abs(r)^BETA; diag(A) <- 0
## TOM (eq. 30.2): two genes are close if they share neighbours, not only if
## they correlate. This is much more robust to single spurious edges.
L <- A %*% A
k <- rowSums(A)
TOM <- (L + A)/(outer(k, k, pmin) + 1 - A)
diag(TOM) <- 1
dissTOM <- 1 - TOM
## Ward linkage on the TOM dissimilarity. Average linkage - WGCNA's default
## before dynamic tree cutting - chains the 180 unstructured noise genes onto
## the real modules and returns one giant cluster plus singletons; Ward's
## variance criterion keeps the modules balanced. WGCNA solves the same problem
## with `cutreeDynamic` rather than a fixed height.
hc <- hclust(as.dist(dissTOM), method = "ward.D2")
mods <- cutree(hc, k = 4)
truth_mod <- c(rep(1, 80), rep(2, 80), rep(3, 60), rep(0, 180))
tab <- table(module = mods, truth = truth_mod)
cat("\n  Modules from TOM dissimilarity vs the true factor assignment\n")
cat("  (truth 0 = noise genes, loading on nothing):\n")
print(tab)
## Module eigengene (eq. 30.3) = first PC of the module's expression.
eig <- sapply(sort(unique(mods)), function(m) {
  sub <- X[, mods == m, drop = FALSE]
  pc <- prcomp(sub, center = TRUE, scale. = FALSE)
  pc$x[, 1]
})
cat("\n  Module eigengenes (eq. 30.3) vs the TRUE latent factors, |correlation|:\n")
cm <- abs(cor(eig, d$Z))
colnames(cm) <- colnames(d$Z); rownames(cm) <- paste0("ME", sort(unique(mods)))
print(round(cm, 3))
cat("\n  Three eigengenes track one latent factor each, and the fourth - the\n")
cat("  large module of unstructured genes - tracks nothing. That module is\n")
cat("  WGCNA's 'grey' module: not a pathway, just the leftovers. Always check\n")
cat("  whether a module HAS a coherent eigengene before interpreting it.\n")
cat("\n  More important: the eigengene for the BATCH factor is exactly as clean\n")
cat("  as the biological ones. A co-expression module is a module of\n")
cat("  CORRELATED genes, and technical artefacts correlate genes beautifully.\n")
cat("  A module is not evidence of a pathway.\n")
```

## 3. Marginal vs partial correlation, eq. (10.5)-(10.6), (30.4)

A correlation network draws an edge whenever two genes covary - including
when both are driven by a third. A **partial** correlation conditions on
everything else and keeps only direct edges:

$$\rho_{ij\cdot\text{rest}} = \frac{-\Theta_{ij}}{\sqrt{\Theta_{ii}\Theta_{jj}}},
  \quad \Theta = \Sigma^{-1} \tag{30.4}$$

```{r partial}
header("3. Direct vs indirect edges (30.4)")
set.seed(4002)
## A chain: h -> g1, g2, g3. NO gene connects to another directly.
n3 <- 200
h <- rnorm(n3)
g <- cbind(g1 = 0.8*h + rnorm(n3, 0, 0.6),
           g2 = 0.8*h + rnorm(n3, 0, 0.6),
           g3 = 0.8*h + rnorm(n3, 0, 0.6))
cat("  Truth: ONE hidden driver h causes g1, g2, g3. No gene->gene edge exists.\n\n")
cat("  marginal correlations (what a co-expression network sees):\n")
print(round(cor(g), 3))
Theta <- solve(cor(g))
pc3 <- -Theta/sqrt(outer(diag(Theta), diag(Theta)))
diag(pc3) <- 1
cat("\n  partial correlations, h NOT measured (eq. 30.4):\n")
print(round(pc3, 3))
Theta2 <- solve(cor(cbind(g, h = h)))
pc4 <- -Theta2/sqrt(outer(diag(Theta2), diag(Theta2)))
diag(pc4) <- 1
cat("\n  partial correlations WITH h included:\n")
print(round(pc4, 3))
cat("\n  With h unmeasured, every pair is strongly correlated AND retains a\n")
cat("  sizeable partial correlation - the network shows a triangle of edges\n")
cat("  that do not exist. Measure h and the gene-gene partials collapse\n")
cat("  towards zero, which is the correct answer.\n")
cat("\n  Estimating (30.4) needs Sigma inverted, impossible when p > n. Graphical\n")
cat("  lasso penalises Theta to make it sparse and invertible - but you are\n")
cat("  then ASSUMING sparsity, and a hidden driver still creates phantom edges.\n")
cat("  No network method recovers causal structure from observational data\n")
cat("  alone (Topic 32); it recovers conditional independence, at best.\n")
```

## 4. Multi-block factor analysis, eq. (30.6)

```{r mofa}
header("4. Recovering latent factors across assays (30.6)")
## MOFA-style integration, done simply: scale each block so no assay dominates
## by virtue of having more features or a larger variance, then take an SVD of
## the concatenation. Real MOFA uses a sparse Bayesian model, but the estimand
## and the pitfalls are the same.
blocks <- list(rna = scale(d$rna), met = scale(d$met), pro = scale(d$pro))
cat(sprintf("  %-6s%10s%18s%22s\n", "block", "features", "total variance",
            "weight if unscaled"))
tot <- sapply(blocks, function(B) sum(apply(B, 2, var)))
for (a in names(blocks))
  cat(sprintf("  %-6s%10d%18.0f%21.0f%%\n", a, ncol(blocks[[a]]), tot[a],
              100*tot[a]/sum(tot)))
## Divide each block by sqrt(its total variance): every assay now contributes
## equally regardless of feature count. Skipping this makes "integration" mean
## "analyse whichever assay has the most probes".
Xc <- do.call(cbind, lapply(blocks, function(B) B/sqrt(sum(apply(B, 2, var)))))
sv <- svd(Xc)
F_est <- sv$u[, 1:4] %*% diag(sv$d[1:4])
cat("\n  |correlation| between estimated factors and the TRUTH:\n")
cm2 <- abs(cor(F_est, d$Z))
colnames(cm2) <- colnames(d$Z); rownames(cm2) <- paste0("Factor", 1:4)
print(round(cm2, 3))
## Variance explained per block, per factor - the plot MOFA is known for, and
## the one that tells you whether a factor is shared or assay-specific.
cat("\n  variance explained (%) by factor within each block:\n")
cat(sprintf("  %-10s%10s%10s%10s\n", "", "rna", "met", "pro"))
starts <- c(0, cumsum(sapply(blocks, ncol)))
for (f in 1:4) {
  u <- sv$u[, f, drop = FALSE]
  ve <- sapply(seq_along(blocks), function(b) {
    cols <- (starts[b] + 1):starts[b + 1]
    Xb <- Xc[, cols, drop = FALSE]
    proj <- u %*% (t(u) %*% Xb)
    100*sum(proj^2)/sum(Xb^2)
  })
  cat(sprintf("  Factor%-4d%10.1f%10.1f%10.1f\n", f, ve[1], ve[2], ve[3]))
}
cat("\n  Read this table as the MAIN OUTPUT of an integration. A factor loading\n")
cat("  on all three blocks is shared biology; one loading on RNA alone is\n")
cat("  either RNA-specific biology or - far more often - a technical artefact\n")
cat("  of that assay. Our injected batch factor is RNA-only by construction.\n")
```

## 5. Module and factor stability, and the n problem

```{r stability}
header("5. Would this structure appear again?")
set.seed(4003)
jaccard <- function(a, b) length(intersect(a, b))/length(union(a, b))
stability_run <- function(Xd, frac = 0.75, B = 30, K = 4) {
  base <- cutree(hclust(as.dist(1 - {A <- abs(cor(Xd))^6; diag(A) <- 0
                                     k <- rowSums(A)
                                     T <- (A %*% A + A)/(outer(k, k, pmin) + 1 - A)
                                     diag(T) <- 1; T}), method = "ward.D2"), K)
  sapply(1:B, function(b) {
    idx <- sample(nrow(Xd), round(frac*nrow(Xd)))
    Xs <- Xd[idx,, drop = FALSE]
    A <- abs(cor(Xs))^6; diag(A) <- 0
    k <- rowSums(A)
    T <- (A %*% A + A)/(outer(k, k, pmin) + 1 - A); diag(T) <- 1
    m <- cutree(hclust(as.dist(1 - T), method = "ward.D2"), K)
    ## For each true module, the best-matching resampled module.
    mean(sapply(sort(unique(base)), function(g)
      max(sapply(sort(unique(m)), function(gg)
        jaccard(which(base == g), which(m == gg))))))
  })
}
cat("  Resample 75% of PATIENTS, rebuild the network, match modules:\n")
cat(sprintf("  %-34s%12s%12s\n", "data", "mean Jaccard", "SD"))
st_real <- stability_run(X)
cat(sprintf("  %-34s%12.3f%12.3f\n", "real structure (n = 60)", mean(st_real), sd(st_real)))
X_null <- matrix(rnorm(nrow(X)*120), nrow(X))
st_null <- stability_run(X_null)
cat(sprintf("  %-34s%12.3f%12.3f\n", "PURE NOISE, same n, 120 genes",
            mean(st_null), sd(st_null)))
X_small <- X[sample(nrow(X), 20), ]
st_small <- stability_run(X_small)
cat(sprintf("  %-34s%12.3f%12.3f\n", "real structure, n = 20 only",
            mean(st_small), sd(st_small)))
cat("\n  Pure noise still yields modules, and they are somewhat reproducible\n")
cat("  under resampling, because resamples share most of their patients. A\n")
cat("  stability score is therefore only interpretable AGAINST A NULL of the\n")
cat("  same size - never on its own. And with n = 20 even genuine structure\n")
cat("  becomes unstable: correlation networks need patients, not features.\n")
```

## 6. Cross-validation at the biological-unit level

```{r cv}
header("6. If a factor is used to predict, validate at the patient level")
set.seed(4004)
cv_r2 <- function(build, y, n, K = 5) {
  ## A single 5-fold split of 60 patients is very noisy, so every number below
  ## is averaged over REPS independent fold assignments.
  fold <- sample(rep_len(1:K, n))
  pred <- numeric(n)
  for (k in 1:K) {
    tr <- fold != k; te <- fold == k
    f <- build(tr, te)
    m <- lm(y[tr] ~ f$train)
    pred[te] <- cbind(1, f$test) %*% coef(m)
  }
  1 - sum((y - pred)^2)/sum((y - mean(y))^2)
}
## (a) LEAKY: factors estimated once on ALL patients, then cross-validated.
sv_all <- svd(scale(d$rna))
f_all <- sv_all$u[, 1]*sv_all$d[1]
REPS <- 20
avg <- function(f, y) mean(replicate(REPS, cv_r2(f, y, nrow(d$rna))))
leaky_build <- function(tr, te) list(train = f_all[tr], test = f_all[te])
honest_build <- function(tr, te) {
  Xtr <- scale(d$rna[tr, ])
  s <- svd(Xtr)
  ## Project the held-out patients using the TRAINING centring/scaling and
  ## the TRAINING loadings. Re-scaling on the test set would leak again.
  Xte <- scale(d$rna[te, ], center = attr(Xtr, "scaled:center"),
               scale = attr(Xtr, "scaled:scale"))
  list(train = s$u[, 1]*s$d[1], test = Xte %*% s$v[, 1])
}
r2_leaky <- avg(leaky_build, d$y)
## (b) HONEST: the factor is re-derived inside each fold and the test patients
## are PROJECTED onto the training loadings.
r2_honest <- avg(honest_build, d$y)
## (c) The same comparison when the outcome is PURE NOISE.
y_null <- rnorm(nrow(d$rna))
r2_leaky_null <- avg(leaky_build, y_null)
r2_honest_null <- avg(honest_build, y_null)
cat(sprintf("  %-44s%12s%12s\n", "factor estimated", "R2 (real y)", "R2 (null y)"))
cat(sprintf("  %-44s%12.3f%12.3f\n", "on ALL patients, then CV (leaky)",
            r2_leaky, r2_leaky_null))
cat(sprintf("  %-44s%12.3f%12.3f\n", "inside each fold, test projected (honest)",
            r2_honest, r2_honest_null))
cat("\n  Here the leak is small, and that is worth understanding rather than\n")
cat("  celebrating: the first PC of a 400-gene matrix is driven by strong,\n")
cat("  widely shared factors, so it barely moves when 20% of patients are\n")
cat("  removed, and it is estimated WITHOUT using y at all. Unsupervised\n")
cat("  feature construction leaks far less than SUPERVISED selection did in\n")
cat("  Modules 21 and 38 - it is the use of y outside the fold that is deadly.\n")
cat("\n  The null-outcome column is the real check: both columns sit near or\n")
cat("  below 0, as they must when y is noise. Report it in every analysis -\n")
cat("  a pipeline that cannot produce ~0 on a permuted outcome is broken.\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "networks_multiomics.png"), width = 1150, height = 850, res = 110)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.5, 1))
ord <- order(mods)
image(t(abs(r)[ord, rev(ord)]), col = hcl.colors(50, "YlOrRd", rev = TRUE),
      axes = FALSE, main = "|correlation|, genes ordered by module")
box()
plot(hc, labels = FALSE, hang = -1, main = "TOM dissimilarity (eq. 30.2)",
     xlab = "", sub = "", cex = 0.4)
image(1:4, 1:3, sapply(1:3, function(b) {
  cols <- (starts[b] + 1):starts[b + 1]
  sapply(1:4, function(f) {
    u <- sv$u[, f, drop = FALSE]; Xb <- Xc[, cols, drop = FALSE]
    100*sum((u %*% (t(u) %*% Xb))^2)/sum(Xb^2)
  })
}), col = hcl.colors(50, "Blues", rev = TRUE), xlab = "factor", ylab = "",
    axes = FALSE, main = "Variance explained per block (30.6)")
axis(1, 1:4, paste0("F", 1:4)); axis(2, 1:3, names(blocks), las = 1); box()
boxplot(list(real = st_real, noise = st_null, `n=20` = st_small),
        col = c("steelblue","grey70","darkorange"), ylab = "mean Jaccard",
        main = "Module stability needs a null")
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "networks_multiomics.png"), "\n")
```

# PROBLEMS

### Problem 1: Module-trait associations need a null

Correlate every module eigengene with a RANDOM trait. How often does at
least one module reach p < 0.05, and what is the right way to report it?

```{r problem1}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# set.seed(21)
# R <- 500; K <- ncol(eig)
# any_hit <- min_p <- numeric(R)
# for (i in 1:R) {
#   trait <- rnorm(nrow(eig))
#   ps <- apply(eig, 2, function(e) cor.test(e, trait)$p.value)
#   any_hit[i] <- any(ps < 0.05); min_p[i] <- min(ps)
# }
# cat(sprintf("  %d modules, %d random traits:\n", K, R))
# cat(sprintf("    P(at least one module p < 0.05)        = %.1f%%\n", 100*mean(any_hit)))
# cat(sprintf("    expected if independent, 1-(0.95)^%d    = %.1f%%\n", K,
#             100*(1 - 0.95^K)))
# cat(sprintf("    P(min p survives Bonferroni over %d)   = %.1f%%\n", K,
#             100*mean(min_p < 0.05/K)))
#
# ## A WGCNA module-trait heatmap is a grid of correlations, each with its own
# ## p-value, and the eye goes straight to the largest. With K modules and
# ## several traits that is dozens of implicit tests. Correct over the WHOLE
# ## GRID, and remember the modules were themselves derived from these data -
# ## so even a corrected p is optimistic. Validation in a second cohort is the
# ## only real evidence.
```

### Problem 2: Would you have found the batch factor without the labels?

Pretend you do not know which factor is technical. Using only the variance
decomposition and the assay structure, which factor would you flag?

```{r problem2}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# ve_mat <- t(sapply(1:4, function(f) {
#   u <- sv$u[, f, drop = FALSE]
#   sapply(1:3, function(b) {
#     cols <- (starts[b] + 1):starts[b + 1]; Xb <- Xc[, cols, drop = FALSE]
#     100*sum((u %*% (t(u) %*% Xb))^2)/sum(Xb^2)
#   })
# }))
# colnames(ve_mat) <- names(blocks); rownames(ve_mat) <- paste0("Factor", 1:4)
# spec <- apply(ve_mat, 1, function(v) max(v)/sum(v))
# for (f in 1:4)
#   cat(sprintf("  Factor%d: rna %.1f%% met %.1f%% pro %.1f%%  specificity %.2f  %s\n",
#               f, ve_mat[f,1], ve_mat[f,2], ve_mat[f,3], spec[f],
#               if (spec[f] > 0.8) "<- single-assay: SUSPECT" else ""))
# cat(sprintf("\n  true correlation of each factor with the known batch variable:\n"))
# print(round(abs(cor(F_est, d$Z[, "batch"]))[, 1], 3))
#
# ## A factor confined to ONE assay is the signature of a technical artefact,
# ## because real biology usually leaves a trace in more than one molecular
# ## layer. This is a HYPOTHESIS, not proof: the protein-specific factor here
# ## is genuine biology (post-translational regulation) and looks identical by
# ## this criterion. Single-assay factors are the ones to check against
# ## processing date, batch, RIN and library size - which is why you must
# ## record that metadata (Module 19).
```

### Problem 3: Preservation in an independent cohort

Simulate a second cohort from the same generative model and ask whether the
modules found in cohort 1 are preserved. Compare with a scrambled control.

```{r problem3}
```
```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# set.seed(33)
# d2 <- simulate_multiomics(n = 60)
# X2 <- scale(d2$rna)
# ## Preservation statistic: mean |correlation| WITHIN each cohort-1 module,
# ## evaluated in cohort 2. A preserved module stays densely connected.
# dens <- function(Xd, mem) sapply(sort(unique(mem)), function(m) {
#   cc <- abs(cor(Xd[, mem == m, drop = FALSE]))
#   mean(cc[upper.tri(cc)])
# })
# d1_dens <- dens(X, mods); d2_dens <- dens(X2, mods)
# ## Null: random modules of the SAME SIZES in cohort 2.
# null_dens <- replicate(100, dens(X2, sample(mods)))
# cat(sprintf("  %-10s%8s%12s%12s%14s%10s\n", "module", "size", "dens coh1",
#             "dens coh2", "null mean", "Z"))
# for (m in sort(unique(mods))) {
#   nd <- null_dens[m, ]
#   cat(sprintf("  %-10d%8d%12.3f%12.3f%14.3f%10.1f\n", m, sum(mods == m),
#               d1_dens[m], d2_dens[m], mean(nd), (d2_dens[m] - mean(nd))/sd(nd)))
# }
#
# ## The Z score against a size-matched null is the whole point: raw density
# ## depends on module size, so it cannot be compared across modules directly.
# ## This is exactly what WGCNA's `modulePreservation` computes. A module with
# ## a high within-cohort density but a Z near 0 in the second cohort was an
# ## artefact of cohort 1. Preservation in an INDEPENDENT cohort - not
# ## stability under resampling - is what makes a module believable.
```

## What to take away

1. A co-expression module is a set of **correlated** genes. Batch effects
   correlate genes just as well as pathways do.
2. Correlation networks show indirect edges; partial correlation (30.4)
   removes them, but only for variables you actually measured.
3. Scale blocks before integrating, or "multi-omics" means "the biggest assay".
4. The variance-explained-per-block table is the main output: single-assay
   factors are the ones to check against technical metadata.
5. Stability under resampling is meaningless without a same-size null, and
   these methods need **patients**, not features.
6. Unsupervised factors leak little; using $y$ outside the fold leaks a lot.
7. Preservation in an independent cohort is the only strong evidence.

**This is the final applied module.** See `statsR/exercises/` for integrative
problem sets that combine several topics at once.
