---
title: "Applied 42 - Deconvolution, batch integration, and zero-inflation"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 42, equations (42.1)-(42.6)
**Core modules used:** 11, 19, 13, 26, 36

## Dataset card

| | |
|---|---|
| **Real analogue** | Bulk RNA-seq deconvolved with CIBERSORTx; scRNA-seq integrated with Harmony or MNN |
| **Key threats** | Wrong signature; collinear cell types; compositional output; over-correction; imaginary zero-inflation |
| **Here** | Simulated mixtures with known proportions, and UMI counts with a known zero mechanism |

## Three claims to test

1. Deconvolution is a constrained regression, and it inherits every problem
   regression has with collinear predictors.
2. Integration trades mixing against biology, and one metric alone can
   always be made to look good.
3. Zero-inflation in UMI data is mostly a myth, and testable.

```{r setup, message = FALSE}
MODULE_NAME <- "42_deconvolution_and_integration"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
set.seed(42)
```

## 1. Deconvolution is constrained regression, eq. (42.1)-(42.2)

$$y_g=\sum_k S_{gk}w_k+\varepsilon_g,\qquad w_k\ge 0,\ \sum_k w_k=1
  \qquad (42.1)-(42.2)$$

```{r deconvolve}
header("1. Recovering cell-type proportions from a mixture (42.1)-(42.2)")
```

Cell-type reference profiles. `similarity` makes two types alike, which is
what happens with, say, CD4 and CD8 T cells.

```{r}
make_signatures <- function(n_genes = 600, n_types = 5, similarity = 0,
                            seed = 0) {
  set.seed(seed)
  S <- matrix(exp(rnorm(n_genes * n_types, 3.0, 1.0)), n_genes, n_types)
  if (similarity > 0) {
    mid <- (S[, 1] + S[, 2]) / 2
    S[, 1] <- (1 - similarity) * S[, 1] + similarity * mid
    S[, 2] <- (1 - similarity) * S[, 2] + similarity * mid
  }
  S
}
```

Eq. (42.2): non-negative least squares, solved with a bound-constrained
optimiser so the module needs no extra packages. Then rescale onto the
simplex.

```{r}
deconvolve <- function(y, S) {
  StS <- crossprod(S); Sty <- as.vector(crossprod(S, y))
  fn <- function(w) sum((S %*% w - y)^2)
  gr <- function(w) 2 * (as.vector(StS %*% w) - Sty)
  start <- pmax(qr.coef(qr(S), y), 0)
  start[!is.finite(start)] <- 0
  w <- optim(start, fn, gr, method = "L-BFGS-B",
             lower = rep(0, ncol(S)))$par
  if (sum(w) > 0) w / sum(w) else rep(1 / ncol(S), ncol(S))
}
S <- make_signatures(seed = 1)
n_types <- ncol(S)
true_w <- c(0.40, 0.25, 0.15, 0.15, 0.05)
cat(sprintf("  %d genes, %d cell types\n\n", nrow(S), n_types))
cat(sprintf("  %-14s%s%15s\n", "noise level",
            paste(sprintf("%10s", paste0("type ", 1:n_types)), collapse = ""),
            "mean |error|"))
cat(sprintf("  %-14s%s%15s\n", "TRUTH",
            paste(sprintf("%10.3f", true_w), collapse = ""), ""))
for (noise in c(0.02, 0.10, 0.30)) {
  est <- t(replicate(200, {
    y <- as.vector(S %*% true_w) * exp(rnorm(nrow(S), 0, noise))
    deconvolve(y, S) }))
  cat(sprintf("  %-14.2f%s%15.4f\n", noise,
              paste(sprintf("%10.3f", colMeans(est)), collapse = ""),
              mean(abs(colMeans(est) - true_w))))
}
cat("\n  With a correct signature matrix, deconvolution is accurate and
  degrades gracefully with noise. Everything below is about what happens when
  one of its assumptions fails.\n")
```

## 2. Collinear cell types, eq. (42.2)

```{r collinear}
header("2. Similar cell types are collinear predictors")
cat(sprintf("  %-30s%13s%11s%14s\n", "similarity of types 1 and 2",
            "corr(S1,S2)", "SD of w1", "SD of w1+w2"))
for (sim in c(0.0, 0.5, 0.85, 0.97)) {
  S2 <- make_signatures(similarity = sim, seed = 2)
  est <- t(replicate(300, {
    y <- as.vector(S2 %*% true_w) * exp(rnorm(nrow(S2), 0, 0.10))
    deconvolve(y, S2) }))
  cat(sprintf("  %-30.2f%13.3f%11.4f%14.4f\n", sim, cor(S2[, 1], S2[, 2]),
              sd(est[, 1]), sd(est[, 1] + est[, 2])))
}
cat("\n  As the two signatures converge, the individual estimates become
  wildly unstable while their SUM stays precise. The data determines how much
  'type 1 or type 2' there is, and cannot split it.

  This is the collinearity behaviour of Module 11, and it has a practical
  consequence: report related cell types as a COMBINED fraction unless you can
  show the signatures are separable. A paper reporting a shift from CD4 to CD8
  with a constant total is often reporting numerical instability.\n")
```

## 3. The output is compositional

```{r compositional}
header("3. Proportions cannot move independently")
S3 <- make_signatures(seed = 3)
# One cell type genuinely doubles; the others are untouched in absolute terms.
abs_ctrl <- c(1000, 600, 400, 400, 150)
abs_case <- abs_ctrl; abs_case[1] <- abs_case[1] * 2
sim_prop <- function(ab) t(replicate(300, {
  w <- ab / sum(ab)
  y <- as.vector(S3 %*% w) * exp(rnorm(nrow(S3), 0, 0.08))
  deconvolve(y, S3) }))
c0 <- sim_prop(abs_ctrl); c1 <- sim_prop(abs_case)
cat("  Only cell type 1 changed, and it DOUBLED in absolute abundance.\n\n")
cat(sprintf("  %-8s%18s%18s%18s%18s\n", "type", "absolute change",
            "proportion ctrl", "proportion case", "apparent change"))
for (i in 1:n_types)
  cat(sprintf("  %-8d%18.2f%18.3f%18.3f%18.2f\n", i,
              abs_case[i] / abs_ctrl[i], mean(c0[, i]), mean(c1[, i]),
              mean(c1[, i]) / mean(c0[, i])))
cat("\n  Four cell types did not change at all and every one of them appears
  to have fallen by about a quarter. That is closure (Topic 26), not biology.

  Deconvolution output is compositional and must be analysed as such: use
  log-ratios, or anchor to an absolute measurement such as total cell count
  per gram of tissue. Testing proportions directly with a t-test per cell type
  produces exactly the correlated false positives Module 36 demonstrates.\n")
```

## 4. Batch integration, eq. (42.3)

$$x_i^{corr}=x_i-\frac{1}{|P|}\sum_{(a,b)\in P}(x_a-x_b) \qquad (42.3)$$

```{r integration}
header("4. MNN-style correction, and the metric that can be gamed (42.3)")
```

Two batches, two cell types. If shared is FALSE, type 2 exists in only one
batch, which is what stops naive alignment being safe.

```{r}
make_batches <- function(n_per = 300, n_genes = 40, batch_shift = 3.0,
                         seed = 0, shared = TRUE) {
  set.seed(seed)
  X <- list(); bb <- cc <- integer(0)
  for (b in 0:1) {
    types <- if (shared || b == 0) 0:1 else 0
    for (t in types) {
      centre <- rep(b * batch_shift, n_genes)
      centre[1:5] <- centre[1:5] + t * 4.0             # biology
      X[[length(X) + 1]] <- matrix(rnorm(n_per * n_genes, 0, 1), n_per) +
        rep(centre, each = n_per)
      bb <- c(bb, rep(b, n_per)); cc <- c(cc, rep(t, n_per))
    }
  }
  list(X = do.call(rbind, X), batch = bb, ctype = cc)
}
sqdist <- function(A, B)
  outer(rowSums(A^2), rowSums(B^2), "+") - 2 * A %*% t(B)
```

Eq. (42.3): find mutual nearest neighbours across batches and subtract the
average difference.

```{r}
mnn_correct <- function(X, batch, k = 20) {
  a <- X[batch == 0, , drop = FALSE]; b <- X[batch == 1, , drop = FALSE]
  d <- sqdist(a, b)
  nn_a <- t(apply(d, 1, order))[, 1:k, drop = FALSE]   # b-neighbours of each a
  nn_b <- t(apply(d, 2, order))[, 1:k, drop = FALSE]   # a-neighbours of each b
  vecs <- list()
  for (i in seq_len(nrow(a))) for (j in nn_a[i, ])
    if (i %in% nn_b[j, ]) vecs[[length(vecs) + 1]] <- a[i, ] - b[j, ]
  if (!length(vecs)) return(X)
  vec <- colMeans(do.call(rbind, vecs))
  out <- X
  out[batch == 1, ] <- sweep(out[batch == 1, , drop = FALSE], 2, vec, "+")
  out
}
```

Fraction of each cell's neighbours from the other batch. 0.5 is ideal.

```{r}
mixing <- function(X, batch, k = 25) {
  d <- sqdist(X, X); diag(d) <- Inf
  nn <- t(apply(d, 1, order))[, 1:k, drop = FALSE]
  mean(matrix(batch[nn], nrow(nn)) != batch)
}
```

Silhouette-like separation of the true cell types.

```{r}
bio_preserved <- function(X, ctype) {
  c0 <- X[ctype == 0, , drop = FALSE]; c1 <- X[ctype == 1, , drop = FALSE]
  between <- sqrt(sum((colMeans(c0) - colMeans(c1))^2))
  within <- (mean(apply(c0, 2, sd)) + mean(apply(c1, 2, sd))) / 2
  between / within
}
B <- make_batches(seed = 5)
X <- B$X; batch <- B$batch; ctype <- B$ctype
Xc <- mnn_correct(X, batch)
# An "integration" that destroys everything: shrink the real structure to
# almost nothing and let noise dominate.
X_over <- sweep(X, 2, colMeans(X)) * 0.02 +
  matrix(rnorm(length(X), 0, 0.5), nrow(X))
cat(sprintf("  %-28s%15s%20s\n", "state", "batch mixing",
            "biology preserved"))
for (z in list(list("before correction", X), list("MNN corrected (42.3)", Xc),
               list("aggressive over-correction", X_over)))
  cat(sprintf("  %-28s%15.3f%20.2f\n", z[[1]], mixing(z[[2]], batch),
              bio_preserved(z[[2]], ctype)))
cat("\n  Read the two columns together, never one alone. The over-corrected
  version has the BEST mixing score of the three, and it achieved that by
  deleting the biology.

  Any integration strong enough to remove all batch signal will remove biology
  correlated with batch. Report a mixing metric AND a biology-preservation
  metric, and if your design is confounded no correction can separate them
  (Module 19).\n")
```

## 5. Integration when a population exists in only one batch

```{r unshared}
header("5. Mutual nearest neighbours protects against forced alignment")
U <- make_batches(seed = 6, shared = FALSE)
Xu <- U$X; bu <- U$batch; cu <- U$ctype
Xu_c <- mnn_correct(Xu, bu)
only_in_0 <- cu == 1
cat(sprintf("  Cell type 2 appears in batch 0 only (%d cells).\n\n",
            sum(only_in_0)))
cat(sprintf("  %-26s%10s%38s\n", "", "mixing",
            "type-2 cells pulled toward batch 1"))
for (z in list(list("before", Xu), list("MNN corrected", Xu_c))) {
  D <- z[[2]]
  d_to_b1 <- sqrt(sum((colMeans(D[only_in_0, , drop = FALSE]) -
                       colMeans(D[bu == 1, , drop = FALSE]))^2))
  cat(sprintf("  %-26s%10.3f%38.2f\n", z[[1]], mixing(D, bu), d_to_b1))
}
cat("\n  The mutual criterion is what makes this safe. A cell in batch 0 with
  no counterpart in batch 1 will rarely be part of a RECIPROCAL
  nearest-neighbour pair, so it contributes little to the correction vector.

  Methods that align batches globally, by matching distributions rather than
  cells, have no such protection and will happily map a population that exists
  in one batch onto an unrelated population in the other. Check what happens
  to batch-specific populations before trusting any integration.\n")
```

## 6. Is there really zero-inflation? eq. (42.4)-(42.5)

$$P(Y=0)=\left(\frac{1/\phi}{1/\phi+\mu}\right)^{1/\phi} \qquad (42.5)$$

```{r zero-inflation}
header("6. Testing for zero-inflation instead of assuming it (42.4)-(42.5)")
nb_zero_prob <- function(mu, disp) {
  size <- 1 / disp
  (size / (size + mu))^size
}
simulate_umi <- function(n_cells = 1500, n_genes = 800, disp = 0.25,
                         pi_extra = 0, seed = 0) {
  set.seed(seed)
  mu <- exp(rnorm(n_genes, 0.5, 1.6))
  size <- 1 / disp
  lam <- matrix(rgamma(n_cells * n_genes, size,
                       rate = rep(size / mu, each = n_cells)), n_cells)
  y <- matrix(rpois(length(lam), lam), n_cells)
  if (pi_extra > 0)
    y <- y * (matrix(runif(length(y)), nrow(y)) > pi_extra)  # true dropout
  list(y = y, mu = mu)
}
cat(sprintf("  %-34s%17s%14s%10s\n", "data", "mean obs zeros", "NB predicts",
            "excess"))
for (z in list(list("plain NB (no inflation)", 0.0),
               list("NB + 20% true dropout", 0.20))) {
  U2 <- simulate_umi(pi_extra = z[[2]], seed = 17)
  y <- U2$y
  obs <- colMeans(y == 0)
  mu_hat <- colMeans(y); v <- apply(y, 2, var)
  disp_hat <- pmin(pmax((v - mu_hat) / pmax(mu_hat^2, 1e-9), 1e-3), 10)
  pred <- nb_zero_prob(mu_hat, disp_hat)
  cat(sprintf("  %-34s%17.3f%14.3f%10.3f\n", z[[1]], mean(obs), mean(pred),
              mean(obs) - mean(pred)))
}
cat("\n  When the data really is negative binomial, the observed zeros match
  the NB prediction, and fitting an extra inflation parameter would be fitting
  noise. When there IS extra dropout, the excess shows up plainly.

  This is a diagnostic anyone can run in three lines, and it is the reason the
  scRNA-seq field moved away from zero-inflated models for UMI data. The zeros
  were never unexplained; they are what a small mean produces.\n")

```
```{r zi-cost}
header("6b. What fitting an unnecessary zero-inflation costs")
cat("  Data has NO zero-inflation. Forcing a ZI parameter anyway, by
  attributing every zero beyond the POISSON expectation to dropout:\n\n")
cat(sprintf("  %-14s%10s%10s%10s%10s%12s\n", "dispersion", "Q1", "Q2", "Q3",
            "Q4", "worst gene"))
for (dsp in c(0.25, 0.80, 2.00)) {
  U3 <- simulate_umi(pi_extra = 0, disp = dsp, seed = 23)
  y <- U3$y; mu <- U3$mu
  mu_hat <- colMeans(y); obs0 <- colMeans(y == 0)
  pi_forced <- pmin(pmax(obs0 - exp(-mu_hat), 0), 0.95)
  mu_zinb <- mu_hat / (1 - pi_forced)
  qs <- quantile(mu, seq(0, 1, length.out = 5))
  rr <- sapply(1:4, function(i) {
    m <- mu >= qs[i] & mu <= qs[i + 1]; mean(mu_zinb[m] / mu[m]) })
  cat(sprintf("  %-14.2f%9.2fx%9.2fx%9.2fx%9.2fx%11.2fx\n", dsp, rr[1], rr[2],
              rr[3], rr[4], max(mu_zinb / mu)))
}
cat("  (expression quartiles, Q1 lowest; each cell is the mean inflation factor)\n")
cat("\n  Read along a row first. The inflation is worst in the MIDDLE of the
  expression range, not at the bottom. Very low-expressed genes have few
  counts under either model, so the Poisson and NB zero predictions nearly
  agree and little is misattributed. Very high-expressed genes have almost no
  zeros at all. In between, the NB produces many more zeros than a Poisson
  would, and every one of those gets blamed on dropout.

  Now read down the columns. The size of the error is set by the DISPERSION,
  because dispersion is what creates the extra zeros in the first place. At
  dispersion 2.0 the typical mid-range gene's mean is inflated by nearly half,
  and the worst gene by two thirds.

  Test first (eq. 42.5). Model dropout only if the excess is there, and expect
  the answer to depend on the protocol: plate-based and read-based assays
  behave differently from UMI ones.\n")
```

## 7. Cell-cell communication, eq. (42.6)

$$\text{score}(a\to b)=\bar{L}_a\cdot\bar{R}_b \qquad (42.6)$$

```{r communication}
header("7. Communication scores and the null they are tested against (42.6)")
set.seed(4207)
n_cc <- 2000; K <- 4
ctype_cc <- sample(0:(K - 1), n_cc, replace = TRUE)
# Ligand and receptor expression depend ONLY on cell type abundance and a
# shared technical factor. There is no communication in this simulation.
depth <- rlnorm(n_cc, 0, 0.4)
lig <- rpois(n_cc, exp(1.2 + 0.4 * ctype_cc) * depth)
rec <- rpois(n_cc, exp(1.0 + 0.3 * ctype_cc) * depth)
comm_score <- function(lig, rec, ctype, a, b)
  mean(lig[ctype == a]) * mean(rec[ctype == b])          # eq. (42.6)
pairs <- expand.grid(a = 0:(K - 1), b = 0:(K - 1))
obs <- mapply(function(a, b) comm_score(lig, rec, ctype_cc, a, b),
              pairs$a, pairs$b)
null <- replicate(400, {
  perm <- sample(ctype_cc)
  mapply(function(a, b) comm_score(lig, rec, perm, a, b), pairs$a, pairs$b) })
sig <- (1 + rowSums(null >= obs)) / 401
cat(sprintf("  %dx%d = %d sender-receiver pairs, NO communication simulated\n",
            K, K, nrow(pairs)))
cat(sprintf("  pairs called significant by label permutation: %d\n",
            sum(sig < 0.05)))
cat(sprintf("\n  %-12s%12s%16s\n", "pair", "score", "permutation p"))
for (i in order(-obs)[1:5])
  cat(sprintf("  %-12s%12.1f%16.4f\n",
              sprintf("(%d, %d)", pairs$a[i], pairs$b[i]), obs[i], sig[i]))
cat("\n  There is no signalling anywhere in this simulation. Ligand and
  receptor levels rise with cell-type index and with sequencing depth, and
  that is enough to produce significant 'interactions'.

  Two reasons. The score in (42.6) is a product of expression magnitudes, so
  abundant ligands and abundant cell types dominate it. And permuting cell
  labels asks 'is this pair unusual for these cell types', which is not the
  same question as 'do these cells communicate'.

  Treat communication output as a ranked hypothesis list. A claim of
  signalling needs perturbation evidence, not a permutation p-value.\n")
```

## 8. Figure

```{r figure, fig.width = 13, fig.height = 4.5}
png(file.path(OUT, "deconvolution.png"), width = 1300, height = 450)
par(mfrow = c(1, 3), mar = c(4.5, 4.5, 3, 1))
sims <- c(0, 0.3, 0.6, 0.85, 0.95, 0.99)
sd1 <- sdsum <- numeric(length(sims))
for (i in seq_along(sims)) {
  S2 <- make_signatures(similarity = sims[i], seed = 2)
  est <- t(replicate(150, deconvolve(as.vector(S2 %*% true_w) *
                                       exp(rnorm(nrow(S2), 0, 0.10)), S2)))
  sd1[i] <- sd(est[, 1]); sdsum[i] <- sd(est[, 1] + est[, 2])
}
plot(sims, sd1, type = "b", pch = 16, col = "firebrick", lwd = 2,
     ylim = c(0, max(sd1)), xlab = "similarity of the two signatures",
     ylab = "SD of the estimate", main = "Collinear types cannot be split")
lines(sims, sdsum, type = "b", pch = 17, col = "steelblue", lwd = 2)
legend("topleft", c("w1 alone", "w1 + w2"), pch = c(16, 17), lwd = 2,
       bty = "n", col = c("firebrick", "steelblue"))
plot(mixing(X, batch), bio_preserved(X, ctype), pch = 16, cex = 2,
     col = "grey40", xlim = c(0, 0.6), ylim = c(0, 12),
     xlab = "batch mixing", ylab = "biology preserved",
     main = "Both metrics, or neither")
points(mixing(Xc, batch), bio_preserved(Xc, ctype), pch = 16, cex = 2,
       col = "steelblue")
points(mixing(X_over, batch), bio_preserved(X_over, ctype), pch = 16, cex = 2,
       col = "firebrick")
text(c(mixing(X, batch), mixing(Xc, batch), mixing(X_over, batch)),
     c(bio_preserved(X, ctype), bio_preserved(Xc, ctype),
       bio_preserved(X_over, ctype)),
     c("before", "MNN", "over-corrected"), pos = c(4, 2, 2), cex = 0.8)
U4 <- simulate_umi(pi_extra = 0, seed = 17)
mh <- colMeans(U4$y); ob <- colMeans(U4$y == 0)
vv <- apply(U4$y, 2, var)
dh <- pmin(pmax((vv - mh) / pmax(mh^2, 1e-9), 1e-3), 10)
plot(log10(mh), ob, pch = 16, cex = 0.4, col = adjustcolor("grey40", 0.5),
     xlab = "log10 mean expression", ylab = "fraction of zeros",
     main = "Zeros are what a small mean gives")
o <- order(mh)
lines(log10(mh[o]), nb_zero_prob(mh, dh)[o], col = "firebrick", lwd = 2)
legend("topright", c("observed", "NB prediction (42.5)"), pch = c(16, NA),
       lwd = c(NA, 2), bty = "n", col = c("grey40", "firebrick"))
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "deconvolution.png"), "\n")
```

# PROBLEMS

### Problem 1: A signature from the wrong tissue

Deconvolve using a signature matrix that is systematically wrong, as
happens when the reference came from blood and the sample is tumour.

```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# S_true <- make_signatures(seed = 11)
# cat(sprintf("  %-34s%s%13s\n", "reference used",
#             paste(sprintf("%9s", paste0("type ", 1:n_types)), collapse = ""),
#             "mean |err|"))
# cat(sprintf("  %-34s%s%13s\n", "TRUTH",
#             paste(sprintf("%9.3f", true_w), collapse = ""), ""))
# for (z in list(list("correct reference", 0.0),
#                list("mildly wrong platform", 0.25),
#                list("wrong tissue", 0.8))) {
#   set.seed(12)
#   S_used <- S_true * exp(matrix(rnorm(length(S_true), 0, z[[2]]),
#                                 nrow(S_true)))
#   est <- t(replicate(200, deconvolve(as.vector(S_true %*% true_w) *
#                        exp(rnorm(nrow(S_true), 0, 0.08)), S_used)))
#   cat(sprintf("  %-34s%s%13.4f\n", z[[1]],
#               paste(sprintf("%9.3f", colMeans(est)), collapse = ""),
#               mean(abs(colMeans(est) - true_w))))
# }
#
# ## The estimates degrade steadily and, importantly, they degrade SILENTLY:
# ## the fit always returns proportions that sum to one and always looks like
# ## a plausible answer. There is no residual diagnostic in the standard
# ## output that screams "wrong reference".
# ##
# ## The check worth running is the residual: compare the observed bulk
# ## profile against S %*% w_hat. A reference that does not fit leaves
# ## structured, not random, residuals. Always report that fit, and always
# ## match the reference platform and tissue.
```

### Problem 2: Does correction help or harm when the design is confounded?

Compare integration on a balanced design and a fully confounded one.

```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# confounded_batches <- function(confound, seed = 0, n_per = 300,
#                                n_genes = 40) {
#   set.seed(seed)
#   X <- list(); bb <- cc <- integer(0)
#   for (b in 0:1) for (t in 0:1) {
#     # When confounded, batch 0 is almost all type 0 and vice versa.
#     n <- if (confound) round(n_per * if (b == t) 0.9 else 0.1) else n_per
#     if (n == 0) next
#     centre <- rep(b * 3.0, n_genes); centre[1:5] <- centre[1:5] + t * 4.0
#     X[[length(X) + 1]] <- matrix(rnorm(n * n_genes), n) +
#       rep(centre, each = n)
#     bb <- c(bb, rep(b, n)); cc <- c(cc, rep(t, n))
#   }
#   list(X = do.call(rbind, X), batch = bb, ctype = cc)
# }
# cat(sprintf("  %-22s%-18s%10s%15s\n", "design", "state", "mixing",
#             "biology kept"))
# for (z in list(list("balanced", FALSE), list("confounded", TRUE))) {
#   Cb <- confounded_batches(z[[2]], seed = 21)
#   Cbc <- mnn_correct(Cb$X, Cb$batch)
#   for (w in list(list("before", Cb$X), list("after MNN", Cbc)))
#     cat(sprintf("  %-22s%-18s%10.3f%15.2f\n", z[[1]], w[[1]],
#                 mixing(w[[2]], Cb$batch), bio_preserved(w[[2]], Cb$ctype)))
# }
#
# ## On the balanced design, correction improves mixing and leaves the biology
# ## intact. On the confounded design, correction improves mixing by
# ## destroying the biological separation, because the batch difference and
# ## the biological difference are the same direction in the data.
# ##
# ## The mixing metric improves in BOTH cases. If you reported only mixing,
# ## the confounded design would look like the bigger success. This is why
# ## Module 19 insists that confounding is a design failure with no analytical
# ## remedy, and why the two metrics must always be read as a pair.
```

### Problem 3: When does zero-inflation actually appear?

Compare a UMI-like protocol with a read-based one where amplification adds
genuine dropout, and check whether the diagnostic distinguishes them.

```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# cat(sprintf("  %-34s%12s%13s%9s%16s\n", "protocol", "obs zeros",
#             "NB predicts", "excess", "verdict"))
# for (z in list(list("UMI, well behaved", 0.00, 0.25),
#                list("UMI, high dispersion", 0.00, 0.80),
#                list("read-based, real dropout", 0.25, 0.25),
#                list("severe dropout", 0.50, 0.25))) {
#   Uz <- simulate_umi(pi_extra = z[[2]], disp = z[[3]], seed = 41)
#   obs <- mean(colMeans(Uz$y == 0))
#   mh2 <- colMeans(Uz$y); vv2 <- apply(Uz$y, 2, var)
#   dh2 <- pmin(pmax((vv2 - mh2) / pmax(mh2^2, 1e-9), 1e-3), 20)
#   pred <- mean(nb_zero_prob(mh2, dh2))
#   exc <- obs - pred
#   cat(sprintf("  %-34s%12.3f%13.3f%9.3f%16s\n", z[[1]], obs, pred, exc,
#               if (exc > 0.02) "inflated" else "no inflation"))
# }
#
# ## The diagnostic separates the cases correctly, and note the second row in
# ## particular: high dispersion produces MANY zeros, and the NB still
# ## predicts them. Lots of zeros is not evidence of zero-inflation. Only
# ## zeros in excess of what the fitted mean-variance relationship predicts
# ## are.
# ##
# ## That distinction is the whole argument. The scRNA-seq literature spent
# ## years treating "mostly zeros" as self-evident proof of a dropout process,
# ## when the negative binomial had already accounted for them. Estimate the
# ## dispersion first, then ask what is left over.
```

## What to take away

1. Deconvolution is constrained regression (42.1)-(42.2) and inherits
   collinearity: report similar cell types as a combined fraction.
2. Its output is **compositional**. One type rising forces the others down.
3. Integration must be judged on **mixing and biology together**.
   Over-correction wins on mixing alone.
4. Mutual nearest neighbours protects batch-specific populations; global
   distribution matching does not.
5. **Test** for zero-inflation with eq. (42.5) rather than assuming it.
   Forcing one inflates low-expression means.
6. Communication scores (42.6) are driven by expression magnitude and
   abundance. They rank hypotheses; they do not establish signalling.

**Next:** `43_advanced_statistical_genetics.R`
