---
title: "Applied 46 - Model interpretation and pretrained models"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 46, equations (46.1)-(46.3)
**Core modules used:** 19, 21, 28, 31

## The question

The model predicts well. What is it using, can you believe the explanation,
and what changes when the model was pretrained on data you did not control?

## The claim to keep in mind

An explanation is a statement about the **model**, not about biology. A
feature can be important to a model because it proxies batch.

```{r setup, message = FALSE}
suppressPackageStartupMessages(library(rpart))
MODULE_NAME <- "46_interpretation_and_foundation_models"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
set.seed(46)
```

Python's module uses scikit-learn's random forest. R's base image has no
such package, so we build the same thing directly: bag regression trees,
and let each tree see a random subset of `mtry` features. That subset size
turns out to matter in section 2b.

```{r}
rf_fit <- function(X, y, ntree = 120, mtry = max(1, floor(ncol(X) / 3)),
                   minsplit = 10) {
  n <- nrow(X)
  lapply(seq_len(ntree), function(b) {
    idx <- sample(n, n, replace = TRUE)
    cols <- sample(ncol(X), mtry)
    df <- data.frame(y = y[idx], X[idx, cols, drop = FALSE])
    list(tree = rpart(y ~ ., data = df, method = "anova",
                      control = rpart.control(minsplit = minsplit, cp = 0.001)),
         cols = cols)
  })
}
rf_pred <- function(fit, X) {
  rowMeans(sapply(fit, function(b) {
    df <- data.frame(X[, b$cols, drop = FALSE])
    predict(b$tree, df)
  }))
}
mse <- function(a, b) mean((a - b)^2)
```

## 1. Permutation importance, eq. (46.1)

$$I_j=\mathbb{E}\big[L(y,f(X^{\pi_j}))\big]-L(y,f(X)) \qquad (46.1)$$

```{r perm-importance}
header("1. Permutation importance on independent features (46.1)")
```

Eq. (46.1): the loss increase when feature j is shuffled.

```{r}
perm_importance <- function(predict_fn, X, y, metric = mse, n_rep = 8,
                            seed = 0) {
  set.seed(seed)
  base <- metric(y, predict_fn(X))
  sapply(seq_len(ncol(X)), function(j) {
    v <- replicate(n_rep, {
      Xp <- X; Xp[, j] <- sample(Xp[, j]); metric(y, predict_fn(Xp))
    })
    mean(v) - base
  })
}
n <- 1200; p <- 8
X <- matrix(rnorm(n * p), n)
beta <- c(2.0, 1.0, 0.5, 0, 0, 0, 0, 0)
y <- as.vector(X %*% beta) + rnorm(n)
# mtry = p: every split may consider every feature. Section 2b shows why
# this choice is not innocent once the features are correlated.
fit <- rf_fit(X, y, mtry = p)
pi_ <- perm_importance(function(Z) rf_pred(fit, Z), X, y)
cat(sprintf("  %-10s%11s%18s\n", "feature", "true beta", "perm importance"))
for (j in 1:p)
  cat(sprintf("  %-10d%11.1f%18.4f\n", j, beta[j], pi_[j]))
cat("\n  With independent features, permutation importance recovers the truth:
  the ranking matches the coefficients and the null features sit near zero.

  Everything from here is about what happens when that independence fails,
  which in omics data it always does.\n")
```

## 2. Correlated features split the credit

```{r correlated}
header("2. Correlated features split the credit")
cat("  Case A: x1 CAUSES y, and x2 is a noisy copy of x1 with no effect of its own.\n\n")
cat(sprintf("  %-14s%13s%13s%11s\n", "corr(x1,x2)", "x1 (causal)",
            "x2 (a copy)", "x2 share"))
for (rho in c(0.0, 0.5, 0.8, 0.95, 0.99)) {
  set.seed(7)
  x1 <- rnorm(n)
  x2 <- rho * x1 + sqrt(max(1 - rho^2, 1e-9)) * rnorm(n)
  Xc <- cbind(x1, x2, matrix(rnorm(n * 3), n))
  yc <- 2.0 * x1 + rnorm(n)
  fc <- rf_fit(Xc, yc, mtry = 2)
  p_ <- perm_importance(function(Z) rf_pred(fc, Z), Xc, yc, seed = 2)
  cat(sprintf("  %-14.2f%13.3f%13.3f%10.1f%%\n", rho, p_[1], p_[2],
              100 * p_[2] / (p_[1] + p_[2])))
}
cat("\n  Here the leak is mild. Even at correlation 0.99, x1 keeps most of the
  credit, because it is the genuine cause and therefore splits the data
  slightly better than its copy does.

  Case B is the one that matters in omics. Neither measured feature is the
  cause: both are noisy readouts of a latent driver z, the way two
  co-expressed genes are both readouts of one transcription factor.\n\n")
cat(sprintf("  %-14s%13s%9s%9s%11s\n", "proxy noise", "corr(x1,x2)", "x1",
            "x2", "x1 share"))
for (tau in c(2.0, 1.0, 0.5, 0.2, 0.05)) {
  set.seed(11)
  z <- rnorm(n)
  x1 <- z + rnorm(n, 0, tau); x2 <- z + rnorm(n, 0, tau)
  Xc <- cbind(x1, x2, matrix(rnorm(n * 3), n))
  yc <- 2.0 * z + rnorm(n)
  fc <- rf_fit(Xc, yc, mtry = 2)
  p_ <- perm_importance(function(Z) rf_pred(fc, Z), Xc, yc, seed = 2)
  cat(sprintf("  %-14.2f%13.2f%9.3f%9.3f%10.1f%%\n", tau, cor(x1, x2), p_[1],
              p_[2], 100 * p_[1] / (p_[1] + p_[2])))
}
cat("\n  The credit splits close to half at every noise level, and each
  feature alone looks about half as important as the pair really is.

  Two consequences. A gene downstream of a real driver can look unimportant
  because a co-expressed gene absorbs half its credit. And \"gene X was the top
  feature\" is close to a coin flip between the two: Problem 2 measures how
  close.

  Report importance for CORRELATED GROUPS, not individual features, or cluster
  the features first and permute whole clusters.\n")

```
```{r hyperparameter}
header("2b. The split also depends on a hyperparameter, not on biology")
cat(sprintf("  %-16s%13s%13s%11s\n", "mtry", "x1 (causal)", "x2 (a copy)",
            "x2 share"))
for (mt in c(5, 2)) {
  set.seed(7)
  x1 <- rnorm(n)
  x2 <- 0.95 * x1 + sqrt(1 - 0.95^2) * rnorm(n)
  Xc <- cbind(x1, x2, matrix(rnorm(n * 3), n))
  yc <- 2.0 * x1 + rnorm(n)
  fc <- rf_fit(Xc, yc, mtry = mt)
  p_ <- perm_importance(function(Z) rf_pred(fc, Z), Xc, yc, seed = 2)
  cat(sprintf("  %-16s%13.3f%13.3f%10.1f%%\n",
              if (mt == 5) "5 (all features)" else "2 (a subset)", p_[1],
              p_[2], 100 * p_[2] / (p_[1] + p_[2])))
}
cat("\n  Same data, same correlation, same true effect. When every split
  considers all features the forest keeps choosing x1, so x2 looks
  unimportant. When each split sees a random subset, x2 gets used whenever x1
  is absent and its importance rises.

  The share of credit given to a feature with no effect is therefore partly a
  function of a tuning parameter. Always state which model and which settings
  produced an importance ranking.\n")
```

## 3. Permutation evaluates the model off the data manifold

```{r manifold}
header("3. Shuffling creates samples that could never exist")
set.seed(4601)
rho <- 0.95
z <- rnorm(n)
x1 <- z; x2 <- rho * z + sqrt(1 - rho^2) * rnorm(n)
Xm <- cbind(x1, x2)
Xperm <- Xm; Xperm[, 2] <- sample(Xperm[, 2])
cat(sprintf("  corr(x1, x2) in the real data      : %.3f\n",
            cor(Xm[, 1], Xm[, 2])))
cat(sprintf("  corr(x1, x2) after permuting x2    : %.3f\n",
            cor(Xperm[, 1], Xperm[, 2])))
d_real <- abs(Xm[, 1] - Xm[, 2]); d_perm <- abs(Xperm[, 1] - Xperm[, 2])
cat(sprintf("\n  |x1 - x2| in real data  : median %.2f, 99th pct %.2f\n",
            median(d_real), quantile(d_real, 0.99)))
cat(sprintf("  |x1 - x2| after permuting: median %.2f, 99th pct %.2f\n",
            median(d_perm), quantile(d_perm, 0.99)))
cat(sprintf("\n  fraction of permuted rows outside the real data's 99th percentile: %.1f%%\n",
            100 * mean(d_perm > quantile(d_real, 0.99))))
cat("\n  Permuting breaks the correlation, so a large share of the rows the
  model is asked to score are combinations that never occur in nature: a high
  x1 with a low x2 when the two are 95% correlated.

  The importance you get is therefore partly a measurement of the model's
  behaviour in a region where it was never trained and where its output is
  arbitrary. Conditional or grouped permutation schemes exist for this
  reason.\n")
```

## 4. Shapley values, eq. (46.2)-(46.3)

$$\phi_j=\sum_{S\subseteq F\setminus\{j\}}
  \frac{|S|!(|F|-|S|-1)!}{|F|!}\big[v(S\cup\{j\})-v(S)\big] \qquad (46.2)$$
$$f(x)=v(\emptyset)+\sum_j \phi_j \qquad (46.3)$$

```{r shapley}
header("4. Exact Shapley values on a small model (46.2)-(46.3)")
```

Eq. (46.2) by full enumeration over orderings. Feasible only for a handful
of features, which is why real implementations approximate.

```{r}
shapley_exact <- function(predict_fn, x, background) {
  k <- length(x)
  perms <- as.matrix(expand.grid(rep(list(seq_len(k)), k)))
  perms <- perms[apply(perms, 1, function(r) length(unique(r)) == k), ,
                 drop = FALSE]
  phi <- numeric(k)
  for (i in seq_len(nrow(perms))) {
    present <- integer(0)
    prev <- predict_fn(background)
    for (j in perms[i, ]) {
      present <- c(present, j)
      xx <- background; xx[, present] <- x[present]
      cur <- predict_fn(xx)
      phi[j] <- phi[j] + cur - prev
      prev <- cur
    }
  }
  phi / nrow(perms)
}
set.seed(4602)
Xs <- matrix(rnorm(800 * 4), 800)
colnames(Xs) <- paste0("V", 1:4)
ys <- 1.5 * Xs[, 1] - 1.0 * Xs[, 2] + 0.0 * Xs[, 3] + 0.5 * Xs[, 4] +
  rnorm(800, 0, 0.5)
ridge <- lm(ys ~ Xs)
pred_lin <- function(A) {
  A <- matrix(A, ncol = 4); colnames(A) <- colnames(Xs)
  as.vector(cbind(1, A) %*% coef(ridge))
}
bg <- matrix(colMeans(Xs), nrow = 1)
target <- Xs[1, ]
phi <- shapley_exact(pred_lin, target, bg)
cat(sprintf("  %-10s%13s%10s%14s\n", "feature", "coefficient", "x value",
            "Shapley phi"))
for (j in 1:4)
  cat(sprintf("  %-10d%13.3f%10.3f%14.4f\n", j, coef(ridge)[j + 1],
              target[j], phi[j]))
cat("\n  additivity check, eq. (46.3):\n")
cat(sprintf("    baseline + sum(phi) = %.4f\n", pred_lin(bg) + sum(phi)))
cat(sprintf("    model prediction    = %.4f\n",
            pred_lin(matrix(target, nrow = 1))))
cat("\n  For a linear model the Shapley value of feature j is simply
  coefficient x (value minus baseline), and the additivity property (46.3)
  holds exactly.

  That transparency is why Shapley values are attractive. It is also why they
  are over-trusted: additivity is a mathematical property of the attribution,
  not evidence that the model captured a real mechanism.\n")
```

## 5. An explanation can be entirely about batch

```{r batch-proxy}
header("5. The most important feature is a batch proxy")
set.seed(21)
n_s <- 600; n_g <- 40
y_out <- sample(0:1, n_s, replace = TRUE)
# A confounded design: 85% of cases were processed in batch 1, 85% of
# controls in batch 0. Nobody planned it; it is how the samples arrived.
FRAC <- 0.85
batch <- as.numeric(runif(n_s) < ifelse(y_out == 1, FRAC, 1 - FRAC))
bio <- rnorm(n_s) + 1.0 * y_out              # a genuinely causal gene
G <- matrix(rnorm(n_s * n_g), n_s)
G[, 1] <- bio
G[, 2] <- 3.0 * batch + rnorm(n_s, 0, 0.3)   # a purely batch-driven gene
auc <- function(lab, score) {
  r <- rank(score)
  n1 <- sum(lab == 1); n0 <- sum(lab == 0)
  (sum(r[lab == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
```

Importance is computed on HELD-OUT data. A bagged ensemble interpolates its
training set, so its in-sample AUC is 1.000 and permuting any feature leaves
it at 1.000: in-sample permutation importance measures nothing here.

```{r}
set.seed(99)
tr <- sample(n_s, n_s * 0.6); te <- setdiff(seq_len(n_s), tr)
fit2 <- rf_fit(G[tr, ], y_out[tr], ntree = 150, mtry = 7)
pi2 <- perm_importance(function(Z) rf_pred(fit2, Z), G[te, ], y_out[te],
                       metric = function(a, b) -auc(a, b), n_rep = 6, seed = 7)
top <- order(pi2, decreasing = TRUE)[1:4]
cat(sprintf("  held-out AUC of the model: %.3f\n",
            auc(y_out[te], rf_pred(fit2, G[te, ]))))
cat(sprintf("  confounding: corr(batch, outcome) = %.2f\n\n",
            cor(batch, y_out)))
cat(sprintf("  %-7s%-10s%13s%26s\n", "rank", "feature", "importance",
            "what it really is"))
labels <- c("real biology", "batch proxy")
for (i in seq_along(top))
  cat(sprintf("  %-7d%-10d%13.4f%26s\n", i, top[i], pi2[top[i]],
              if (top[i] <= 2) labels[top[i]] else "noise"))
cat(sprintf("\n  gene 2 vs batch   : r = %.3f\n", cor(G[, 2], batch)))
cat(sprintf("  gene 2 vs outcome : r = %.3f\n", cor(G[, 2], y_out)))
cat("\n  The batch-driven gene is the TOP feature, ahead of the one gene that
  actually causes the outcome. Its explanation is stable, reproducible, and
  has a plausible story attached to it as soon as you look up what the gene
  does.

  Nothing inside the importance calculation can detect this. The defences are
  all outside the model: check top features against technical metadata
  (Module 19), design so batch is not confounded with the outcome (Module 01),
  and validate any claimed mechanism experimentally.\n")
```

## 6. Evaluating a pretrained model

```{r leakage}
header("6. Pretraining breaks the usual meaning of a test split")
set.seed(4603)
n_pat <- 24; cells_per <- 60
pid <- rep(seq_len(n_pat), each = cells_per)
pat_effect <- rnorm(n_pat)
label <- as.numeric(pat_effect > 0)[pid]
sig <- pat_effect[pid] + rnorm(length(pid), 0, 0.8)
noise <- matrix(rnorm(length(pid) * 5), length(pid))
Xp <- cbind(sig, noise)
# A "pretrained embedding" that saw every patient, including the test ones.
emb_leaky <- cbind(pat_effect[pid] + rnorm(length(pid), 0, 0.15), noise)
# An embedding built without the held-out patients would not carry this.
emb_clean <- cbind(sig, noise)
```

Grouped cross-validation: whole patients are held out, never single cells.

```{r}
grouped_auc <- function(Xd, seed = 0, K = 4) {
  set.seed(seed)
  fold_of_pat <- sample(rep_len(seq_len(K), n_pat))
  pred <- rep(NA_real_, length(label))
  for (k in seq_len(K)) {
    te <- fold_of_pat[pid] == k
    df <- data.frame(y = label[!te], Xd[!te, , drop = FALSE])
    m <- suppressWarnings(glm(y ~ ., binomial(), df))
    pred[te] <- predict(m, data.frame(Xd[te, , drop = FALSE]),
                        type = "response")
  }
  ok <- !is.na(pred)
  auc(label[ok], pred[ok])
}
cat(sprintf("  %-46s%16s\n", "representation", "grouped CV AUC"))
cat(sprintf("  %-46s%16.3f\n", "raw features", grouped_auc(Xp)))
cat(sprintf("  %-46s%16.3f\n", "embedding built WITHOUT held-out patients",
            grouped_auc(emb_clean)))
cat(sprintf("  %-46s%16.3f\n", "embedding that saw every patient (leaky)",
            grouped_auc(emb_leaky)))
cat("\n  The clean embedding is deliberately identical to the raw features
  here. An embedding built without the held-out patients cannot carry
  information about them beyond what the features already contain, so its row
  SHOULD match. That is the baseline the leaky row is being compared against.

  The leaky embedding wins, and the cross-validation is grouped correctly by
  patient. The split is not the problem: the REPRESENTATION already encodes
  information about the held-out patients, so no downstream split can undo it.

  This is the structural issue when evaluating pretrained models. If your
  evaluation cohort was part of the pretraining corpus, your held-out AUC is
  not held out. The questions to ask about any foundation model result:

    1. Was the evaluation data in the pretraining corpus?
    2. Does it beat a simple baseline on the SAME splits?
    3. Does the embedding separate batch, or biology?\n")

```
```{r baseline}
header("6b. Always report the simple baseline")
cat(sprintf("  %-46s%16.3f\n", "logistic regression on raw features",
            grouped_auc(Xp, seed = 1)))
# Pseudobulk baseline: one row per patient.
pb <- t(sapply(seq_len(n_pat), function(i) colMeans(Xp[pid == i, ])))
pl <- sapply(seq_len(n_pat), function(i) label[pid == i][1])
aucs <- sapply(0:4, function(s) {
  set.seed(s)
  fold <- sample(rep_len(1:4, n_pat))
  pr <- rep(NA_real_, n_pat)
  for (k in 1:4) {
    te <- fold == k
    m <- suppressWarnings(glm(y ~ ., binomial(),
                              data.frame(y = pl[!te], pb[!te, ])))
    pr[te] <- predict(m, data.frame(pb[te, , drop = FALSE]),
                      type = "response")
  }
  auc(pl, pr)
})
cat(sprintf("  %-46s%16.3f\n", "pseudobulk, one row per patient", mean(aucs)))
cat("\n  The pseudobulk baseline is competitive with every per-cell model
  here, because the label is a property of the PATIENT and averaging cells
  removes per-cell noise without discarding anything relevant. Running the
  comparison per cell instead of per patient also inflates the apparent sample
  size from 24 to 1440 (Module 01).

  A large model that does not beat a logistic regression on pseudobulk has not
  been shown to be useful. Report the baseline on the same splits, every
  time.\n")
```

## 7. Figure

```{r figure, fig.width = 13, fig.height = 4.5}
png(file.path(OUT, "interpretation.png"), width = 1300, height = 450)
par(mfrow = c(1, 3), mar = c(4.5, 4.5, 3, 1))
grid <- c(0.2, 0.5, 0.8, 0.95, 0.99)
share_copy <- share_proxy <- numeric(length(grid))
for (i in seq_along(grid)) {
  rr <- grid[i]
  set.seed(7)
  a <- rnorm(n); b <- rr * a + sqrt(max(1 - rr^2, 1e-9)) * rnorm(n)
  Xc <- cbind(a, b, matrix(rnorm(n * 3), n)); yc <- 2.0 * a + rnorm(n)
  fc <- rf_fit(Xc, yc, ntree = 80, mtry = 2)
  pp <- perm_importance(function(Z) rf_pred(fc, Z), Xc, yc, n_rep = 5, seed = 3)
  share_copy[i] <- pp[2] / (pp[1] + pp[2])
  tau <- sqrt(max(1 / rr - 1, 1e-9))        # corr(x1,x2) = 1/(1+tau^2)
  set.seed(11)
  zz <- rnorm(n); a <- zz + rnorm(n, 0, tau); b <- zz + rnorm(n, 0, tau)
  Xc <- cbind(a, b, matrix(rnorm(n * 3), n)); yc <- 2.0 * zz + rnorm(n)
  fc <- rf_fit(Xc, yc, ntree = 80, mtry = 2)
  pp <- perm_importance(function(Z) rf_pred(fc, Z), Xc, yc, n_rep = 5, seed = 3)
  share_proxy[i] <- pp[2] / (pp[1] + pp[2])
}
plot(grid, share_copy, type = "b", pch = 16, col = "steelblue", lwd = 2,
     ylim = c(0, 0.7), xlab = "correlation between the two features",
     ylab = "share of credit given to feature B",
     main = "Where the credit goes")
lines(grid, share_proxy, type = "b", pch = 17, col = "firebrick", lwd = 2)
abline(h = 0.5, lty = 2)
legend("topleft", c("B is a copy of the cause", "both are proxies"),
       pch = c(16, 17), lwd = 2, bty = "n", col = c("steelblue", "firebrick"))
o <- order(pi2, decreasing = TRUE)[1:8]
barplot(rev(pi2[o]), horiz = TRUE, names.arg = rev(paste("gene", o)),
        las = 1, cex.names = 0.7, xlab = "importance",
        col = rev(ifelse(o == 2, "firebrick",
                         ifelse(o == 1, "steelblue", "grey60"))),
        main = "Top feature is a pure batch proxy")
barplot(c(grouped_auc(Xp), grouped_auc(emb_clean), grouped_auc(emb_leaky)),
        names.arg = c("raw", "clean\nembedding", "leaky\nembedding"),
        col = c("grey60", "steelblue", "firebrick"), ylim = c(0.4, 1.0),
        ylab = "grouped CV AUC", xpd = FALSE,
        main = "Pretraining leakage survives a correct split")
abline(h = 0.5, lty = 2)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "interpretation.png"), "\n")
```

# PROBLEMS

### Problem 1: Do the importance methods agree?

Compare permutation importance against a linear model's coefficients on the
same correlated data.

```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# set.seed(4610)
# z <- rnorm(n)
# Xa <- cbind(z, 0.9 * z + 0.44 * rnorm(n), matrix(rnorm(n * 3), n))
# ya <- 2.0 * Xa[, 1] + rnorm(n)
# fa <- rf_fit(Xa, ya, mtry = 2)
# ri <- lm(ya ~ Xa)
# pim <- perm_importance(function(Z) rf_pred(fa, Z), Xa, ya, seed = 8)
# cat(sprintf("  %-10s%8s%14s%13s\n", "feature", "truth", "permutation",
#             "linear coef"))
# for (j in 1:5)
#   cat(sprintf("  %-10d%8.1f%14.4f%13.3f\n", j, if (j == 1) 2.0 else 0.0,
#               pim[j], coef(ri)[j + 1]))
# cat(sprintf("\n  rank correlation permutation vs |linear coef|: %.3f\n",
#             cor(pim, abs(coef(ri)[-1]), method = "spearman")))
#
# ## The two methods rank the correlated pair differently, and neither
# ## recovers "feature 1 matters, feature 2 does not", because the data cannot
# ## distinguish them. The linear model splits the coefficient between the
# ## pair; the forest splits the importance.
# ##
# ## The practical rule: state which method you used, because the answer
# ## depends on it, and never present a single importance ranking as though it
# ## were a property of the biology. Where methods disagree, that disagreement
# ## is the finding.
```

### Problem 2: How stable is the top feature?

Resample the data and record how often the same feature comes top.

```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# top_counts <- function(kind, par, reps = 25, m = 600) {
#   tops <- integer(reps)
#   for (b in seq_len(reps)) {
#     set.seed(300 + b)
#     zz <- rnorm(m)
#     if (kind == "copy") {          # x2 is a copy of the causal x1
#       a <- zz; bb <- par * zz + sqrt(max(1 - par^2, 1e-9)) * rnorm(m)
#       tgt <- a
#     } else {                       # both are proxies of a latent cause
#       a <- zz + rnorm(m, 0, par); bb <- zz + rnorm(m, 0, par); tgt <- zz
#     }
#     xb <- cbind(a, bb, matrix(rnorm(m * 3), m))
#     yb <- 2.0 * tgt + rnorm(m)
#     mb <- rf_fit(xb, yb, ntree = 60, mtry = 2)
#     pb_ <- perm_importance(function(Z) rf_pred(mb, Z), xb, yb, n_rep = 3,
#                            seed = b)
#     tops[b] <- which.max(pb_)
#   }
#   c(mean(tops == 1), mean(tops == 2), mean(tops > 2))
# }
# cat(sprintf("  %-40s%9s%9s%9s\n", "setting", "A top", "B top", "other"))
# for (z in list(list("copy of the cause, r = 0.00", "copy", 0.0),
#                list("copy of the cause, r = 0.95", "copy", 0.95),
#                list("two proxies, low noise (r = 0.96)", "proxy", 0.2),
#                list("two proxies, high noise (r = 0.51)", "proxy", 1.0))) {
#   r_ <- top_counts(z[[2]], z[[3]])
#   cat(sprintf("  %-40s%8.0f%%%8.0f%%%8.0f%%\n", z[[1]], 100 * r_[1],
#               100 * r_[2], 100 * r_[3]))
# }
#
# ## When one feature is the true cause, it comes top nearly always, even when
# ## a near-perfect copy sits beside it. When neither feature is the cause and
# ## both are proxies, which one comes top is much closer to a coin flip.
# ##
# ## So "the top predictor was gene X" is a draw from this distribution, and
# ## how wide the distribution is depends on something you cannot see: whether
# ## the feature is the driver or a readout of one.
# ##
# ## The honest output is a stability measure. Resample, refit, and report how
# ## often each feature appears in the top k. It costs one loop and is almost
# ## never done.
```

### Problem 3: Detect pretraining leakage without seeing the corpus

You cannot inspect a foundation model's training data. Devise a check that
still reveals contamination.

```{r YOUR CODE HERE}

```
```{r SOLUTION (uncomment to check)}
# ## The signature of leakage: the embedding predicts patient IDENTITY far
# ## better than raw features do, even for patients it should not know.
# identity_auc <- function(Xd, seed = 0, K = 3) {
#   set.seed(seed)
#   target <- sample(n_pat, 1)
#   lab <- as.numeric(pid == target)
#   fold_of_pat <- sample(rep_len(seq_len(K), n_pat))
#   pred <- rep(NA_real_, length(lab))
#   for (k in seq_len(K)) {
#     te <- fold_of_pat[pid] == k
#     if (sum(lab[!te]) == 0) next
#     m <- suppressWarnings(glm(y ~ ., binomial(),
#                               data.frame(y = lab[!te], Xd[!te, ])))
#     pred[te] <- predict(m, data.frame(Xd[te, , drop = FALSE]),
#                         type = "response")
#   }
#   ok <- !is.na(pred)
#   if (length(unique(lab[ok])) < 2) return(NA_real_)
#   auc(lab[ok], pred[ok])
# }
# cat(sprintf("  %-44s%18s%18s\n", "representation", "predicts outcome",
#             "encodes patient"))
# for (z in list(list("raw features", Xp), list("clean embedding", emb_clean),
#                list("leaky embedding", emb_leaky))) {
#   ids <- sapply(0:5, function(s) identity_auc(z[[2]], s))
#   cat(sprintf("  %-44s%18.3f%18.3f\n", z[[1]], grouped_auc(z[[2]]),
#               mean(ids, na.rm = TRUE)))
# }
#
# ## The leaky embedding is unusual on BOTH axes: it predicts the outcome
# ## better and it carries more patient-specific information. That second
# ## number is the diagnostic, because it can be computed without any access
# ## to the pretraining corpus.
# ##
# ## Other checks in the same spirit: compare performance on cohorts collected
# ## AFTER the model's training cutoff, which cannot have been included; and
# ## test whether the embedding clusters by processing site. None of these
# ## proves contamination, but a model that is unremarkable on all of them is
# ## much easier to believe.
```

## What to take away

1. Permutation importance (46.1) works on independent features and misleads
   on correlated ones, where credit **splits** between copies.
2. Permuting evaluates the model **off the data manifold**, in a region
   where its behaviour is arbitrary.
3. Shapley values (46.2) are exactly additive (46.3). Additivity is a
   property of the attribution, not evidence of a mechanism.
4. An explanation describes the **model**. The top feature can be a pure
   batch proxy, and no importance method will tell you.
5. Report **stability**: how often does the same feature come top under
   resampling?
6. For pretrained models, a correct split does not undo contamination in
   the representation. Ask what was in the corpus, and always report a
   simple baseline on the same splits.

**This is the final applied module.** See `statsR/exercises/` for the
integrative problem sets.
