---
title: "Module 02 - Data structures, measurement scales, transformations"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 2, equations (2.1)-(2.7)

## What you will learn

1. The assay triple and why the alignment contract (2.1) must be ASSERTED.
2. How the support of a measurement determines its likelihood.
3. log+c (2.2)-(2.3), logit/M (2.4), arcsinh (2.5) -- and what each does to
   the variance.
4. Why compositional closure (2.6) forces negative correlations (2.7).

```{r setup, message = FALSE}
MODULE_NAME <- "02_data_structures_and_scales"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. The assay triple and the alignment contract, eq. (2.1)

Bioconductor's `SummarizedExperiment` exists to enforce (2.1) automatically
under subsetting and reordering. With plain matrices you must do it yourself.
A violated contract produces code that RUNS WITHOUT ERROR and is entirely
wrong -- the worst possible failure mode.

```{r assay-triple}
header("1. Asserting the alignment contract (2.1)")

new_assay_data <- function(assay, row_data, col_data) {
  ## A minimal SummarizedExperiment. The point is not the structure, it is the
  ## validate() call: alignment is checked on construction AND after subsetting.
  obj <- list(assay = assay, row_data = row_data, col_data = col_data)
  class(obj) <- "assay_data"
  validate_assay_data(obj)
}

validate_assay_data <- function(obj) {
  if (!identical(colnames(obj$assay), rownames(obj$col_data))) {
    stop("ALIGNMENT VIOLATION: assay columns != col_data rows.\n  assay: ",
         paste(head(colnames(obj$assay), 4), collapse = ", "),
         "\n  meta : ", paste(head(rownames(obj$col_data), 4), collapse = ", "))
  }
  if (!identical(rownames(obj$assay), rownames(obj$row_data))) {
    stop("ALIGNMENT VIOLATION: assay rows != row_data rows.")
  }
  obj
}

subset_samples <- function(obj, keep) {
  new_assay_data(obj$assay[, keep, drop = FALSE], obj$row_data,
                 obj$col_data[keep,, drop = FALSE])
}

set.seed(11)
genes <- sprintf("GENE%03d", 1:6)
samples <- sprintf("S%d", 1:8)
se <- new_assay_data(
  assay = matrix(rpois(48, 50), 6, 8, dimnames = list(genes, samples)),
  row_data = data.frame(chrom = rep(c("chr1", "chr2"), each = 3),
                        row.names = genes),
  col_data = data.frame(condition = rep(c("ctrl", "trt"), each = 4),
                        batch = rep(c("b1", "b1", "b2", "b2"), 2),
                        row.names = samples))
cat(sprintf("AssayData: %d features x %d samples\n",
            nrow(se$assay), ncol(se$assay)))
print(se$col_data)

sub <- subset_samples(se, c("S1", "S3", "S6"))
cat("\nAfter subsetting, metadata travels with the data:\n")
print(sub$col_data)

cat("\nNow the classic bug - metadata sorted, assay not:\n")
scrambled <- se
scrambled$col_data <- se$col_data[order(se$col_data$condition,
                                        decreasing = TRUE),, drop = FALSE]
res <- try(validate_assay_data(scrambled), silent = TRUE)
cat("  Caught before it could do damage:\n   ",
    strsplit(attr(res, "condition")$message, "\n")[[1]][1], "\n")
cat("\nRULE: assert eq. (2.1) at the top of every script and after every merge.\n")
cat("In R, `merge()` REORDERS rows by default - a very common source of this.\n")
```

## 2. Support determines the likelihood

```{r scales}
header("2. Support -> natural model")
scale_table <- data.frame(
  Measurement = c("Read counts", "Counts out of N", "Log intensity",
                  "Methylation beta", "Relative abundance", "Time to event",
                  "Pathology grade"),
  Support = c("{0,1,2,...}", "{0..N}", "R", "[0,1]", "simplex", "[0,Inf) + censoring",
              "ordered categories"),
  Model = c("Poisson / negative binomial", "Binomial / beta-binomial", "Normal",
            "Beta / logit-normal", "Log-ratio / Dirichlet-mult.",
            "Survival / hazard", "Ordinal (proportional odds)"))
print(scale_table, row.names = FALSE)
```

## 3. log with a pseudocount, eq. (2.2)-(2.3)

$$\operatorname{Var}(\log_2(Y+c)) \approx
  \frac{1}{(\ln 2)^2}\cdot\frac{\mu+\phi\mu^2}{(\mu+c)^2}$$

```{r log-transform}
header("3. log(y + c): delta-method prediction (2.3) vs simulation")

## NOTE R's parameterisation: rnbinom(n, mu =, size = ) where size = 1/phi.
nb_sample <- function(n, mu, phi) rnbinom(n, mu = mu, size = 1 / phi)
delta_var_log2 <- function(mu, phi, c) (mu + phi * mu^2) / ((mu + c)^2 * log(2)^2)

PHI <- 0.16   # typical human bulk RNA-seq dispersion (BCV = 0.4)
cat(sprintf("dispersion phi = %.2f  (asymptote phi/(ln2)^2 = %.4f)\n\n",
            PHI, PHI / log(2)^2))
cat(sprintf("%8s %5s %13s %11s %7s\n", "mu", "c", "delta (2.3)", "simulated", "ratio"))
set.seed(21)
for (mu in c(1, 5, 20, 100, 1000)) {
  for (cc in c(0.5, 1, 8)) {
    y <- nb_sample(2e5, mu, PHI)
    emp <- var(log2(y + cc))
    theo <- delta_var_log2(mu, PHI, cc)
    cat(sprintf("%8d %5.1f %13.4f %11.4f %7.2f\n", mu, cc, theo, emp, emp / theo))
  }
}
cat("\nRead the c=0.5 rows: at mu=1 the variance is ~10x the asymptote.\n")
cat("Small pseudocounts hugely inflate the variance of low-count features,\n")
cat("which is why you filter them (stats.md Topics 8, 20).\n")
```

## 4. logit / M-value, eq. (2.4)

```{r mvalues}
header("4. beta vs M-values: heteroscedasticity (2.4)")
beta_to_m <- function(b, eps = 1e-6) { b <- pmin(pmax(b, eps), 1 - eps); log2(b / (1 - b)) }
m_to_beta <- function(m) 2^m / (2^m + 1)

set.seed(31)
cat(sprintf("%15s %10s %8s\n", "true mean beta", "SD(beta)", "SD(M)"))
for (target in c(0.02, 0.10, 0.30, 0.50, 0.70, 0.95)) {
  m_draws <- rnorm(5e4, log2(target / (1 - target)), 0.5)  # constant SD on M
  cat(sprintf("%15.2f %10.4f %8.4f\n", target, sd(m_to_beta(m_draws)), sd(m_draws)))
}
cat("\nSD on beta varies ~4x across the range; SD on M is constant by\n")
cat("construction. MODEL ON M, REPORT ON BETA (stats.md Topic 23).\n")
cat(sprintf("Round trip: M = 1.5 corresponds to beta = %.4f\n", m_to_beta(1.5)))
```

## 5. arcsinh, eq. (2.5)

```{r arcsinh}
header("5. arcsinh with cofactor c (2.5)")
## R has no asinh-with-cofactor helper, so define it. Note base R DOES have
## asinh(), which handles negatives natively - unlike log().
arcsinh_transform <- function(y, cofactor) asinh(y / cofactor)

test_values <- c(-50, -5, 0, 1, 5, 50, 500, 5000)
cat(sprintf("%10s %12s %13s %11s\n", "raw", "asinh c=5", "asinh c=150", "log10(y)"))
for (v in test_values) {
  lg <- if (v > 0) sprintf("%.3f", log10(v)) else "undefined"
  cat(sprintf("%10.1f %12.3f %13.3f %11s\n",
              v, arcsinh_transform(v, 5), arcsinh_transform(v, 150), lg))
}
cat("\nc=5 is the CyTOF convention; c=150 typical for fluorescence flow.\n")
cat("arcsinh handles the negative values compensation produces.\n")
```

## 6. Compositional closure, eq. (2.6)-(2.7)

```{r closure}
header("6. Closure manufactures correlations (2.6)-(2.7)")
set.seed(41)
D <- 5; n <- 3000
absolute <- sapply(c(3, 2.5, 2, 1.5, 1), function(m) rlnorm(n, m, 0.6))
relative <- absolute / rowSums(absolute)                    # closure (2.6)

corr_abs <- cor(absolute); corr_rel <- cor(relative)
off <- upper.tri(corr_abs)
cat(sprintf("Correlation among ABSOLUTE abundances (truth: independent):\n"))
cat(sprintf("  max |off-diagonal r| = %.3f\n", max(abs(corr_abs[off]))))
cat(sprintf("\nCorrelation among RELATIVE abundances (after closure):\n"))
cat(sprintf("  max |off-diagonal r| = %.3f\n", max(abs(corr_rel[off]))))
cat(sprintf("  mean off-diagonal r  = %+.3f  (negative!)\n", mean(corr_rel[off])))
cat("\nRow sums of the relative covariance matrix - eq. (2.7) predicts 0:\n  ")
cat(sprintf("%.2e ", rowSums(cov(relative))), "\n")
cat("\nEvery correlation you see in closed data was manufactured by the\n")
cat("constraint. This is why microbiome analysis needs log-ratios (Topic 26).\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "transformations.png"), width = 1300, height = 420, res = 110)
par(mfrow = c(1, 3), mar = c(4.2, 4.2, 2.5, 1))

set.seed(51)
mus <- 10^seq(0, 3.5, length.out = 30)
raw_sd <- sapply(mus, function(m) sd(nb_sample(2e4, m, PHI)))
log_sd <- sapply(mus, function(m) sd(log2(nb_sample(2e4, m, PHI) + 1)))
plot(mus, raw_sd, log = "xy", type = "b", pch = 16, cex = 0.6, col = "steelblue",
     xlab = "mean count mu", ylab = "standard deviation",
     main = "Eq. (2.2)-(2.3): log stabilises variance")
lines(mus, log_sd, type = "b", pch = 15, cex = 0.6, col = "darkorange")
abline(h = sqrt(PHI) / log(2), lty = 3)
legend("topleft", c("SD(raw counts)", "SD(log2(y+1))", "asymptote"),
       col = c("steelblue", "darkorange", "black"), lty = c(1, 1, 3),
       pch = c(16, 15, NA), bty = "n", cex = 0.7)

betas <- seq(0.01, 0.99, length.out = 60)
sd_beta <- sapply(betas, function(b) sd(m_to_beta(rnorm(4000, log2(b/(1-b)), 0.5))))
plot(betas, sd_beta, type = "l", lwd = 2, col = "steelblue",
     xlab = "mean methylation beta", ylab = "SD on the beta scale",
     main = "Eq. (2.4): beta is heteroscedastic")
abline(h = 0.5, col = "darkorange", lty = 2)
legend("topright", c("SD(beta)", "SD(M) = 0.5 (constant)"),
       col = c("steelblue", "darkorange"), lty = c(1, 2), bty = "n", cex = 0.7)

yy <- seq(-100, 5000, length.out = 2000)
plot(yy, asinh(yy / 5), type = "l", lwd = 2, col = "steelblue",
     xlab = "raw intensity", ylab = "transformed",
     main = "Eq. (2.5): arcsinh handles negatives")
lines(yy, asinh(yy / 150), lwd = 2, col = "darkorange")
lines(yy[yy > 0], log(yy[yy > 0]), lwd = 2, col = "forestgreen")
abline(v = 0, lty = 3)
legend("bottomright", c("arcsinh c=5", "arcsinh c=150", "ln(y)"),
       col = c("steelblue", "darkorange", "forestgreen"), lty = 1, bty = "n", cex = 0.7)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "transformations.png"), "\n")
```

## Decision rules (from `stats.md` Topic 2)

1. Ask what the number counts or measures before choosing a test.
2. Keep counts as counts for inference; transformed values for
   visualisation, distances and PCA.
3. Model on the scale where variance is stable; report on the scale a
   biologist can interpret.
4. Assert eq. (2.1) programmatically. Assert, do not assume.

**Next:** `03_exploratory_data_analysis.R`
