---
title: "Module 11 - Linear regression as the core framework"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 11, equations (11.1)-(11.13)

## What you will learn

1. OLS from the normal equations (11.2), the hat matrix (11.3), and why R
   uses QR rather than inverting `X'X`.
2. Coefficient SEs and t-tests (11.4)-(11.5).
3. Nested-model F tests (11.6); why R^2 cannot select models (11.7).
4. **Confidence vs prediction intervals** (11.8)-(11.9).
5. Leverage, Cook's distance (11.10)-(11.11), VIF (11.12).
6. Splines (11.13), and the assumption hierarchy.

```{r setup, message = FALSE}
suppressPackageStartupMessages(library(splines))
MODULE_NAME <- "11_linear_models"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. OLS from scratch, eq. (11.2)-(11.5)

```{r ols}
header("1. Normal equations, hat matrix, standard errors (11.2)-(11.5)")
set.seed(1101)
n <- 60
dat <- data.frame(age = runif(n, 25, 75), treated = rbinom(n, 1, 0.5),
                  batch = factor(sample(0:2, n, TRUE)))
dat$y <- with(dat, 2 + 0.05*age + 1.4*treated + 0.6*(batch=="1") -
                0.4*(batch=="2") + rnorm(n))

ols_from_scratch <- function(X, y) {
  ## QR, never solve(t(X) %*% X): forming the normal equations SQUARES the
  ## condition number (eq. 16.10) and loses half your significant digits.
  qrX <- qr(X); beta <- qr.coef(qrX, y)                      # eq. (11.2)
  fitted <- X %*% beta; resid <- y - fitted
  p <- ncol(X); sigma2 <- sum(resid^2) / (n - p)             # eq. (11.4)
  XtX_inv <- chol2inv(qr.R(qrX))
  se <- sqrt(sigma2 * diag(XtX_inv))
  t <- beta / se                                             # eq. (11.5)
  H <- qr.Q(qrX) %*% t(qr.Q(qrX))                            # eq. (11.3)
  list(beta = beta, se = se, t = t, p = 2*pt(abs(t), n-p, lower.tail = FALSE),
       H = H, leverage = diag(H), sigma2 = sigma2)
}
X <- model.matrix(~ age + treated + batch, data = dat)
mine <- ols_from_scratch(X, dat$y)
fit <- lm(y ~ age + treated + batch, data = dat)
cat(sprintf("  %-14s%13s%11s%9s%19s\n", "term", "beta (mine)", "SE (mine)", "t",
            "lm() beta"))
for (i in seq_along(mine$beta))
  cat(sprintf("  %-14s%13.5f%11.5f%9.3f%19.5f\n", names(mine$beta)[i],
              mine$beta[i], mine$se[i], mine$t[i], coef(fit)[i]))
cat(sprintf("\n  max |mine - lm()| = %.2e\n", max(abs(mine$beta - coef(fit)))))
cat(sprintf("\n  Hat matrix (11.3): symmetric? %s  idempotent? %s  trace = %.6f (= p = %d)\n",
            isTRUE(all.equal(mine$H, t(mine$H))),
            isTRUE(all.equal(mine$H %*% mine$H, mine$H)),
            sum(diag(mine$H)), ncol(X)))

cat("\n--- Numerical stability on a nearly-collinear design ---\n")
tt <- seq(0, 1, length.out = 40)
Xbad <- cbind(1, tt, tt^2, tt^3, tt^4, tt^5)
ybad <- 1 + 2*tt + rnorm(40, 0, 0.01)
cat(sprintf("  condition number of X   : %.3e\n", kappa(Xbad, exact = TRUE)))
cat(sprintf("  condition number of X'X : %.3e\n", kappa(crossprod(Xbad), exact = TRUE)))
b_normal <- solve(crossprod(Xbad), crossprod(Xbad, ybad))
b_qr <- qr.coef(qr(Xbad), ybad)
cat(sprintf("  max |normal-equations - QR|: %.3e\n", max(abs(b_normal - b_qr))))
cat("  lm() uses QR for exactly this reason.\n")
```

## 2. Nested-model F test and R^2, eq. (11.6)-(11.7)

```{r nested-f}
header("2. Partial F test (11.6); why R^2 cannot select models (11.7)")
fit_red <- lm(y ~ age + treated, data = dat)
av <- anova(fit_red, fit)
p1 <- length(coef(fit)); p0 <- length(coef(fit_red))
F_manual <- ((sum(residuals(fit_red)^2) - sum(residuals(fit)^2)) / (p1 - p0)) /
  (sum(residuals(fit)^2) / (n - p1))
cat(sprintf("  manual F (11.6) = %.4f, df=(%d, %d), p = %.4f\n",
            F_manual, p1 - p0, n - p1, pf(F_manual, p1-p0, n-p1, lower.tail = FALSE)))
cat(sprintf("  anova() F       = %.4f, p = %.4f\n", av[["F"]][2], av[["Pr(>F)"]][2]))

cat("\n  R^2 always increases when you add ANY predictor - even pure noise:\n")
d2 <- dat
cat(sprintf("    %-34s%9s%10s\n", "model", "R^2", "adj R^2"))
f0 <- lm(y ~ age, data = d2)
cat(sprintf("    %-34s%9.4f%10.4f\n", "y ~ age", summary(f0)$r.squared,
            summary(f0)$adj.r.squared))
set.seed(1102)
for (k in 1:5) {
  d2[[paste0("noise", k)]] <- rnorm(n)
  form <- as.formula(paste("y ~ age +", paste0("noise", 1:k, collapse = " + ")))
  f_ <- lm(form, data = d2)
  cat(sprintf("    %-34s%9.4f%10.4f\n", sprintf("+ %d pure-noise predictors", k),
              summary(f_)$r.squared, summary(f_)$adj.r.squared))
}
cat("  Adjusted R^2 penalises p, but neither is a validation metric (Module 31).\n")
```

## 3. Confidence vs prediction intervals, eq. (11.8)-(11.9)

```{r intervals}
header("3. CI for the mean vs PI for an observation (11.8)-(11.9)")
fit_s <- lm(y ~ age, data = dat)
nd <- data.frame(age = 50)
ci <- predict(fit_s, nd, interval = "confidence")
pi_ <- predict(fit_s, nd, interval = "prediction")
cat(sprintf("  at age = 50: fitted = %.4f\n", ci[1]))
cat(sprintf("    95%% CI for the MEAN  [%.4f, %.4f]  width %.4f\n",
            ci[2], ci[3], ci[3] - ci[2]))
cat(sprintf("    95%% PI for an OBS    [%.4f, %.4f]  width %.4f\n",
            pi_[2], pi_[3], pi_[3] - pi_[2]))
cat("\n  As n grows the CI shrinks to zero but the PI converges to ~ +/-1.96 sigma:\n")
set.seed(1103)
for (nn in c(30, 300, 3000, 30000)) {
  a <- runif(nn, 25, 75); yy <- 2 + 0.05*a + rnorm(nn)
  f_ <- lm(yy ~ a)
  c_ <- predict(f_, data.frame(a = 50), interval = "confidence")
  p_ <- predict(f_, data.frame(a = 50), interval = "prediction")
  cat(sprintf("    n=%6d: CI width %.4f   PI width %.4f\n", nn,
              c_[3]-c_[2], p_[3]-p_[2]))
}
cat("  The extra 1 in eq. (11.9) is irreducible noise. The PI does NOT shrink.\n")
```

## 4. Leverage, influence, collinearity, eq. (11.10)-(11.12)

```{r diagnostics}
header("4. Diagnostics: leverage, Cook's D, VIF (11.10)-(11.12)")
lev <- hatvalues(fit); ck <- cooks.distance(fit); p_ <- length(coef(fit))
cat(sprintf("  leverage rule of thumb 2p/n = %.4f: %d points flagged\n",
            2*p_/n, sum(lev > 2*p_/n)))
cat(sprintf("  Cook's D rule of thumb 4/n  = %.4f: %d points flagged\n",
            4/n, sum(ck > 4/n)))
cat(sprintf("  from-scratch leverage matches hatvalues(): %s\n",
            isTRUE(all.equal(unname(sort(lev)), sort(mine$leverage)))))
sr <- rstandard(fit)
cat("\n  High leverage != influential. Both are needed:\n")
cat(sprintf("    %-8s%10s%13s%10s\n", "point", "leverage", "|std resid|", "Cook D"))
for (i in order(-ck)[1:4])
  cat(sprintf("    %-8d%10.4f%13.4f%10.4f\n", i, lev[i], abs(sr[i]), ck[i]))

vif_manual <- function(X, j) {
  others <- X[, -j, drop = FALSE]
  1 / (1 - summary(lm(X[, j] ~ others - 1))$r.squared)          # eq. (11.12)
}
set.seed(1104)
z1 <- rnorm(n); z2 <- 0.97*z1 + rnorm(n, 0, 0.24)
Xv <- cbind(1, age = dat$age, z1 = z1, z2 = z2)
cat("\n  VIF (11.12) for a design with an induced collinearity:\n")
for (j in 2:4) {
  v <- vif_manual(Xv, j)
  cat(sprintf("    %-6s VIF = %7.2f  -> SE inflated %.2fx%s\n",
              colnames(Xv)[j], v, sqrt(v), if (v > 10) "   <-- serious" else ""))
}
```

## 5. Nonlinearity via splines, eq. (11.13)

```{r splines}
header("5. Splines: linear in the PARAMETERS (11.13)")
set.seed(1105)
tt <- sort(runif(120, 0, 24))
grp <- rep(0:1, 60)
expr <- 3 + 2*sin(tt/24*2*pi) + 0.06*tt + 0.8*grp*sin(tt/24*2*pi) + rnorm(120, 0, 0.35)
dt <- data.frame(expr = expr, t = tt, grp = factor(grp))
m_lin <- lm(expr ~ t, data = dt)
m_spl <- lm(expr ~ ns(t, df = 5), data = dt)
m_int <- lm(expr ~ ns(t, df = 5) * grp, data = dt)
cat(sprintf("  %-34s%5s%10s%10s\n", "model", "df", "RSS", "adj R^2"))
for (z in list(list("expr ~ t (linear)", m_lin), list("expr ~ ns(t, df=5)", m_spl),
               list("expr ~ ns(t, df=5) * group", m_int)))
  cat(sprintf("  %-34s%5d%10.3f%10.4f\n", z[[1]], z[[2]]$rank - 1,
              sum(residuals(z[[2]])^2), summary(z[[2]])$adj.r.squared))
av2 <- anova(m_spl, m_int)
cat(sprintf("\n  Treatment x time interaction test (11.6): F = %.3f, p = %.3e\n",
            av2[["F"]][2], av2[["Pr(>F)"]][2]))
cat("  This is how to ask 'do the groups follow different smooth trajectories?'\n")
cat("  NOTE ns() (natural splines) constrains the fit to be LINEAR beyond the\n")
cat("  boundary knots, which tames the wild tails of unrestricted polynomials.\n")
```

## 6. Which assumptions actually matter

```{r assumptions}
header("6. Ranking the assumption violations by damage done")
fpr_under <- function(violation, n_sim = 2000, n = 30) {
  mean(replicate(n_sim, {
    if (violation == "none") { x <- rnorm(n); y <- rnorm(n) }
    else if (violation == "non-normal residuals") { x <- rnorm(n); y <- rt(n, 3)*0.7 }
    else if (violation == "heteroscedasticity") {
      x <- rnorm(n); y <- rnorm(n, 0, 0.2 + 1.5*abs(x)) }
    else {   # dependence: 6 clusters of 5
      cl <- rep(rnorm(6, 0, 1.2), each = 5)
      x <- rep(rnorm(6), each = 5) + rnorm(n, 0, 0.3)
      y <- cl + rnorm(n, 0, 0.5) }
    coef(summary(lm(y ~ x)))["x", "Pr(>|t|)"] < 0.05
  }))
}
set.seed(1106)
cat(sprintf("  %-28s%21s  (nominal 5%%)\n", "violation", "false-positive rate"))
for (v in c("none", "non-normal residuals", "heteroscedasticity", "dependence (clusters)"))
  cat(sprintf("  %-28s%20.1f%%\n", v, 100 * fpr_under(v)))
cat("\n  Non-normal residuals: barely matters at n=30 (the CLT, eq. 4.5).\n")
cat("  Heteroscedasticity : distorts SEs - fix with HC3 (Module 05).\n")
cat("  DEPENDENCE         : catastrophic, and INVISIBLE in residual plots.\n")
cat("  Fix dependence by design or with a mixed model (Module 14).\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "linear_models.png"), width = 1100, height = 800, res = 110)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.5, 1))
plot(fit, which = 1, caption = "", main = "Residuals vs fitted (linearity)")
plot(fit, which = 2, caption = "", main = "Normal Q-Q (least important)")
plot(fit, which = 5, caption = "", main = "Eq. (11.10)-(11.11): leverage")
grid_t <- data.frame(t = seq(min(dt$t), max(dt$t), length.out = 200))
plot(dt$t, dt$expr, col = c("steelblue", "darkorange")[dt$grp], pch = 16, cex = 0.6,
     xlab = "time (h)", ylab = "expression",
     main = "Eq. (11.13): spline x group interaction")
for (g in 0:1) {
  gg <- grid_t; gg$grp <- factor(g, levels = c("0", "1"))
  lines(gg$t, predict(m_int, gg), col = c("steelblue", "darkorange")[g+1], lwd = 2)
}
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "linear_models.png"), "\n")
```

## Decision rules (from `stats.md` Topic 11)

1. Write the design matrix and check its rank first.
2. Choose covariates from a causal diagram (Module 32), NOT stepwise
   selection, which invalidates all subsequent p-values.
3. Centre continuous predictors involved in interactions.
4. Prefer robust SEs over dropping data when variance is non-constant.

**Next:** `12_anova_and_contrasts.R`
