---
title: "Module 15 - Missing data, censoring, and measurement limits"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 15, equations (15.1)-(15.9)

## What you will learn

1. MCAR / MAR / MNAR (15.1)-(15.3) simulated side by side.
2. Multiple imputation and **Rubin's rules (15.4)-(15.7)** from scratch.
3. Left-censoring as a **Tobit likelihood (15.8)**, not as missingness.
4. The one diagnostic plot every proteomics report needs.

```{r setup, message = FALSE}
MODULE_NAME <- "15_missing_data_and_censoring"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. The three mechanisms, eq. (15.1)-(15.3)

Critically, **MAR and MNAR cannot be distinguished from the observed data.**

```{r mechanisms}
header("1. Simulating MCAR, MAR and MNAR (15.1)-(15.3)")
make_data <- function(n = 600) {
  x <- rnorm(n); data.frame(x = x, y = 2 + 1.5*x + rnorm(n))
}
impose_missing <- function(df, mechanism, rate = 0.4) {
  n <- nrow(df)
  p <- switch(mechanism,
    MCAR = rep(rate, n),                                      # eq. (15.1)
    MAR  = plogis(2.2*df$x - log((1-rate)/rate)),             # eq. (15.2): on x
    MNAR = plogis(2.2*scale(df$y)[, 1] - log((1-rate)/rate))) # eq. (15.3): on y
  miss <- runif(n) < p
  df$missing <- miss; df$y[miss] <- NA
  df
}
TRUE_BETA <- 1.5
set.seed(1501)
cat(sprintf("  target estimand: beta_x = %.1f\n\n", TRUE_BETA))
cat(sprintf("  %-12s%11s%21s%9s\n", "mechanism", "% missing",
            "complete-case beta", "bias"))
for (mech in c("MCAR", "MAR", "MNAR")) {
  b <- replicate(200, {
    obs <- impose_missing(make_data(), mech)
    coef(lm(y ~ x, data = obs[!is.na(obs$y), ]))["x"]
  })
  obs1 <- impose_missing(make_data(), mech)
  cat(sprintf("  %-12s%10.1f%%%21.4f%9.4f\n", mech, 100*mean(obs1$missing),
              mean(b), mean(b) - TRUE_BETA))
}
cat("\n  MCAR: complete-case is unbiased (just wasteful).\n")
cat("  MAR : unbiased HERE because missingness depends only on x, which is IN\n")
cat("        the model. Not general - depend on an omitted covariate and it\n")
cat("        would be biased.\n")
cat("  MNAR: biased, and no imputation fixes it without an untestable\n")
cat("        assumption.\n")

```
```{r mar-covariate}
cat("\n--- MAR in a COVARIATE, depending on the outcome ---\n")
set.seed(1502)
bias_cc <- replicate(150, {
  n <- 500
  x <- rnorm(n); zz <- 0.8*x + rnorm(n)
  y <- 2 + 1.5*x + 0.7*zz + rnorm(n)
  x[runif(n) < plogis(1.5*scale(y)[, 1] - 0.4)] <- NA     # MAR: depends on y
  d <- data.frame(x = x, z = zz, y = y)
  cc <- d[complete.cases(d), ]
  if (nrow(cc) > 30) coef(lm(y ~ x + z, data = cc))["x"] - 1.5 else NA
})
cat(sprintf("  complete-case bias in beta_x: %+.4f\n", mean(bias_cc, na.rm = TRUE)))
cat("  Under MAR the fix is a method that CONDITIONS on the observed data\n")
cat("  driving missingness - multiple imputation INCLUDING y.\n")
```

## 2. Multiple imputation and Rubin's rules, eq. (15.4)-(15.7)

$$\bar Q=\frac1M\sum\hat Q_m,\qquad T=\bar U+\Big(1+\frac1M\Big)B$$

The `(1+1/M)B` term is exactly what single imputation omits.

```{r rubin}
header("2. Rubin's rules from scratch (15.4)-(15.7)")
impute_once <- function(df, target, predictors) {
  ## Stochastic regression imputation (the core step of MICE). Adding residual
  ## noise AND drawing beta from its sampling distribution is what makes it
  ## 'proper'; predicting the conditional mean alone shrinks variability.
  cc <- df[complete.cases(df[, c(target, predictors), drop = FALSE]), ]
  f <- lm(reformulate(predictors, target), data = cc)
  miss <- is.na(df[[target]])
  if (!any(miss)) return(df[[target]])
  beta_draw <- MASS::mvrnorm(1, coef(f), vcov(f))
  Xm <- model.matrix(reformulate(predictors), data = df[miss,, drop = FALSE])
  out <- df[[target]]
  out[miss] <- as.vector(Xm %*% beta_draw) + rnorm(sum(miss), 0, sigma(f))
  out
}
rubin_pool <- function(est, vars, M) {
  Q <- mean(est); U <- mean(vars)                      # eq. (15.4)-(15.5)
  B <- if (M > 1) var(est) else 0
  Tt <- U + (1 + 1/M) * B                              # eq. (15.6)
  gamma <- if (Tt > 0) (1 + 1/M) * B / Tt else 0
  nu <- if (B > 0) (M - 1) * (1 + U / ((1 + 1/M) * B))^2 else Inf   # eq. (15.7)
  list(estimate = Q, se = sqrt(Tt), df = nu, fmi = gamma)
}
run_mi <- function(df, M) {
  est <- numeric(M); vars <- numeric(M)
  for (m in seq_len(M)) {
    d <- df; d$y <- impute_once(d, "y", "x")
    f <- lm(y ~ x, data = d)
    est[m] <- coef(f)["x"]; vars[m] <- coef(summary(f))["x", "Std. Error"]^2
  }
  rubin_pool(est, vars, M)
}
set.seed(1503)
cat("  Coverage of a nominal 95% CI for beta_x, MCAR with 40% of y missing:\n")
cover <- c(`complete case` = 0, `single imputation` = 0, `MI (M=20)` = 0)
width <- list(a = c(), b = c(), c = c())
N_REP <- 200
for (rep in seq_len(N_REP)) {
  obs <- impose_missing(make_data(n = 300), "MCAR")
  f_cc <- lm(y ~ x, data = obs[!is.na(obs$y), ])
  ci <- confint(f_cc)["x", ]
  cover[1] <- cover[1] + (ci[1] <= TRUE_BETA && TRUE_BETA <= ci[2])
  width$a <- c(width$a, diff(ci))
  r1 <- run_mi(obs, 1)
  ci1 <- r1$estimate + c(-1, 1) * 1.96 * r1$se
  cover[2] <- cover[2] + (ci1[1] <= TRUE_BETA && TRUE_BETA <= ci1[2])
  width$b <- c(width$b, diff(ci1))
  r20 <- run_mi(obs, 20)
  ci20 <- r20$estimate + c(-1, 1) * qt(0.975, min(r20$df, 1e6)) * r20$se
  cover[3] <- cover[3] + (ci20[1] <= TRUE_BETA && TRUE_BETA <= ci20[2])
  width$c <- c(width$c, diff(ci20))
}
cat(sprintf("  %-22s%11s%16s\n", "method", "coverage", "mean CI width"))
for (i in 1:3)
  cat(sprintf("  %-22s%10.1f%%%16.4f\n", names(cover)[i], 100*cover[i]/N_REP,
              mean(width[[i]])))
cat("\n  Single imputation treats imputed values as OBSERVED, so its CI is too\n")
cat("  narrow and UNDERCOVERS. Multiple imputation adds (1+1/M)B and recovers\n")
cat("  nominal coverage.\n")
set.seed(99)
r <- run_mi(impose_missing(make_data(n = 300), "MCAR"), 20)
cat(sprintf("\n  Example pooled result: beta = %.4f, SE = %.4f, df = %.1f\n",
            r$estimate, r$se, r$df))
cat(sprintf("  fraction of missing information = %.3f   (eq. 15.7)\n", r$fmi))
cat("  Use M >= 20 when the FMI is substantial.\n")
```

## 3. Left-censoring is NOT missingness, eq. (15.8)

```{r tobit}
header("3. Limit of detection: censored likelihood vs substitution (15.8)")
tobit_mle <- function(y_obs, censored, L) {
  n_cen <- sum(censored)
  nll <- function(par) {
    mu <- par[1]; sigma <- exp(par[2])
    ll_obs <- sum(dnorm(y_obs[!censored], mu, sigma, log = TRUE))
    ## Each censored point contributes log Phi((L-mu)/sigma). Multiplying by
    ## n_cen is essential - the term is identical for every censored point.
    ll_cen <- n_cen * pnorm((L - mu) / sigma, log.p = TRUE)
    -(ll_obs + ll_cen)
  }
  st <- c(mean(y_obs[!censored]), log(max(sd(y_obs[!censored]), 1e-3)))
  o <- optim(st, nll, method = "Nelder-Mead",
             control = list(reltol = 1e-12, maxit = 5000))
  c(mu = o$par[1], sigma = exp(o$par[2]))
}
set.seed(1504)
TRUE_MU <- 20; TRUE_SD <- 3; L <- 18
res <- t(replicate(400, {
  z <- rnorm(120, TRUE_MU, TRUE_SD); cen <- z < L
  y_half <- ifelse(cen, L/2, z); y_sq2 <- ifelse(cen, L/sqrt(2), z)
  y_lod <- ifelse(cen, L, z)
  tob <- tobit_mle(ifelse(cen, L, z), cen, L)
  c(drop_mu = mean(z[!cen]), drop_sd = sd(z[!cen]),
    half_mu = mean(y_half), half_sd = sd(y_half),
    sq2_mu = mean(y_sq2), sq2_sd = sd(y_sq2),
    lod_mu = mean(y_lod), lod_sd = sd(y_lod),
    ## unname(): tob carries names, and replicate() would otherwise mangle
    ## the column labels into "tobit_mu.mu".
    tobit_mu = unname(tob["mu"]), tobit_sd = unname(tob["sigma"]),
    pct = mean(cen))
}))
cat(sprintf("  truth: mu = %g, sigma = %g, LOD = %g; mean censored %.1f%%\n\n",
            TRUE_MU, TRUE_SD, L, 100*mean(res[, "pct"])))
cat(sprintf("  %-28s%13s%9s%16s%9s\n", "strategy", "mean mu_hat", "bias",
            "mean sigma_hat", "bias"))
for (row_ in list(c("drop", "drop censored values"), c("half", "substitute L/2"),
                  c("sq2", "substitute L/sqrt(2)"), c("lod", "substitute L"),
                  c("tobit", "censored MLE (15.8)"))) {
  mu_m <- mean(res[, paste0(row_[1], "_mu")]); sd_m <- mean(res[, paste0(row_[1], "_sd")])
  cat(sprintf("  %-28s%13.4f%9.4f%16.4f%9.4f\n", row_[2], mu_m, mu_m - TRUE_MU,
              sd_m, sd_m - TRUE_SD))
}
cat("\n  Every substitution rule is biased in BOTH the mean and the SD. The\n")
cat("  censored likelihood recovers both essentially without bias.\n")
```

## 4. The diagnostic every proteomics report needs

```{r diagnostic}
header("4. Diagnosing the mechanism from missingness vs abundance")
set.seed(1505)
G <- 1500; n <- 20
true_abundance <- rnorm(G, 22, 2.5)
Y <- matrix(rnorm(G * n, true_abundance, 0.8), nrow = G)
is_mnar <- runif(G) < 0.7
p_missing <- ifelse(is_mnar, plogis(-(true_abundance - 20.5) / 0.6), 0.15)
M <- matrix(runif(G * n), nrow = G) < p_missing
Yobs <- Y; Yobs[M] <- NA
miss_rate <- rowMeans(is.na(Yobs))
mean_obs <- rowMeans(Yobs, na.rm = TRUE)
ok <- is.finite(mean_obs)
sp <- cor(mean_obs[ok], miss_rate[ok], method = "spearman")
cat(sprintf("  Spearman(mean observed abundance, missingness rate) = %+.3f\n", sp))
cat("  A strong NEGATIVE correlation => abundance-dependent (MNAR) missingness.\n")
qs <- cut(mean_obs[ok], quantile(mean_obs[ok], 0:5/5), include.lowest = TRUE,
          labels = paste0("Q", 1:5))
cat(sprintf("\n  %-26s%18s\n", "mean-abundance quintile", "mean missingness"))
for (lv in levels(qs))
  cat(sprintf("  %-26s%18.3f\n", lv, mean(miss_rate[ok][qs == lv])))
cat("\n  Treat MNAR and MAR features DIFFERENTLY. Applying one imputer to both\n")
cat("  is the standard mistake (stats.md Topic 25, Module 35).\n")
```

## 5. Imputing before cross-validation leaks information

Imputation is a LEARNED step. Fit it on all the rows and the held-out rows
have already contributed to how they themselves are filled.

```{r impute-leak}
header("5. Imputing before cross-validation leaks information")
impute_mean <- function(X) {
  cm <- colMeans(X, na.rm = TRUE); cm[!is.finite(cm)] <- 0
  for (j in seq_len(ncol(X))) X[is.na(X[, j]), j] <- cm[j]
  X
}
```

kNN imputation: fill each missing cell from the k most similar ROWS. This
is what makes it leak - the neighbours can include the test rows.

```{r}
impute_knn <- function(X, k = 5) {
  cm <- colMeans(X, na.rm = TRUE); cm[!is.finite(cm)] <- 0
  Xf <- impute_mean(X)
  out <- X
  D <- as.matrix(dist(Xf))
  for (i in seq_len(nrow(X))) {
    miss <- which(is.na(X[i, ]))
    if (!length(miss)) next
    nb <- order(D[i, ])[2:(k + 1)]
    for (j in miss) {
      v <- X[nb, j]; v <- v[!is.na(v)]
      out[i, j] <- if (length(v)) mean(v) else cm[j]
    }
  }
  out
}
cv_r2 <- function(X, y, impute_inside, imputer, seed = 0) {
  set.seed(seed)
  fold <- sample(rep_len(1:5, nrow(X)))
  if (!impute_inside) X <- imputer(X)     # LEAKY: uses every row, then CV
  sc <- numeric(5)
  for (k in 1:5) {
    tr <- fold != k; te <- !tr
    Xtr <- X[tr, , drop = FALSE]; Xte <- X[te, , drop = FALSE]
    if (impute_inside) {
      ## Fit the imputer on the training rows, then apply it to the test rows
      ## WITHOUT letting them influence each other.
      Xtr <- imputer(Xtr)
      cm <- colMeans(Xtr); cm[!is.finite(cm)] <- 0
      for (j in seq_len(ncol(Xte))) Xte[is.na(Xte[, j]), j] <- cm[j]
    }
    m <- lm(y[tr] ~ ., data.frame(Xtr))
    pr <- predict(m, data.frame(Xte))
    sc[k] <- 1 - sum((y[te] - pr)^2) / sum((y[te] - mean(y[tr]))^2)
  }
  mean(sc)
}
n <- 120; p <- 40
cat("  Data contain NO signal; honest R^2 should be <= 0.\n")
cat("  Averaged over 20 datasets, so the comparison is not one draw:\n\n")
cat(sprintf("  %-34s%19s%18s%11s\n", "imputer", "before CV (leaky)",
            "inside the fold", "optimism"))
for (z in list(list("column mean (no row borrowing)", impute_mean),
               list("kNN, k=5 (borrows across ROWS)", impute_knn))) {
  lk <- hn <- numeric(20)
  for (rep_ in 1:20) {
    set.seed(1600 + rep_)
    Xr <- matrix(rnorm(n * p), n)
    yr <- rnorm(n)                                # NO real signal at all
    Xr[matrix(runif(n * p), n) < 0.3] <- NA
    lk[rep_] <- cv_r2(Xr, yr, FALSE, z[[2]], seed = rep_)
    hn[rep_] <- cv_r2(Xr, yr, TRUE, z[[2]], seed = rep_)
  }
  cat(sprintf("  %-34s%19.4f%18.4f%+11.4f\n", z[[1]], mean(lk), mean(hn),
              mean(lk) - mean(hn)))
}
cat("
  Both rows should be at or below zero, because there is nothing to predict.

  Mean imputation leaks little: it only ever uses COLUMN statistics, so a
  held-out row contributes almost nothing to how it is filled.

  kNN imputation leaks more, and the reason is structural: it fills a cell by
  copying from the most similar ROWS, and when it is fitted on all the data
  those neighbours include the test rows themselves. The test row is partly
  reconstructed from its own values before the model ever sees it.

  This is the rule behind the classification in Module 21: what matters is not
  whether a step is 'preprocessing' but whether it BORROWS ACROSS ROWS. Fit
  every learned step inside the fold so the boundary is enforced structurally
  rather than by discipline.\n")
```

## 6. Figure

```{r figure}
png(file.path(OUT, "missing_data.png"), width = 1300, height = 420, res = 110)
par(mfrow = c(1, 3), mar = c(4.2, 4.2, 2.5, 1))
set.seed(77)
full <- make_data(n = 800)
plot(NA, xlim = range(full$x), ylim = range(full$y), xlab = "x",
     ylab = "y (true value)", main = "Which points go missing")
cols <- c("steelblue", "darkorange", "firebrick")
for (i in seq_along(c("MCAR", "MAR", "MNAR"))) {
  obs <- impose_missing(full, c("MCAR", "MAR", "MNAR")[i])
  points(full$x[obs$missing], full$y[obs$missing], pch = 16, cex = 0.4,
         col = adjustcolor(cols[i], 0.5))
}
legend("topleft", c("MCAR", "MAR", "MNAR"), col = cols, pch = 16, bty = "n", cex = 0.7)
zz <- rnorm(3000, TRUE_MU, TRUE_SD)
hist(zz, breaks = 60, col = adjustcolor("steelblue", 0.5), border = "white",
     main = "Eq. (15.8): left-censoring", xlab = "value")
hist(zz[zz >= L], breaks = 60, col = adjustcolor("darkorange", 0.7), border = "white",
     add = TRUE)
abline(v = L, col = "red", lwd = 2)
plot(mean_obs[ok], miss_rate[ok], pch = ".", col = "steelblue",
     xlab = "mean observed abundance", ylab = "missingness rate",
     main = sprintf("MNAR signature (Spearman %+.2f)", sp))
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "missing_data.png"), "\n")
```

## Decision rules (from `stats.md` Topic 15)

1. Name the mechanism for each variable; you cannot TEST MAR vs MNAR.
2. Never impute the outcome to create observations.
3. Use M >= 20 imputations, and INCLUDE the outcome in the imputation model.
4. Model detection limits as CENSORING (15.8), not as missingness.
5. Report a sensitivity analysis over plausible MNAR departures.

**Next:** `16_matrix_algebra.R`
