---
title: "Module 17 - Principal component analysis"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 17, equations (17.1)-(17.6)

## What you will learn

1. PCA is one eigenproblem (17.1), computed via SVD (17.2)-(17.3).
2. Centering is mandatory; **scaling is a decision** (17.4).
3. How many components: parallel analysis and Marchenko-Pastur (17.5).
4. Sign and rotation indeterminacy.
5. **PC x metadata association (17.6)** -- the key bioinformatics output.
6. Over-reading separation, and double dipping.

```{r setup, message = FALSE}
MODULE_NAME <- "17_pca"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. PCA from scratch: eigen and SVD agree, eq. (17.1)-(17.3)

```{r pca-scratch}
header("1. Eigen-decomposition == SVD (17.1)-(17.3)")
pca_manual <- function(X, scale. = FALSE) {
  Xc <- scale(X, center = TRUE, scale = scale.)      # centering is MANDATORY
  n <- nrow(Xc)
  sv <- svd(Xc)
  list(scores = sv$u %*% diag(sv$d), loadings = sv$v,
       eigenvalues = sv$d^2 / (n - 1), pve = sv$d^2 / sum(sv$d^2), centered = Xc)
}
set.seed(1701)
n_s <- 60; p_f <- 300
Z <- matrix(rnorm(n_s * 2), n_s)
X <- Z %*% matrix(rnorm(2 * p_f), 2) + matrix(rnorm(n_s * p_f, 0, 0.8), n_s)
res <- pca_manual(X)
S <- cov(res$centered)
ev_eig <- sort(eigen(S, only.values = TRUE)$values, TRUE)[1:6]
cat(sprintf("  eigenvalues from covariance eigen: %s\n",
            paste(sprintf("%.4f", ev_eig), collapse = " ")))
cat(sprintf("  eigenvalues from SVD, d^2/(n-1)  : %s\n",
            paste(sprintf("%.4f", res$eigenvalues[1:6]), collapse = " ")))
cat(sprintf("  max |difference| = %.2e\n", max(abs(ev_eig - res$eigenvalues[1:6]))))
pr <- prcomp(X, center = TRUE, scale. = FALSE)
cat(sprintf("\n  prcomp() PVE : %s\n",
            paste(sprintf("%.4f", (pr$sdev^2/sum(pr$sdev^2))[1:6]), collapse = " ")))
cat(sprintf("  manual   PVE : %s\n",
            paste(sprintf("%.4f", res$pve[1:6]), collapse = " ")))
recon <- res$scores[, 1:2] %*% t(res$loadings[, 1:2])
cat(sprintf("\n  rank-2 reconstruction captures %.1f%% of the total variance\n",
            100 * (1 - sum((res$centered - recon)^2) / sum(res$centered^2))))
```

## 2. Centering is mandatory; scaling is a decision, eq. (17.4)

```{r centering-scaling}
header("2. Centering and scaling (17.4)")
X_unc <- X + 50
sv_unc <- svd(X_unc)
cat(sprintf("  WITHOUT centering, PC1 explains %.1f%% of variance\n",
            100 * sv_unc$d[1]^2 / sum(sv_unc$d^2)))
cat(sprintf("  |cor(PC1 loading, mean vector)| = %.4f\n",
            abs(cor(sv_unc$v[, 1], colMeans(X_unc)))))
cat("  Without centering, PC1 just points at the mean. ALWAYS centre.\n")
cat("  (prcomp() centres by default; princomp() also does. Never turn it off.)\n")
set.seed(1702)
clin <- cbind(HbA1c = rnorm(100, 5.5, 0.8), chol = rnorm(100, 200, 40),
              wbc = rnorm(100, 7000, 2500))
cat("\n  Scaling matters when features are on different scales:\n")
for (sc in c(FALSE, TRUE)) {
  r <- pca_manual(clin, scale. = sc)
  cat(sprintf("    scale=%-5s: PVE1=%.3f, |PC1 loadings| = %s\n", sc, r$pve[1],
              paste(sprintf("%.3f", abs(r$loadings[, 1])), collapse = " ")))
}
cat("  Unscaled, PC1 is entirely the variable with the largest UNITS (wbc),\n")
cat("  an artefact of measurement units, not biology.\n")
cat("\n  RULE: scale when units differ (clinical panels); do NOT scale when\n")
cat("  features share units and the variance differences ARE the biology\n")
cat("  (log-expression, M-values, arcsinh cytometry).\n")
```

## 3. How many components? eq. (17.5)

```{r how-many}
header("3. Selecting components: parallel analysis and Marchenko-Pastur (17.5)")
parallel_analysis <- function(X, n_perm = 60, scale. = TRUE) {
  obs <- pca_manual(X, scale.)$eigenvalues
  null <- replicate(n_perm, {
    Xp <- apply(X, 2, sample)      # permute each feature INDEPENDENTLY
    pca_manual(Xp, scale.)$eigenvalues[seq_along(obs)]
  })
  list(obs = obs, thresh = apply(null, 1, quantile, 0.95))
}
set.seed(1703)
pa <- parallel_analysis(X, n_perm = 40)
gamma <- p_f / n_s
mp_edge <- (1 + sqrt(gamma))^2                                  # eq. (17.5)
cat(sprintf("  data: n=%d, p=%d, gamma = p/n = %.2f\n", n_s, p_f, gamma))
cat(sprintf("  Marchenko-Pastur upper edge (1+sqrt(gamma))^2 = %.3f\n\n", mp_edge))
cat(sprintf("  %4s%13s%23s%8s\n", "PC", "eigenvalue", "95th pct of permuted", "keep?"))
for (j in 1:6)
  cat(sprintf("  %4d%13.3f%23.3f%8s\n", j, pa$obs[j], pa$thresh[j],
              ifelse(pa$obs[j] > pa$thresh[j], "yes", "no")))
cat(sprintf("\n  parallel analysis keeps %d component(s); the data were generated\n",
            sum(pa$obs > pa$thresh)))
cat("  from 2 latent factors.\n")
cat("\n  How large do PURE NOISE eigenvalues get?\n")
cat(sprintf("  %6s%7s%8s%14s%10s\n", "n", "p", "gamma", "observed max", "MP edge"))
for (cfg in list(c(50,50), c(50,500), c(50,2000))) {
  set.seed(cfg[1]*cfg[2])
  Xn <- matrix(rnorm(cfg[1]*cfg[2]), cfg[1])
  e <- pca_manual(Xn, scale. = TRUE)$eigenvalues
  g <- cfg[2]/cfg[1]
  cat(sprintf("  %6d%7d%8.1f%14.2f%10.2f\n", cfg[1], cfg[2], g, e[1], (1+sqrt(g))^2))
}
cat("  With p >> n, noise alone produces enormous leading eigenvalues.\n")
```

## 4. Sign and rotation indeterminacy

```{r sign}
header("4. PC signs are arbitrary; near-tied PCs are unstable")
r1 <- pca_manual(X); r2 <- pca_manual(X[, ncol(X):1])
cat(sprintf("  PC1 PVE, original order : %.6f\n", r1$pve[1]))
cat(sprintf("  PC1 PVE, reversed order : %.6f   <- identical\n", r2$pve[1]))
cat(sprintf("  correlation of the two PC1 score vectors: %+.6f\n",
            cor(r1$scores[, 1], r2$scores[, 1])))
cat("  |correlation| = 1, but the SIGN can flip. Never interpret the sign of\n")
cat("  a PC without anchoring it to a known feature.\n")
set.seed(1704)
base <- matrix(rnorm(80*2), 80) %*% diag(c(2.0, 1.98))    # NEARLY equal
Xtie <- cbind(base, matrix(rnorm(80*20, 0, 0.3), 80))
orig <- pca_manual(Xtie)$loadings[, 1]
sims <- replicate(50, {
  idx <- sample(80, 80, TRUE)
  abs(sum(pca_manual(Xtie[idx, ])$loadings[, 1] * orig))
})
cat(sprintf("\n  Near-tied eigenvalues -> unstable individual eigenvectors:\n"))
cat(sprintf("    eigenvalues 1,2 = %s\n",
            paste(sprintf("%.3f", pca_manual(Xtie)$eigenvalues[1:2]), collapse = " ")))
cat(sprintf("    mean |cos| between bootstrap PC1 and original PC1: %.3f\n", mean(sims)))
cat("    Only the SUBSPACE spanned by tied PCs is well defined.\n")
```

## 5. PC x metadata association, eq. (17.6)

```{r pc-metadata}
header("5. Which covariate explains each PC? (17.6)")
set.seed(1705)
n_samp <- 40; G <- 2000
meta <- data.frame(condition = rep(c("ctrl","trt"), n_samp/2),
                   batch = rep(c("b1","b2","b3","b4"), each = n_samp/4),
                   sex = sample(c("F","M"), n_samp, TRUE),
                   lib_size = rlnorm(n_samp, log(2e7), 0.3))
expr <- matrix(rnorm(n_samp * G), n_samp)
batch_load <- matrix(rnorm(4 * G), 4)
for (i in 1:4) expr[meta$batch == paste0("b", i), ] <-
  expr[meta$batch == paste0("b", i), ] + 2.2 * rep(batch_load[i, ], each = sum(meta$batch == paste0("b", i)))
expr[meta$condition == "trt", 1:200] <- expr[meta$condition == "trt", 1:200] + 0.9
expr <- expr + 0.6 * as.vector(scale(log(meta$lib_size)))
pc <- pca_manual(expr)
pc_metadata_r2 <- function(scores, meta, n_pc = 5) {
  out <- data.frame(PC = 1:n_pc, PVE = NA)
  for (col in names(meta)) out[[col]] <- NA
  for (j in 1:n_pc) {
    zj <- scores[, j]; out$PVE[j] <- NA
    for (col in names(meta)) {
      v <- meta[[col]]
      out[[col]][j] <- summary(lm(zj ~ if (is.numeric(v)) v else factor(v)))$r.squared
    }
  }
  out
}
tabm <- pc_metadata_r2(pc$scores, meta)
tabm$PVE <- pc$pve[1:5]
print(format(tabm, digits = 3), row.names = FALSE)
cat("\n  PC1 is overwhelmingly explained by BATCH, not condition. That is a\n")
cat("  technical finding, not a discovery. Build this table for EVERY dataset\n")
cat("  before interpreting a single PCA plot (stats.md Topic 19).\n")
```

## 6. Two errors: over-reading separation, and double dipping

```{r errors}
header("6. Separation in PC space is not evidence")
set.seed(1706)
n_samp <- 20; G <- 5000
Xnull <- matrix(rnorm(n_samp * G), n_samp)          # NO group structure at all
labels <- rep(0:1, n_samp/2)
pn <- pca_manual(Xnull)
obs_sep <- abs(mean(pn$scores[labels == 0, 1]) - mean(pn$scores[labels == 1, 1]))
null_sep <- replicate(2000, {
  perm <- sample(labels)
  abs(mean(pn$scores[perm == 0, 1]) - mean(pn$scores[perm == 1, 1]))
})
cat(sprintf("  n=%d, p=%d, NO real group difference\n", n_samp, G))
cat(sprintf("  observed PC1 separation between arbitrary labels: %.3f\n", obs_sep))
cat(sprintf("  permutation p-value: %.3f\n", (1 + sum(null_sep >= obs_sep))/2001))
cat("  Random high-dimensional data ALWAYS spread out in the top PCs.\n")
cat("  Permute the labels to get a null for any separation you plan to claim.\n")
grp <- as.integer(pn$scores[, 1] > median(pn$scores[, 1]))
top_feats <- order(-abs(pn$loadings[, 1]))[1:200]
p_dip <- sapply(top_feats, function(j) t.test(Xnull[grp==0, j], Xnull[grp==1, j])$p.value)
rand_feats <- sample(G, 200)
p_rand <- sapply(rand_feats, function(j) t.test(Xnull[grp==0, j], Xnull[grp==1, j])$p.value)
cat(sprintf("\n--- Double dipping: selecting features by PC1, then testing them ---\n"))
cat(sprintf("  features selected by PC1 loading : %.1f%% 'significant' (pure noise)\n",
            100*mean(p_dip < 0.05)))
cat(sprintf("  randomly selected features       : %.1f%%\n", 100*mean(p_rand < 0.05)))
cat("\n  Random features are only mildly inflated (the GROUPS came from PC1).\n")
cat("  Selecting the FEATURES from the same PC1 as well is what produces the\n")
cat("  large inflation. Split the data, or use a selective-inference test.\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "pca.png"), width = 1100, height = 800, res = 110)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.5, 1))
plot(1:15, pa$obs[1:15], type = "b", pch = 16, col = "steelblue",
     xlab = "component", ylab = "eigenvalue", main = "Eq. (17.5): how many components?")
lines(1:15, pa$thresh[1:15], type = "b", pch = 4, col = "darkorange")
abline(h = mp_edge, col = "forestgreen", lty = 3)
legend("topright", c("observed", "95th pct permuted", "MP edge"),
       col = c("steelblue", "darkorange", "forestgreen"), pch = c(16, 4, NA),
       lty = c(1, 1, 3), bty = "n", cex = 0.65)
plot(pc$scores[, 1], pc$scores[, 2], pch = c(16,17,15,18)[factor(meta$batch)],
     col = "steelblue", cex = 1.2, asp = 1,
     xlab = sprintf("PC1 (%.0f%%)", 100*pc$pve[1]),
     ylab = sprintf("PC2 (%.0f%%)", 100*pc$pve[2]), main = "PCA coloured by BATCH")
legend("topright", levels(factor(meta$batch)), pch = c(16,17,15,18), bty = "n", cex = 0.65)
plot(pc$scores[, 1], pc$scores[, 2], pch = c(16,17)[factor(meta$condition)],
     col = "firebrick", cex = 1.2, asp = 1,
     xlab = sprintf("PC1 (%.0f%%)", 100*pc$pve[1]),
     ylab = sprintf("PC2 (%.0f%%)", 100*pc$pve[2]),
     main = "Same PCA coloured by CONDITION")
legend("topright", levels(factor(meta$condition)), pch = c(16,17), bty = "n", cex = 0.65)
hist(null_sep, breaks = 50, col = "lightsteelblue", border = "white",
     main = "Separation on NOISE: permutation null", xlab = "PC1 separation")
abline(v = obs_sep, col = "red", lwd = 2)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "pca.png"), "\n")
```

## Decision rules (from `stats.md` Topic 17)

1. Always centre. Scale only with a stated reason.
2. Interpret PCs through loadings and metadata associations (17.6).
3. Report the PVE for every displayed component; use equal aspect ratio.
4. Never test features on the same data that defined the component.

**Next:** `18_distances_and_clustering.R`
