---
title: "Module 18 - Distances, clustering, and embeddings"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 18, equations (18.1)-(18.14)

Clustering always returns clusters. Which would survive a different distance,
a different seed, or a different half of the data?

## What you will learn

1. The distance zoo (18.1)-(18.5) and the Euclidean/correlation identity (18.6).
2. The curse of dimensionality (18.7).
3. Linkage choice (18.8)-(18.9) and k-means (18.10).
4. Choosing K: silhouette (18.11), gap statistic (18.12), **stability**.
5. Double dipping: testing the features that defined the clusters.

```{r setup, message = FALSE}
suppressPackageStartupMessages(library(cluster))
MODULE_NAME <- "18_distances_and_clustering"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. The distance zoo, eq. (18.1)-(18.6)

```{r distances}
header("1. Different distances answer different questions (18.1)-(18.5)")
profiles <- rbind(geneA = 1:6, geneB = 11:16, geneC = 6:1)
corr_distance <- function(a, b) 1 - cor(a, b)                   # eq. (18.2)
cosine_distance <- function(a, b) 1 - sum(a*b)/(sqrt(sum(a^2))*sqrt(sum(b^2)))
bray_curtis <- function(a, b) sum(abs(a-b)) / sum(a+b)          # eq. (18.4)
cat(sprintf("  %-14s%11s%13s%9s%13s\n", "pair", "Euclidean", "correlation",
            "cosine", "Bray-Curtis"))
for (pr in list(c("geneA","geneB"), c("geneA","geneC"), c("geneB","geneC"))) {
  a <- profiles[pr[1], ]; b <- profiles[pr[2], ]
  cat(sprintf("  %-14s%11.3f%13.3f%9.3f%13.3f\n", paste(pr, collapse="-"),
              sqrt(sum((a-b)^2)), corr_distance(a,b), cosine_distance(a,b),
              bray_curtis(a,b)))
}
cat("\n  A-B: identical SHAPE, different level. Correlation distance says 0;\n")
cat("  Euclidean says 24.5. Choose the distance from the QUESTION.\n")
set.seed(1801)
p <- 50; xv <- rnorm(p); yv <- 0.6*xv + rnorm(p)
zx <- scale(xv)[, 1]; zy <- scale(yv)[, 1]
cat(sprintf("\n  Verifying eq. (18.6): d^2(z_x, z_y) = 2(p-1)(1-r)\n"))
cat(sprintf("    d^2 computed = %.6f\n", sum((zx-zy)^2)))
cat(sprintf("    2(p-1)(1-r)  = %.6f   <- identical\n", 2*(p-1)*(1-cor(xv,yv))))
cat("  Clustering z-scored features by Euclidean distance IS clustering by\n")
cat("  correlation. Knowing this prevents a lot of method-shopping.\n")
```

## 2. The curse of dimensionality, eq. (18.7)

```{r curse}
header("2. In high dimensions all points are equidistant (18.7)")
set.seed(1802)
cat(sprintf("  %14s%25s\n", "dimensions p", "(max-min)/min distance"))
for (pp in c(2, 10, 100, 1000, 10000)) {
  Xd <- matrix(rnorm(200*pp), 200)
  D <- as.matrix(dist(Xd)); diag(D) <- NA
  cat(sprintf("  %14s%25.4f\n", format(pp, big.mark = ","),
              mean((apply(D, 1, max, na.rm=TRUE) - apply(D, 1, min, na.rm=TRUE)) /
                     apply(D, 1, min, na.rm=TRUE))))
}
cat("\n  The contrast between nearest and farthest neighbour COLLAPSES. This is\n")
cat("  why every scRNA-seq pipeline builds its neighbour graph in the top ~30\n")
cat("  PCs, not in 20,000-dimensional gene space.\n")
```

## 3. Linkage and k-means, eq. (18.8)-(18.10)

```{r linkage}
header("3. Linkage choice changes the answer (18.8)-(18.10)")
set.seed(1803)
truth <- rep(1:3, c(80, 40, 15))
centres <- rbind(c(0,0), c(5,0), c(2.5,4.5))
Xb <- do.call(rbind, lapply(1:3, function(k)
  matrix(rnorm(c(80,40,15)[k]*2, rep(centres[k,], each=c(80,40,15)[k]), 0.9),
         ncol = 2)))
adj_rand <- function(a, b) {
  tab <- table(a, b); n <- sum(tab)
  sum_comb <- function(x) sum(choose(x, 2))
  idx <- sum_comb(tab); ea <- sum_comb(rowSums(tab)); eb <- sum_comb(colSums(tab))
  exp_idx <- ea * eb / choose(n, 2)
  (idx - exp_idx) / ((ea + eb)/2 - exp_idx)
}
D <- dist(Xb)
cat(sprintf("  %-14s%25s\n", "linkage", "adjusted Rand vs truth"))
for (meth in c("single", "complete", "average", "ward.D2"))
  cat(sprintf("  %-14s%25.3f\n", meth,
              adj_rand(truth, cutree(hclust(D, method = meth), 3))))
km <- kmeans(Xb, 3, nstart = 25)
cat(sprintf("  %-14s%25.3f\n", "k-means", adj_rand(truth, km$cluster)))
cat("\n  Single linkage chains and usually fails on omics data. Ward.D2\n")
cat("  REQUIRES Euclidean distance; applying it to a correlation distance is\n")
cat("  not meaningful. (Note ward.D2 expects ORDINARY distances, ward.D\n")
cat("  expects SQUARED ones - a common silent error.)\n")
cat("\n  k-means depends on INITIALISATION. On these well-separated blobs even\n")
cat("  nstart=1 succeeds, so a harder case is needed to see the problem -\n")
cat("  6 overlapping clusters in 10 dimensions:\n")
set.seed(18031)
n_k <- 6
cent <- matrix(rnorm(n_k * 10, 0, 1.3), n_k)
truth_h <- rep(seq_len(n_k), each = 30)
Xh <- cent[truth_h, ] + matrix(rnorm(n_k * 30 * 10, 0, 1.0), ncol = 10)
cat(sprintf("    %-12s%12s%10s%14s\n", "nstart", "ARI mean", "ARI min", "failures/20"))
for (ns in c(1, 5, 25)) {
  aris <- sapply(1:20, function(s) {
    set.seed(s); adj_rand(truth_h, kmeans(Xh, n_k, nstart = ns)$cluster) })
  cat(sprintf("    %-12d%12.3f%10.3f%14s\n", ns, mean(aris), min(aris),
              sprintf("%d/20", sum(aris < mean(aris) - 0.05))))
}
cat("    Lloyd's algorithm converges to a LOCAL optimum; nstart takes the best\n")
cat("    of several random starts. Use nstart >= 25 and report the seed.\n")
```

## 4. Choosing K: silhouette, gap statistic, stability

```{r choosing-k}
header("4. How many clusters? (18.11)-(18.12) and stability")
gap_statistic <- function(X, k_max = 8, B = 20) {
  Wk <- function(X, k) { km <- kmeans(X, k, nstart = 10)
                         sum(km$withinss) }
  mins <- apply(X, 2, min); maxs <- apply(X, 2, max)
  gaps <- sks <- numeric(k_max)
  for (k in 1:k_max) {
    logW <- log(Wk(X, k))
    ref <- replicate(B, log(Wk(mapply(function(lo, hi) runif(nrow(X), lo, hi),
                                      mins, maxs), k)))
    gaps[k] <- mean(ref) - logW                                 # eq. (18.12)
    sks[k] <- sd(ref) * sqrt(1 + 1/B)
  }
  list(gap = gaps, sk = sks)
}
cluster_stability <- function(X, k, n_boot = 40, frac = 0.8) {
  base <- kmeans(X, k, nstart = 25)$cluster
  mean(replicate(n_boot, {
    idx <- sample(nrow(X), floor(frac*nrow(X)))
    adj_rand(base[idx], kmeans(X[idx, ], k, nstart = 25)$cluster)
  }))
}
set.seed(1804)
gs <- gap_statistic(Xb, 8)
cat(sprintf("  %3s%13s%9s%10s%18s\n", "K", "silhouette", "gap", "gap s_k", "stability (ARI)"))
for (k in 2:8) {
  lab <- kmeans(Xb, k, nstart = 25)$cluster
  sil <- mean(silhouette(lab, D)[, 3])                          # eq. (18.11)
  cat(sprintf("  %3d%13.3f%9.3f%10.3f%18.3f\n", k, sil, gs$gap[k], gs$sk[k],
              cluster_stability(Xb, k)))
}
k_gap <- which(sapply(1:7, function(k) gs$gap[k] >= gs$gap[k+1] - gs$sk[k+1]))[1]
cat(sprintf("\n  gap-statistic rule selects K = %d   (truth: 3)\n", k_gap))
cat("\n  Negative control - ONE homogeneous Gaussian cloud, no clusters:\n")
set.seed(1805)
Xnull <- matrix(rnorm(135*2), 135)
gn <- gap_statistic(Xnull, 6)
for (k in 2:5) {
  lab <- kmeans(Xnull, k, nstart = 25)$cluster
  cat(sprintf("    K=%d: silhouette %.3f, gap %+.3f, stability %.3f\n", k,
              mean(silhouette(lab, dist(Xnull))[, 3]), gn$gap[k],
              cluster_stability(Xnull, k, n_boot = 20)))
}
cat("  Silhouette is POSITIVE even with no real structure - it ALWAYS finds\n")
cat("  something. The gap statistic and stability are far more informative.\n")
```

## 5. Embedding distances and cluster areas are meaningless, eq. (18.14)

Python's module uses t-SNE. The base R image has no t-SNE, but the point is
not specific to it: any embedding that preserves only LOCAL structure
discards global geometry. Non-metric MDS (`MASS::isoMDS`) makes this
explicit, because it fits the ORDER of the distances and not their size.

```{r embedding}
header("5. Embedding distances and cluster areas are meaningless (18.14)")
set.seed(1804)
## Three groups at KNOWN, very different true separations.
g1 <- matrix(rnorm(60 * 10), 60)
g2 <- matrix(rnorm(60 * 10), 60); g2[, 1] <- g2[, 1] + 6    # near g1
g3 <- matrix(rnorm(60 * 10), 60); g3[, 1] <- g3[, 1] + 40   # VERY far
Xt <- rbind(g1, g2, g3)
lab <- rep(1:3, each = 60)
cen <- function(M) colMeans(M)
true_d12 <- sqrt(sum((cen(g1) - cen(g2))^2))
true_d13 <- sqrt(sum((cen(g1) - cen(g3))^2))
cat(sprintf("  TRUE centroid distances: d(1,2) = %.2f, d(1,3) = %.2f   ratio %.2f\n",
            true_d12, true_d13, true_d13 / true_d12))
D_emb <- dist(Xt)
cat(sprintf("\n  %-26s%13s%13s%8s\n", "embedding", "emb d(1,2)", "emb d(1,3)",
            "ratio"))
emb_ratio <- function(E) {
  a <- colMeans(E[lab == 1, ]); b <- colMeans(E[lab == 2, ])
  cc <- colMeans(E[lab == 3, ])
  c(sqrt(sum((a - b)^2)), sqrt(sum((a - cc)^2)))
}
## Metric MDS (PCA on distances) DOES preserve global geometry - the control.
e_cmd <- cmdscale(D_emb, k = 2)
r_ <- emb_ratio(e_cmd)
cat(sprintf("  %-26s%13.2f%13.2f%8.2f\n", "classical (metric) MDS", r_[1],
            r_[2], r_[2] / r_[1]))
## Non-metric MDS keeps only the RANK of the distances.
for (sd_ in c(1, 2)) {
  set.seed(sd_)
  init <- cmdscale(D_emb, k = 2) + matrix(rnorm(2 * nrow(Xt), 0, 5), ncol = 2)
  e <- suppressWarnings(MASS::isoMDS(D_emb, y = init, k = 2, trace = FALSE)$points)
  r_ <- emb_ratio(e)
  cat(sprintf("  %-26s%13.2f%13.2f%8.2f\n",
              sprintf("non-metric MDS, seed %d", sd_), r_[1], r_[2],
              r_[2] / r_[1]))
}
cat(sprintf("\n  The TRUE ratio is %.1f. Classical MDS reproduces it, because it
  minimises a METRIC stress and so keeps the geometry. Non-metric MDS does not:
  it only has to get the ORDER right, so it is free to place the far group at
  whatever distance makes the local structure fit, and the answer moves with
  the starting configuration.\n", true_d13 / true_d12))
cat("
  t-SNE and UMAP are further along the same spectrum. Both optimise a
  neighbourhood objective (eq. 18.14) that preserves LOCAL structure and
  deliberately discards global geometry, and both add a random initialisation
  on top. A t-SNE map's between-cluster gaps and cluster areas are artefacts of
  perplexity, initialisation and seed.

  NEVER: measure distances on a nonlinear embedding; compare cluster areas;
  report an embedding without the seed, perplexity or n_neighbors, and n_PCs
  used. If you need distances, compute them in the space you trust (the PCs),
  not on the picture.\n")
```

## 6. Double dipping in clustering

```{r double-dipping}
header("6. Testing the features that defined the clusters")
set.seed(1806)
n <- 100; G <- 1500
Xhom <- matrix(rnorm(n*G), n)               # ONE homogeneous population
lab2 <- kmeans(Xhom, 2, nstart = 25)$cluster
p_naive <- sapply(1:G, function(j) t.test(Xhom[lab2==1, j], Xhom[lab2==2, j])$p.value)
cat(sprintf("  Data: ONE homogeneous Gaussian cloud (n=%d, p=%d), no subgroups.\n", n, G))
cat(sprintf("  k-means K=2 cluster sizes: %s\n", paste(table(lab2), collapse = " ")))
cat(sprintf("  'differential' features at p < 0.05 : %.1f%%  (should be 5%%)\n",
            100*mean(p_naive < 0.05)))
cat(sprintf("  surviving BH FDR 0.05               : %d\n",
            sum(p.adjust(p_naive, "BH") < 0.05)))
split_test <- function(X, seed = 1) {
  ## Learn the separating direction on one half, apply it to the other.
  set.seed(seed)
  idx <- sample(nrow(X)); tr <- idx[1:(nrow(X)/2)]; te <- idx[-(1:(nrow(X)/2))]
  km_tr <- kmeans(X[tr, ], 2, nstart = 25)
  direction <- km_tr$centers[2, ] - km_tr$centers[1, ]
  proj <- X[te, ] %*% direction
  lt <- as.integer(proj > median(proj))
  sapply(seq_len(ncol(X)), function(j) t.test(X[te, j][lt==0], X[te, j][lt==1])$p.value)
}
p_split <- split_test(Xhom)
cat(sprintf("\n  With sample splitting (direction learned on half, tested on the other):\n"))
cat(sprintf("    'differential' features at p < 0.05 : %.1f%%\n", 100*mean(p_split < 0.05)))
cat(sprintf("    surviving BH FDR 0.05               : %d\n",
            sum(p.adjust(p_split, "BH") < 0.05)))
cat("\n  The decisive evidence against a noise cluster is FAILURE TO REPLICATE\n")
cat("  on held-out data. In practice the damage is larger than the p-value\n")
cat("  distribution suggests, because analysts report the TOP 'marker genes' -\n")
cat("  precisely the features the clustering selected on (stats.md Topic 21).\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "clustering.png"), width = 1100, height = 800, res = 110)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.5, 1))
plot(Xb, col = truth, pch = 16, xlab = "", ylab = "",
     main = "Three blobs of unequal size")
ks <- 2:8
plot(ks, sapply(ks, function(k) mean(silhouette(kmeans(Xb,k,nstart=25)$cluster, D)[,3])),
     type = "b", pch = 16, col = "steelblue", ylim = c(0, 1), xlab = "K", ylab = "",
     main = "Eq. (18.11): choosing K")
lines(ks, sapply(ks, function(k) cluster_stability(Xb, k, n_boot = 20)),
      type = "b", pch = 15, col = "darkorange")
abline(v = 3, col = "red", lty = 2)
legend("topright", c("silhouette", "stability (ARI)"), col = c("steelblue","darkorange"),
       pch = c(16,15), bty = "n", cex = 0.7)
hist(p_naive, breaks = 30, col = adjustcolor("steelblue", 0.6), border = "white",
     main = "Double dipping on homogeneous data", xlab = "p-value")
hist(p_split, breaks = 30, col = adjustcolor("darkorange", 0.6), border = "white", add = TRUE)
abline(h = G/30, col = "red", lty = 2)
legend("topright", c("cluster then test", "sample splitting", "uniform"),
       fill = c(adjustcolor("steelblue",0.6), adjustcolor("darkorange",0.6), NA),
       border = c("black","black",NA), lty = c(NA,NA,2), col = c(NA,NA,"red"),
       bty = "n", cex = 0.65)
plot(ks, gs$gap[ks], type = "b", pch = 16, col = "forestgreen", xlab = "K",
     ylab = "gap", main = "Eq. (18.12): gap statistic")
arrows(ks, gs$gap[ks]-gs$sk[ks], ks, gs$gap[ks]+gs$sk[ks], code = 3, angle = 90,
       length = 0.03)
abline(v = 3, col = "red", lty = 2)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "clustering.png"), "\n")
```

## Decision rules (from `stats.md` Topic 18)

1. Pick the distance from the measurement scale.
2. Cluster in a reduced space (top PCs) for high-dimensional data.
3. Report K-selection evidence AND stability, not just a coloured plot.
4. Never test the features that defined the clusters on the same data.
5. Report seed, parameters and n_PCs for every embedding; quantify nothing
   on embedding coordinates.

**Next:** `19_batch_effects.R`
