---
title: "Module 19 - Batch effects and unwanted variation"
output: html_document
---

**Curriculum link:** `stats.md` -> Topic 19, equations (19.1)-(19.6)

## What you will learn

1. The location-scale model (19.1) and a from-scratch ComBat-style
   correction (19.2) -- including **why the design must be supplied**.
2. Identifiability (19.3): complete confounding is unfixable.
3. **Model-based adjustment (19.4) vs correcting the matrix.**
4. Surrogate variables (19.5) when batch is unrecorded.
5. The two-sided diagnostic: batch removal AND biology preservation.

```{r setup, message = FALSE}
MODULE_NAME <- "19_batch_effects"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. Simulating the location-scale model, eq. (19.1)

$$Y_{ijg}=\alpha_g+\mathbf x_j'\boldsymbol\beta_g+\gamma_{ig}+\delta_{ig}\varepsilon_{ijg}$$

```{r simulate}
header("1. A multi-batch experiment with known truth (19.1)")
simulate_batches <- function(n_per_group = 12, G = 2000, n_de = 200, effect = 1.8,
                             batch_shift = 1.5, batch_scale = 1.6, confounding = 0) {
  n <- 2 * n_per_group
  condition <- rep(c("ctrl", "trt"), each = n_per_group)
  follow <- runif(n) < confounding
  batch <- ifelse(follow, ifelse(condition == "ctrl", "b1", "b2"),
                  sample(c("b1", "b2"), n, TRUE))
  ## Guard against n_de > G (the identifiability demo below uses a small G).
  n_de <- min(n_de, G)
  alpha <- rnorm(G, 8, 1.5)
  beta <- numeric(G); beta[seq_len(n_de)] <- effect * sample(c(-1, 1), n_de, TRUE)
  gamma <- rnorm(G, 0, batch_shift)
  delta <- exp(rnorm(G, 0, log(batch_scale)))
  Y <- matrix(0, G, n)
  for (j in seq_len(n)) {
    shift <- if (batch[j] == "b2") gamma else 0
    scl <- if (batch[j] == "b2") delta else 1
    Y[, j] <- alpha + beta * (condition[j] == "trt") + shift + scl * rnorm(G)
  }
  dimnames(Y) <- list(sprintf("G%04d", seq_len(G)), sprintf("S%02d", seq_len(n)))
  list(Y = Y, meta = data.frame(condition = condition, batch = batch,
                                row.names = colnames(Y)),
       truth = beta != 0)
}
set.seed(1901)
sim <- simulate_batches()
print(table(sim$meta$condition, sim$meta$batch))
cat(sprintf("\n  %d of %d genes are truly differential\n", sum(sim$truth), length(sim$truth)))
```

## 2. Diagnose before correcting: PC x metadata

```{r diagnose}
header("2. Which factor dominates the variance?")
pc_metadata_table <- function(Y, meta, n_pc = 4) {
  pca <- prcomp(t(Y), center = TRUE)
  pve <- pca$sdev^2 / sum(pca$sdev^2)
  out <- data.frame(PC = 1:n_pc, PVE = pve[1:n_pc])
  for (col in names(meta))
    out[[col]] <- sapply(1:n_pc, function(j)
      summary(lm(pca$x[, j] ~ factor(meta[[col]])))$r.squared)
  list(tab = out, scores = pca$x, pve = pve)
}
d1 <- pc_metadata_table(sim$Y, sim$meta)
print(format(d1$tab, digits = 3), row.names = FALSE)
cat("\n  PC1 is dominated by BATCH. Build this table before interpreting\n")
cat("  anything (stats.md eq. 17.6).\n")
```

## 3. Identifiability, eq. (19.3)

```{r identifiability}
header("3. Complete vs partial confounding (19.3)")
cat(sprintf("  %-14s%-26s%12s%18s%8s\n", "confounding", "cond x batch table",
            "rank/ncol", "R^2(cond~batch)", "VIF"))
for (conf in c(0, 0.5, 0.8, 1.0)) {
  set.seed(1902 + round(conf*10))
  s <- simulate_batches(confounding = conf, G = 50)
  ct <- as.vector(table(s$meta$condition, s$meta$batch))
  X <- model.matrix(~ factor(condition) + factor(batch), data = s$meta)
  r2 <- suppressWarnings(cor(as.numeric(s$meta$condition == "trt"),
                             as.numeric(s$meta$batch == "b2"))^2)
  if (is.na(r2)) r2 <- 1
  cat(sprintf("  %-14.1f%-26s%12s%18.3f%8s\n", conf, paste(ct, collapse = " "),
              sprintf("%d/%d", qr(X)$rank, ncol(X)), r2,
              if (r2 < 1) sprintf("%.2f", 1/(1-r2)) else "Inf"))
}
cat("\n  At confounding = 1.0 the design matrix is RANK DEFICIENT: the\n")
cat("  condition effect is not estimable by ANY method. At 0.8 it is\n")
cat("  estimable but the SE is inflated by sqrt(VIF).\n")
```

## 4. ComBat-style correction from scratch, eq. (19.2)

```{r combat}
header("4. Location-scale correction, with and without the design (19.2)")
combat_like <- function(Y, batch, mod = NULL) {
  ## The non-empirical-Bayes core of ComBat, stats.md eq. (19.2). `mod` is the
  ## BIOLOGICAL design; supplying it is what PROTECTS the biology.
  batch <- factor(batch); levs <- levels(batch)
  B <- model.matrix(~ batch - 1)
  X <- if (is.null(mod)) B else cbind(B, mod)
  coef <- qr.coef(qr(X), t(Y))
  n_b <- ncol(B)
  w <- as.numeric(table(batch)) / length(batch)
  grand <- as.vector(w %*% coef[1:n_b,, drop = FALSE])
  bio <- if (is.null(mod)) 0 else mod %*% coef[-(1:n_b),, drop = FALSE]
  fitted <- X %*% coef
  resid <- t(Y) - fitted
  pooled_var <- colSums(resid^2) / (nrow(X) - ncol(X))
  Ystar <- Y
  for (l in levs) {
    sel <- batch == l
    delta <- sqrt(pmax(colMeans(resid[sel,, drop = FALSE]^2) /
                         pmax(pooled_var, 1e-12), 1e-12))
    centred <- sweep(t(Y[, sel, drop = FALSE]) - fitted[sel,, drop = FALSE], 2, delta, "/")
    add <- sweep(centred, 2, grand, "+")
    if (!is.null(mod)) add <- add + bio[sel,, drop = FALSE]
    Ystar[, sel] <- t(add)
  }
  Ystar
}
de_test <- function(Y, meta) {
  a <- Y[, meta$condition == "ctrl", drop = FALSE]
  b <- Y[, meta$condition == "trt", drop = FALSE]
  n1 <- ncol(a); n2 <- ncol(b)
  m1 <- rowMeans(a); m2 <- rowMeans(b)
  v1 <- apply(a, 1, var); v2 <- apply(b, 1, var)
  se <- sqrt(v1/n1 + v2/n2)
  df <- (v1/n1 + v2/n2)^2 / ((v1/n1)^2/(n1-1) + (v2/n2)^2/(n2-1))
  2 * pt(abs((m1 - m2)/se), df, lower.tail = FALSE)
}
de_adjusted <- function(Y, meta) {                               # eq. (19.4)
  X <- model.matrix(~ factor(condition) + factor(batch), data = meta)
  coef <- qr.coef(qr(X), t(Y))
  resid <- t(Y) - X %*% coef
  dof <- nrow(X) - ncol(X)
  s2 <- colSums(resid^2) / dof
  se <- sqrt(s2 * chol2inv(qr.R(qr(X)))[2, 2])
  2 * pt(abs(coef[2, ] / se), dof, lower.tail = FALSE)
}
evaluate <- function(pv, truth, label) {
  q <- p.adjust(pv, "BH"); rej <- q < 0.05
  tp <- sum(rej & truth); fp <- sum(rej & !truth)
  cat(sprintf("  %-40s%7d%7d%7d%10.2f%8.3f\n", label, sum(rej), tp, fp,
              tp/sum(truth), fp/max(sum(rej), 1)))
}
compare_all <- function(conf, seed) {
  set.seed(seed)
  s <- simulate_batches(confounding = conf)
  bio <- model.matrix(~ factor(condition), data = s$meta)[, 2, drop = FALSE]
  cat(sprintf("\n  --- confounding = %.1f;  cond x batch table = %s ---\n", conf,
              paste(as.vector(table(s$meta$condition, s$meta$batch)), collapse = " ")))
  cat(sprintf("  %-40s%7s%7s%7s%10s%8s\n", "analysis", "rej", "TP", "FP", "sens", "FDP"))
  evaluate(de_test(s$Y, s$meta), s$truth, "no correction, no adjustment")
  evaluate(de_test(combat_like(s$Y, s$meta$batch, NULL), s$meta), s$truth,
           "matrix corrected WITHOUT the design")
  evaluate(de_test(combat_like(s$Y, s$meta$batch, bio), s$meta), s$truth,
           "matrix corrected WITH the design")
  evaluate(de_adjusted(s$Y, s$meta), s$truth, "batch IN THE MODEL (19.4)")
  s
}
invisible(compare_all(0, 1901))
sim_c <- compare_all(0.6, 1911)
cat("\n  Read the two blocks together:\n")
cat("   * BALANCED: every strategy works. Removing batch means cannot touch\n")
cat("     the condition effect because the two are ORTHOGONAL. This is the\n")
cat("     strongest argument for balanced allocation at the design stage.\n")
cat("   * CONFOUNDED: correcting WITHOUT the design strips out real biology.\n")
cat("     Supplying the design, or adjusting in the model, protects it.\n")
cat("\n  Look at the FDP column in the confounded block. Model-based adjustment\n")
cat("  holds FDP near nominal while the corrected MATRIX overshoots: the\n")
cat("  correction was ESTIMATED, and a corrected matrix discards that\n")
cat("  uncertainty. Correct the matrix for VISUALISATION; adjust in the model\n")
cat("  for INFERENCE.\n")
```

## 5. Surrogate variables when batch is unrecorded, eq. (19.5)

```{r sva}
header("5. Estimating unrecorded batch as surrogate variables (19.5)")
set.seed(1903)
s2 <- simulate_batches()
bio2 <- model.matrix(~ factor(condition), data = s2$meta)
coef_bio <- qr.coef(qr(bio2), t(s2$Y))
R <- t(s2$Y) - bio2 %*% coef_bio                                 # eq. (19.5)
sv <- svd(R)
SV <- sv$u[, 1, drop = FALSE]
hidden <- as.numeric(s2$meta$batch == "b2")
cat(sprintf("  |correlation(SV1, true hidden batch)| = %.4f\n", abs(cor(SV[,1], hidden))))
cat(sprintf("  residual variance explained by SV1    = %.1f%%\n",
            100 * sv$d[1]^2 / sum(sv$d^2)))
de_with_sv <- function(Y, bio, svs) {
  X <- cbind(1, bio, svs)
  coef <- qr.coef(qr(X), t(Y))
  resid <- t(Y) - X %*% coef
  dof <- nrow(X) - ncol(X)
  se <- sqrt((colSums(resid^2)/dof) * chol2inv(qr.R(qr(X)))[2, 2])
  2 * pt(abs(coef[2, ]/se), dof, lower.tail = FALSE)
}
cat(sprintf("\n  %-40s%7s%7s%7s%10s%8s\n", "analysis", "rej", "TP", "FP", "sens", "FDP"))
evaluate(de_test(s2$Y, s2$meta), s2$truth, "ignore the hidden batch")
evaluate(de_with_sv(s2$Y, bio2[, 2], SV), s2$truth, "adjust for 1 surrogate variable")
evaluate(de_adjusted(s2$Y, s2$meta), s2$truth, "adjust for the TRUE batch (oracle)")
cat("\n  The surrogate variable recovers most of what knowing the true batch\n")
cat("  would have given. RUV is the alternative, using NEGATIVE CONTROL\n")
cat("  features (eq. 19.6) - valid only if those controls really are\n")
cat("  unaffected by the biology, an assumption to defend.\n")
```

## 6. The two-sided diagnostic

```{r two-sided}
header("6. Batch removal AND biology preservation")
bio_c <- model.matrix(~ factor(condition), data = sim_c$meta)[, 2, drop = FALSE]
knn_mixing <- function(Y, meta, k = 5) {
  D <- as.matrix(dist(t(Y))); diag(D) <- Inf
  b <- meta$batch
  mean(sapply(seq_along(b), function(i) mean(b[order(D[i, ])[1:k]] != b[i])))
}
cat("  (using the PARTIALLY CONFOUNDED dataset, where the choice matters)\n")
cat(sprintf("  %-40s%15s%14s%12s%14s\n", "matrix", "PC1 R^2 batch", "PC1 R^2 cond",
            "kNN mixing", "true-DE sens"))
for (row_ in list(list("uncorrected", sim_c$Y),
                  list("corrected WITHOUT design", combat_like(sim_c$Y, sim_c$meta$batch, NULL)),
                  list("corrected WITH design", combat_like(sim_c$Y, sim_c$meta$batch, bio_c)))) {
  M <- row_[[2]]
  pcm <- pc_metadata_table(M, sim_c$meta, n_pc = 1)$tab
  q <- p.adjust(de_test(M, sim_c$meta), "BH")
  cat(sprintf("  %-40s%15.3f%14.3f%12.3f%14.2f\n", row_[[1]],
              pcm$batch[1], pcm$condition[1], knn_mixing(M, sim_c$meta),
              sum(q < 0.05 & sim_c$truth)/sum(sim_c$truth)))
}
cat("\n  Read the columns against each other. The design-BLIND correction wins\n")
cat("  on every 'batch removal' metric - PC1 batch R^2 of 0.000 and near-\n")
cat("  perfect kNN mixing - and yet recovers ZERO true genes. It scored best\n")
cat("  on the thing that is easy to measure while destroying the thing you\n")
cat("  cared about. The design-aware correction leaves some batch structure\n")
cat("  behind and keeps the biology.\n")
cat("\n  That asymmetry is the whole lesson: a mixing metric ALONE rewards\n")
cat("  overcorrection. Always report a biology-preservation metric beside it\n")
cat("  (stats.md Topic 19).\n")
```

## 7. Figure

```{r figure}
png(file.path(OUT, "batch_effects.png"), width = 1400, height = 430, res = 110)
par(mfrow = c(1, 3), mar = c(4.2, 4.2, 2.5, 1))
for (row_ in list(list("uncorrected (confounded)", sim_c$Y),
                  list("corrected WITHOUT design", combat_like(sim_c$Y, sim_c$meta$batch, NULL)),
                  list("corrected WITH design", combat_like(sim_c$Y, sim_c$meta$batch, bio_c)))) {
  pcm <- pc_metadata_table(row_[[2]], sim_c$meta)
  plot(pcm$scores[, 1], pcm$scores[, 2], asp = 1,
       pch = c(16, 17)[factor(sim_c$meta$batch)],
       col = c("steelblue", "firebrick")[factor(sim_c$meta$condition)], cex = 1.3,
       xlab = sprintf("PC1 (%.0f%%)", 100*pcm$pve[1]),
       ylab = sprintf("PC2 (%.0f%%)", 100*pcm$pve[2]), main = row_[[1]])
}
legend("topright", c("b1", "b2", "ctrl", "trt"), pch = c(16, 17, 15, 15),
       col = c("black", "black", "steelblue", "firebrick"), bty = "n", cex = 0.6)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "batch_effects.png"), "\n")
```

## Decision rules (from `stats.md` Topic 19)

1. **Prevent**: randomise across batches, balance conditions within batch,
   include bridging samples, record every processing variable.
2. **Diagnose before correcting.** If completely confounded, stop and say so.
3. Adjust in the model (19.4) for INFERENCE; correct the matrix (19.2) only
   for VISUALISATION -- and keep it inside CV folds (Module 21).
4. Always evaluate BOTH batch removal and biology preservation.

**Next:** `20_survival_and_longitudinal.R`
