---
title: "Module 34 - Diagnostics, sensitivity analysis, and reproducibility"
output: html_document
---

**Curriculum link:** `stats.md` -> Topics 34 and 35, equations (34.1)-(34.3)

## What you will learn

1. **Randomised quantile residuals (34.1)**: the residual that works for
   discrete outcomes.
2. Leave-one-sample-out and leave-one-batch-out (34.2).
3. **Specification-curve / multiverse analysis (34.3)**.
4. The permuted-label negative control -- the check that catches most bugs.
5. Simulation-based calibration.
6. Property-based tests for statistical code, and seeding done right.

```{r setup, message = FALSE}
suppressPackageStartupMessages(library(MASS))
MODULE_NAME <- "34_diagnostics_and_reproducibility"
OUT <- file.path(Sys.getenv("STATS_OUT", unset = "results"), MODULE_NAME)
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
header <- function(txt) cat("\n", strrep("=", 72), "\n", txt, "\n",
                            strrep("=", 72), "\n", sep = "")
```

## 1. Randomised quantile residuals, eq. (34.1)

$$r_i=\Phi^{-1}(u_i),\quad u_i\sim U(F(y_i-1),\,F(y_i))$$

Raw residuals from a count model form useless discrete bands. Randomised
quantile residuals are exactly N(0,1) under a CORRECT model.

```{r rqr}
header("1. Residuals that work for discrete outcomes (34.1)")
randomised_quantile_residuals <- function(y, cdf_lo, cdf_hi, seed = 1) {
  set.seed(seed)
  qnorm(pmin(pmax(runif(length(y), cdf_lo, cdf_hi), 1e-12), 1 - 1e-12))
}
set.seed(2401)
n <- 600
xq <- rnorm(n); PHI <- 0.35
mu <- exp(1.6 + 0.5*xq)
yq <- rnbinom(n, mu = mu, size = 1/PHI)
dq <- data.frame(y = yq, x = xq)
m_pois <- glm(y ~ x, data = dq, family = poisson())
m_nb <- glm.nb(y ~ x, data = dq)
rq_pois <- randomised_quantile_residuals(
  yq, ppois(yq - 1, fitted(m_pois)), ppois(yq, fitted(m_pois)))
rq_nb <- randomised_quantile_residuals(
  yq, pnbinom(yq - 1, mu = fitted(m_nb), size = m_nb$theta),
  pnbinom(yq, mu = fitted(m_nb), size = m_nb$theta))
cat(sprintf("  %-22s%9s%8s%12s%17s\n", "model", "mean", "SD", "Shapiro p",
            "KS vs N(0,1) p"))
for (row_ in list(list("Poisson (WRONG)", rq_pois), list("neg. binomial (right)", rq_nb)))
  cat(sprintf("  %-22s%9.3f%8.3f%12.2e%17.4f\n", row_[[1]], mean(row_[[2]]),
              sd(row_[[2]]), shapiro.test(row_[[2]][1:500])$p.value,
              suppressWarnings(ks.test(row_[[2]], "pnorm")$p.value)))
cat("\n  Under the CORRECT model the residuals are exactly N(0,1); under the\n")
cat("  wrong one they are not. Ordinary deviance residuals cannot make this\n")
cat("  distinction nearly as cleanly for count data.\n")
```

## 2. Leave-one-sample-out, eq. (34.2)

```{r loo}
header("2. How much does one sample drive the result? (34.2)")
set.seed(2402)
G <- 3000; n_per <- 6
is_de <- c(rep(TRUE, 150), rep(FALSE, G - 150))
A <- matrix(rnorm(G*n_per, ifelse(is_de, 1.3, 0)), nrow = G)
B <- matrix(rnorm(G*n_per), nrow = G)
A[, 1] <- A[, 1] + rnorm(G, 0, 2.5)            # inject ONE influential outlier
de_hits <- function(A, B) {
  n1 <- ncol(A); n2 <- ncol(B)
  m1 <- rowMeans(A); m2 <- rowMeans(B)
  v1 <- apply(A, 1, var); v2 <- apply(B, 1, var)
  se <- sqrt(v1/n1 + v2/n2)
  df <- (v1/n1 + v2/n2)^2/((v1/n1)^2/(n1-1) + (v2/n2)^2/(n2-1))
  which(p.adjust(2*pt(abs((m1-m2)/se), df, lower.tail = FALSE), "BH") < 0.05)
}
full_hits <- de_hits(A, B)
cat(sprintf("  full analysis: %d discoveries at FDR 5%%\n", length(full_hits)))
cat(sprintf("\n  %-20s%13s%11s%18s\n", "dropped sample", "discoveries", "% of full",
            "Jaccard vs full"))
for (j in seq_len(n_per)) {
  h <- de_hits(A[, -j, drop = FALSE], B)
  jac <- length(intersect(h, full_hits)) / max(length(union(h, full_hits)), 1)
  flag <- if (length(h) < 0.7*length(full_hits) || length(h) > 1.4*length(full_hits))
    "  <-- influential" else ""
  cat(sprintf("  %-20s%13d%10.0f%%%18.3f%s\n", paste0("A", j), length(h),
              100*length(h)/max(length(full_hits), 1), jac, flag))
}
cat("\n  Sample A1 is the injected outlier, and the leave-one-out scan finds it\n")
cat("  without being told. Run this for ANY small-n omics result and report\n")
cat("  the range, not just the headline count.\n")
```

## 3. Specification curve / multiverse analysis, eq. (34.3)

```{r multiverse}
header("3. Multiverse analysis (34.3)")
set.seed(2403)
n <- 220
dm <- data.frame(batch = sample(0:2, n, TRUE), age = rnorm(n, 55, 12),
                 sex = rbinom(n, 1, 0.5), treat = rbinom(n, 1, 0.5))
TRUE_EFF <- 0.30
dm$y_raw <- exp(2 + TRUE_EFF*dm$treat + 0.02*dm$age + 0.15*dm$sex +
                  0.25*dm$batch + rnorm(n, 0, 0.5))
dm$y_raw[sample(n, 6)] <- dm$y_raw[sample(n, 6)] * 6      # a few extreme values
choices <- expand.grid(transform = c("log", "raw", "rank"),
                       outliers = c("keep", "trim 1%", "winsorise"),
                       covariates = c("none", "+age+sex", "+age+sex+batch"),
                       se = c("classical", "HC3"), stringsAsFactors = FALSE)
vcov_hc3 <- function(fit) {
  X <- model.matrix(fit); e <- residuals(fit); h <- hatvalues(fit)
  b <- solve(crossprod(X)); b %*% crossprod(X * sqrt(e^2/(1-h)^2)) %*% b
}
spec <- do.call(rbind, lapply(seq_len(nrow(choices)), function(i) {
  ch <- choices[i, ]
  d2 <- dm
  if (ch$outliers == "trim 1%") {
    qq <- quantile(d2$y_raw, c(0.005, 0.995))
    d2 <- d2[d2$y_raw >= qq[1] & d2$y_raw <= qq[2], ]
  } else if (ch$outliers == "winsorise") {
    qq <- quantile(d2$y_raw, c(0.01, 0.99))
    d2$y_raw <- pmin(pmax(d2$y_raw, qq[1]), qq[2])
  }
  d2$y <- switch(ch$transform, log = log(d2$y_raw),
                 raw = d2$y_raw/sd(d2$y_raw), rank = rank(d2$y_raw)/nrow(d2))
  form <- paste("y ~ treat", switch(ch$covariates, none = "",
                                    `+age+sex` = "+ age + sex",
                                    `+age+sex+batch` = "+ age + sex + factor(batch)"))
  f <- lm(as.formula(form), data = d2)
  se <- if (ch$se == "HC3") sqrt(diag(vcov_hc3(f))["treat"])
        else coef(summary(f))["treat", "Std. Error"]
  est <- coef(f)["treat"]
  tcrit <- qt(0.975, df.residual(f))
  data.frame(ch, estimate = est, se = se,
             p = 2*pt(abs(est/se), df.residual(f), lower.tail = FALSE),
             lo = est - tcrit*se, hi = est + tcrit*se)
}))
spec <- spec[order(spec$estimate), ]
cat(sprintf("  %d defensible specifications (%s)\n", nrow(spec),
            paste(sapply(choices[, 1:4], function(cc) length(unique(cc))), collapse = " x ")))
cat(sprintf("  estimate range : [%.3f, %.3f]\n", min(spec$estimate), max(spec$estimate)))
cat(sprintf("  median estimate: %.3f\n", median(spec$estimate)))
cat(sprintf("  sign stable?   : %s\n", length(unique(sign(spec$estimate))) == 1))
cat(sprintf("  p < 0.05 in    : %.1f%% of specifications\n", 100*mean(spec$p < 0.05)))
cat("\n  Which single choice moves the result most?\n")
for (cc in c("transform", "outliers", "covariates", "se")) {
  g <- tapply(spec$estimate, spec[[cc]], mean)
  cat(sprintf("    %-12s spread across levels = %.4f   %s\n", cc, max(g) - min(g),
              paste(sprintf("%s:%.3f", names(g), g), collapse = ", ")))
}
cat("\n  A conclusion that holds in 5 of 54 specifications is a\n")
cat("  SPECIFICATION-DEPENDENT conclusion, and saying so is a finding. Note\n")
cat("  the transform changes the SCALE of the coefficient, so compare SIGN and\n")
cat("  SIGNIFICANCE across transforms, magnitudes only within one.\n")
```

## 4. The permuted-label negative control

```{r negative-control}
header("4. Negative control: does your pipeline find nothing in noise?")
pipeline_under_test <- function(A, B, broken = FALSE) {
  if (broken) {
    ## OUTCOME-DEPENDENT filtering - invalidates everything downstream.
    d <- abs(rowMeans(A) - rowMeans(B))
    keep <- d > median(d); A <- A[keep,, drop = FALSE]; B <- B[keep,, drop = FALSE]
  }
  n1 <- ncol(A); n2 <- ncol(B)
  m1 <- rowMeans(A); m2 <- rowMeans(B)
  v1 <- apply(A, 1, var); v2 <- apply(B, 1, var)
  se <- sqrt(v1/n1 + v2/n2)
  df <- (v1/n1 + v2/n2)^2/((v1/n1)^2/(n1-1) + (v2/n2)^2/(n2-1))
  2*pt(abs((m1-m2)/se), df, lower.tail = FALSE)
}
set.seed(2404)
G <- 4000; n_per <- 5
cat(sprintf("  %-24s%30s%18s\n", "", "mean discoveries at FDR 5%", "P(p<0.05)"))
for (broken in c(FALSE, TRUE)) {
  res <- t(replicate(30, {
    A <- matrix(rnorm(G*n_per), G); B <- matrix(rnorm(G*n_per), G)
    p <- pipeline_under_test(A, B, broken)     # labels are PURE NOISE
    c(sum(p.adjust(p, "BH") < 0.05), mean(p < 0.05))
  }))
  cat(sprintf("  %-24s%30.1f%18.3f\n",
              if (broken) "pipeline with a bug" else "correct pipeline",
              mean(res[, 1]), mean(res[, 2])))
}
cat("\n  A correct pipeline on permuted labels yields ~0 discoveries and a FLAT\n")
cat("  p-value histogram. Anything else means the pipeline is broken - usually\n")
cat("  leakage, outcome-dependent filtering, or unmodelled dependence. Run\n")
cat("  this before you trust ANY result.\n")
```

## 5. Simulation-based calibration

```{r sbc}
header("5. Do your nominal 95% intervals actually cover 95%?")
TRUE_BETA <- 0.5
sim_independent <- function(n = 40) {
  x <- rnorm(n); data.frame(x = x, y = TRUE_BETA*x + rnorm(n))
}
sim_clustered <- function(n_clust = 8, per = 5) {
  cl <- rep(rnorm(n_clust, 0, 1.2), each = per)
  x <- rep(rnorm(n_clust), each = per) + rnorm(n_clust*per, 0, 0.3)
  data.frame(x = x, y = TRUE_BETA*x + cl + rnorm(n_clust*per, 0, 0.5),
             cl = factor(rep(seq_len(n_clust), each = per)))
}
cluster_ci <- function(d) {
  ## Cluster-robust sandwich, eq. (14.8).
  f <- lm(y ~ x, data = d); X <- model.matrix(f); e <- residuals(f)
  b <- solve(crossprod(X))
  meat <- Reduce(`+`, lapply(split(seq_len(nrow(d)), d$cl), function(idx)
    tcrossprod(crossprod(X[idx,, drop = FALSE], e[idx]))))
  V <- b %*% meat %*% b
  coef(f)["x"] + c(-1, 1)*qt(0.975, nlevels(d$cl) - 2)*sqrt(V[2, 2])
}
calibration_check <- function(fit_ci, simulate, truth, n_sim = 1500, seed = 1) {
  set.seed(seed)
  mean(replicate(n_sim, { ci <- fit_ci(simulate()); ci[1] <= truth && truth <= ci[2] }))
}
ols_ci <- function(d) confint(lm(y ~ x, data = d))["x", ]
cat(sprintf("  %-44s%28s\n", "scenario", "coverage of nominal 95% CI"))
cat(sprintf("  %-44s%27.1f%%\n", "independent data, OLS interval",
            100*calibration_check(ols_ci, sim_independent, TRUE_BETA, seed = 1)))
cat(sprintf("  %-44s%27.1f%%\n", "CLUSTERED data, naive OLS interval",
            100*calibration_check(ols_ci, sim_clustered, TRUE_BETA, seed = 2)))
cat(sprintf("  %-44s%27.1f%%\n", "CLUSTERED data, cluster-robust interval",
            100*calibration_check(cluster_ci, sim_clustered, TRUE_BETA, seed = 3)))
cat("\n  Two findings, and the second is the more interesting one:\n")
cat("   * Naive OLS on clustered data covers far below 95%. That is the\n")
cat("     pseudoreplication of Module 01, measured directly.\n")
cat("   * The cluster-robust interval fixes MOST but NOT ALL of it. With only\n")
cat("     8 clusters the sandwich estimator (eq. 14.8) is itself biased -\n")
cat("     exactly the J >= 40 guidance from Module 14. With few clusters, use\n")
cat("     a mixed model or aggregate to the cluster level.\n")
```

## 6. Property-based tests for statistical code

```{r property-tests}
header("6. The property tests every analysis script should have")
tests <- logical(0)
check <- function(name, ok, detail = "") {
  tests <<- c(tests, isTRUE(ok))
  cat(sprintf("  [%s] %s%s\n", if (isTRUE(ok)) "PASS" else "FAIL", name,
              if (nzchar(detail)) paste0("  ", detail) else ""))
}
set.seed(2405)
X <- matrix(rnorm(150), 50, 3)
## as.vector(): X %*% b returns a 50x1 MATRIX, and all.equal() would then
## compare a matrix to a vector and report an attribute mismatch.
yv <- as.vector(X %*% c(1, -2, 0.5) + rnorm(50))
Xd <- cbind(1, X)
mine <- unname(qr.coef(qr(Xd), yv))
theirs <- unname(coef(lm(yv ~ X)))
check("known-answer: QR solve matches lm()",
      isTRUE(all.equal(mine, theirs, tolerance = 1e-10)),
      sprintf("max diff %.2e", max(abs(mine - theirs))))
perm <- sample(50)
check("invariance: row permutation leaves coefficients unchanged",
      isTRUE(all.equal(mine, unname(qr.coef(qr(Xd[perm, ]), yv[perm])),
                       tolerance = 1e-10)))
check("equivariance: y -> 3y scales coefficients by 3",
      isTRUE(all.equal(3*mine, unname(qr.coef(qr(Xd), 3*yv)), tolerance = 1e-10)))
assay <- matrix(rnorm(60), 10, 6, dimnames = list(NULL, sprintf("S%d", 1:6)))
meta <- data.frame(grp = rep(c("a","b"), each = 3), row.names = sprintf("S%d", 1:6))
check("alignment: assay columns == metadata rows (eq. 2.1)",
      identical(colnames(assay), rownames(meta)))
## Distinct variables: `lm(rnorm(40) ~ rnorm(40))` makes R treat the identical
## call as the SAME term and drop it - a subtle formula trap.
fp <- mean(replicate(2000, {
  dnull <- data.frame(y = rnorm(40), x = rnorm(40))
  coef(summary(lm(y ~ x, data = dnull)))["x", "Pr(>|t|)"] < 0.05
}))
check("null calibration: FPR within 1% of nominal 0.05", abs(fp - 0.05) < 0.01,
      sprintf("observed %.4f", fp))
## Boundary: a perfectly collinear predictor must yield NA, never a number.
dcol <- data.frame(y = rnorm(20), x1 = rnorm(20))
dcol$x2 <- dcol$x1                                   # exact duplicate column
check("boundary: an aliased predictor yields NA, not silent nonsense",
      is.na(coef(lm(y ~ x1 + x2, data = dcol))["x2"]))
cat(sprintf("\n  %d/%d property tests passed.\n", sum(tests), length(tests)))
```

## 7. Seeding and reproducibility mechanics

```{r seeding}
header("7. Reproducibility mechanics")
cat("  (a) Seed at the POINT OF USE, not once at the top:\n")
cat("      set.seed(101); boot_out  <- ...\n")
cat("      set.seed(202); perm_out  <- ...\n")
cat("      -> inserting a line above cannot shift any downstream draw.\n")
cat("\n  (b) Parallel-safe streams with L'Ecuyer-CMRG:\n")
old_kind <- RNGkind("L'Ecuyer-CMRG"); set.seed(2024)
state <- .Random.seed; draws <- numeric(3)
for (i in 1:3) {
  assign(".Random.seed", state, envir = globalenv())
  draws[i] <- rnorm(1); state <- parallel::nextRNGStream(state)
}
cat(sprintf("      three worker first-draws: %s\n",
            paste(round(draws, 4), collapse = " ")))
do.call(RNGkind, as.list(old_kind))
cat("\n  (c) Environment record, emitted with every run:\n")
cat(sprintf("      R %s on %s\n", getRversion(), R.version$platform))
for (pkg in c("MASS", "nlme", "survival", "mgcv"))
  cat(sprintf("      %-10s %s\n", pkg, as.character(packageVersion(pkg))))
cat("      (sessionInfo() is the full record - print it in every report.)\n")
cat("\n  (d) Project layout that makes results/ disposable:\n")
cat("      data/raw/      read-only, checksummed, never written by code\n")
cat("      data/derived/  regenerable; safe to delete\n")
cat("      results/       figures and tables; regenerable\n")
cat("      If one command cannot rebuild results/ from data/raw/, the analysis\n")
cat("      is not reproducible.\n")
cat("\n  (e) Export the FULL statistic table, not the filtered list:\n")
cat("      estimate, SE, statistic, raw p, adjusted p, n, base mean. Someone\n")
cat("      will want to re-threshold or meta-analyse - which needs every row.\n")
```

## 8. Figure

```{r figure}
png(file.path(OUT, "diagnostics.png"), width = 1100, height = 800, res = 110)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.5, 1))
qqnorm(rq_pois, main = "RQ residuals: WRONG (Poisson) model", pch = ".", col = "steelblue")
qqline(rq_pois, col = "red")
qqnorm(rq_nb, main = "Same residuals under the CORRECT (NB) model", pch = ".",
       col = "steelblue")
qqline(rq_nb, col = "red")
plot(seq_len(nrow(spec)), spec$estimate, pch = 16, cex = 0.5, col = "steelblue",
     ylim = range(c(spec$lo, spec$hi)), xlab = "specification (ordered)",
     ylab = "treatment coefficient",
     main = sprintf("Eq. (34.3): specification curve (%d)", nrow(spec)))
segments(seq_len(nrow(spec)), spec$lo, seq_len(nrow(spec)), spec$hi, col = "grey60")
abline(h = 0, lty = 2)
set.seed(99)
A0 <- matrix(rnorm(4000*5), 4000); B0 <- matrix(rnorm(4000*5), 4000)
hist(pipeline_under_test(A0, B0, FALSE), breaks = 40,
     col = adjustcolor("steelblue", 0.6), border = "white",
     main = "Negative control: histogram must be flat", xlab = "p-value on NULL data")
hist(pipeline_under_test(A0, B0, TRUE), breaks = 40,
     col = adjustcolor("firebrick", 0.6), border = "white", add = TRUE)
legend("topright", c("correct pipeline", "pipeline with a bug"),
       fill = c(adjustcolor("steelblue",0.6), adjustcolor("firebrick",0.6)),
       bty = "n", cex = 0.65)
invisible(dev.off())
cat("\nFigure written to", file.path(OUT, "diagnostics.png"), "\n")
```

## Decision rules (from `stats.md` Topics 34-35)

1. Diagnose every model you report, with residuals appropriate to its family.
2. Run leave-one-sample-out and leave-one-batch-out for small-n results.
3. Pre-specify the primary analysis; report the multiverse around it.
4. Always run the permuted-label negative control.
5. Seed at the point of use, pin the environment, version-stamp annotations,
   and export full statistic tables.

**Next:** `36_modern_multiple_testing.R`
